---
title: "Correction of spatial heterogeneity"
author: "T. Flutre"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: TRUE
abstract: "This document shows with a simulated example how to correct for spatial heterogeneity in a field trial."
---


# Preamble

License: [CC BY-SA 4.0](http://creativecommons.org/licenses/by-sa/4.0/)

Dependencies:
```{r}
options(digits=5)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(MASS))
## suppressPackageStartupMessages(library(Matrix))
## suppressPackageStartupMessages(library(sparseMVN))
suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(SpATS))
suppressPackageStartupMessages(library(rutilstimflutre))
```

Execution time (see the appendix):
```{r time_0}
t0 <- proc.time()
```


# Overview

Let us assume a whole field containing a platform of thousands of micro-plots, with several cereal trials, hundreds of micro-plots each, separated by borders.
We will then analyze the data with two different approaches.

* step 1: simulate yield data for each micro-plot, with spatial heterogeneity at the scale of the whole field;

* step 2: compute an index of data quality per trial, i.e., by fitting a simple model for each trial separately, accounting for blocks;

* step 3: fit a more complex model for all trials jointly, accounting for spatial heterogeneity via splines, remove the spatial component from the yield data, and re-compute the quality index per trial, expecting the index values to be better than those from step 2.


# Simulate data

```{r}
set.seed(12345)
```

## Dimensions

```{r}
platformDims <- c("X"=160, "Y"=15)
(totNbMicroPlots <- as.numeric(platformDims["X"] * platformDims["Y"]))

levSeries <- c("A1","A2","A3", "B1","B2","B3", "C")
nbSeries <- 7
nbGenosPerSerie <- 40
(nbGenosInTrial <- nbSeries * nbGenosPerSerie)

nbBlocksPerTrial <- 2
(nbMicroPlotsPerTrial <- nbGenosPerSerie * nbBlocksPerTrial)
```

## Map

```{r}
## whole platform
## coords: columns are along the x-axis and rows along the y-axis
## coords <- expand.grid(row=paste0("y", 1:platformDims["Y"]),
##                       col=paste0("x", 1:platformDims["X"]))
map <- data.frame(x=rep(1:platformDims["X"], each=platformDims["Y"]),
                  y=1:platformDims["Y"])
map$xy <- factor(paste0(map$x, "_", map$y))

## major borders
map$border <- FALSE
map$border[map$y == 11] <- TRUE # separation N+ vs N-
map$border[map$x %in% c(1,2, 159:160)] <- TRUE
map$border[map$x %in% c(73,74) & map$y < 11] <- TRUE # separation T vs NT
```

```{r}
## add managements
## nitrogen fertilization
map$ferti <- "N+"
map$ferti[map$y > 11] <- "N-"
## fungicide treatments
map$fungi <- "T"
map$fungi[map$x > 72 & map$y < 11] <- "NT"
map$mngt <- factor(paste(map$ferti, map$fungi),
                   levels=c("N+ T", "N+ NT", "N- T"))
map$ferti <- factor(map$ferti, levels=c("N+","N-"))
map$fungi <- factor(map$fungi, levels=c("T","NT"))
```

```{r}
## 7 series: A1, A2, A3, B1, B2, B3, C
## 40 genos per serie
## 2 blocks per serie x mngt
map$serie <- NA
## trials in the lower half: ferti=N+
## fungi=T
map$serie[map$x %in% 3:10 & map$y < 11] <- "A1"
map$border[map$x %in% c(11,12) & map$y < 11] <- TRUE
map$serie[map$x %in% 13:20 & map$y < 11] <- "A2"
map$border[map$x %in% c(21,22) & map$y < 11] <- TRUE
map$serie[map$x %in% 23:30 & map$y < 11] <- "A3"
map$border[map$x %in% c(31,32) & map$y < 11] <- TRUE
map$serie[map$x %in% 33:40 & map$y < 11] <- "B1"
map$border[map$x %in% c(41,42) & map$y < 11] <- TRUE
map$serie[map$x %in% 43:50 & map$y < 11] <- "B2"
map$border[map$x %in% c(51,52) & map$y < 11] <- TRUE
map$serie[map$x %in% 53:60 & map$y < 11] <- "B3"
map$border[map$x %in% c(61,62) & map$y < 11] <- TRUE
map$serie[map$x %in% 63:70 & map$y < 11] <- "C"
map$border[map$x %in% c(71,72) & map$y < 11] <- TRUE
## fungi=NT
map$serie[map$x %in% 75:82 & map$y < 11] <- "A1"
map$border[map$x %in% c(83,84) & map$y < 11] <- TRUE
map$serie[map$x %in% 85:92 & map$y < 11] <- "A2"
map$border[map$x %in% c(93,94) & map$y < 11] <- TRUE
map$serie[map$x %in% 95:102 & map$y < 11] <- "A3"
map$border[map$x %in% c(103,104) & map$y < 11] <- TRUE
map$serie[map$x %in% 105:112 & map$y < 11] <- "B1"
map$border[map$x %in% c(113,114) & map$y < 11] <- TRUE
map$serie[map$x %in% 115:122 & map$y < 11] <- "B2"
map$border[map$x %in% c(123,124) & map$y < 11] <- TRUE
map$serie[map$x %in% 125:132 & map$y < 11] <- "B3"
map$border[map$x %in% c(133,134) & map$y < 11] <- TRUE
map$serie[map$x %in% 135:142 & map$y < 11] <- "C"
map$border[map$x %in% c(143,144) & map$y < 11] <- TRUE

## trials in the upper half: ferti=N-
## fungi=T
map$serie[map$x %in% 3:22 & map$y > 11] <- "A1"
map$border[map$x %in% c(23,24) & map$y > 11] <- TRUE
map$serie[map$x %in% 25:44 & map$y > 11] <- "A2"
map$border[map$x %in% c(45,46) & map$y > 11] <- TRUE
map$serie[map$x %in% 47:66 & map$y > 11] <- "A3"
map$border[map$x %in% c(67,68) & map$y > 11] <- TRUE
map$serie[map$x %in% 69:88 & map$y > 11] <- "B1"
map$border[map$x %in% c(89,90) & map$y > 11] <- TRUE
map$serie[map$x %in% 91:110 & map$y > 11] <- "B2"
map$border[map$x %in% c(111,112) & map$y > 11] <- TRUE
map$serie[map$x %in% 113:132 & map$y > 11] <- "B3"
map$border[map$x %in% c(133,134) & map$y > 11] <- TRUE
map$serie[map$x %in% 135:154 & map$y > 11] <- "C"
map$border[map$x %in% c(155,156) & map$y > 11] <- TRUE

map$trial <- paste(map$serie,
                   as.character(map$ferti),
                   as.character(map$fungi),
                   sep="_")
map$trial[is.na(map$serie)] <- "border"
map$serie <- factor(map$serie)
map$trial <- factor(map$trial,
                    levels=c(sort(unique(map$trial[! is.na(map$serie)])),
                             "border"))
map$border[map$trial == "border"] <- TRUE
str(map)
```

```{r}
## blocks
## per trial in the lower half
map$block <- NA
trials <- unique(sort(as.character(map$trial[! map$border & map$y < 11])))
for(trial in trials){
  stopifnot(sum(map$trial == trial, na.rm=TRUE) == nbMicroPlotsPerTrial)
  map$block[which(map$trial == trial & map$y <= 5)] <- "A"
  map$block[which(map$trial == trial & map$y > 5)] <- "B"
}
trials <- unique(sort(as.character(map$trial[! map$border & map$y > 11])))
for(trial in trials){
  stopifnot(sum(map$trial == trial, na.rm=TRUE) == nbMicroPlotsPerTrial)
  rangeX <- range(map$x[map$trial == trial])
  midX <- rangeX[1] + ((rangeX[2] - rangeX[1] + 1) / 2) - 1
  map$block[which(map$trial == trial & map$x <= midX)] <- "A"
  map$block[which(map$trial == trial & map$x > midX)] <- "B"
}
map$block[map$border] <- "C"
map$block <- factor(map$block)
```

```{r, fig.width=12}
pMap <- ggplot(map, aes(x=x, y=y)) +
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  scale_x_continuous(breaks=sort(unique(map$x))) +
  scale_y_continuous(breaks=sort(unique(map$y))) +
  guides(x=guide_axis(angle=90))

pMap +
  labs(title="Map of the whole platform: managements") +
  geom_rect(aes(xmin=x-0.5, xmax=x+0.5,
                ymin=y-0.5, ymax=y+0.5, fill=mngt),
            color="grey", size=0.2)

pMap +
  labs(title="Map of the whole platform: trials",
       fill="Trials") +
  geom_rect(aes(xmin=x-0.5, xmax=x+0.5,
                ymin=y-0.5, ymax=y+0.5, fill=trial),
            color="grey", size=0.2)
## TODO: set border color to grey
```

## Genotypes

```{r}
## genotype IDs
genoBorder <- "g000"
levGenos <- c(sprintf(fmt=paste0("g%0", floor(log10(nbGenosInTrial))+1, "i"),
                      1:nbGenosInTrial),
              genoBorder)
length(levGenos)
```

```{r}
genosPerSerie <- list()
for(s in 1:nlevels(map$serie)){
  serie <- levels(map$serie)[s]
  idx <- (1+40*(s-1)):(1+40*(s-1)+40-1)
  genosPerSerie[[serie]] <- levGenos[idx]
}
sapply(genosPerSerie, length)
```

```{r}
map$geno <- NA
trials <- unique(sort(as.character(map$trial[! map$border])))
for(trial in trials){
  serie <- strsplit(trial, "_")[[1]][1]
  for(block in c("A","B")){
    idx <- which(map$trial == trial & map$block == block)
    stopifnot(length(idx) == nbGenosPerSerie)
    map$geno[idx] <- sample(genosPerSerie[[serie]], size=nbGenosPerSerie)
  }
}
map$geno[map$border] <- genoBorder
map$geno <- factor(map$geno, levels=levGenos)
```

## Design matrices

```{r}
## design matrices
X <- model.matrix(~ 1 + ferti + fungi, data=map
                , contrats.arg=list(ferti="contr.treatment", fungi="contr.treatment"))
## X <- model.matrix(~ 1 + ferti + fungi + trial/block, data=map)#, contrats.arg=list(ferti="contr.sum", fungi="contr.sum"))
dim(X)
colnames(X)

Z_g <- model.matrix(~ -1 + geno, data=map)
dim(Z_g)
colnames(Z_g) <- gsub("^geno", "", colnames(Z_g))

Z_spat <- model.matrix(~ -1 + xy, data=map)
dim(Z_spat)
```

## Effects

### Fixed

```{r}
mu <- 70 # intercept -> mean yield under ferti=N+ and fungi=T
betaFerti <- c("fertiN-"=-20)
betaFungi <- c("fungiNT"=-15)
(beta <- c("(Intercept)"=mu, betaFerti, betaFungi))
stopifnot(all(names(beta) == colnames(X)))
```

### Random

#### Genetic

1. Draw many genetic effects

2. Assume that A1 genotyeps are better on average than A2's, etc

```{r}
allGenoEffs <- rnorm(n=5*nbGenosInTrial, mean=0, sd=4)
summary(allGenoEffs)
allGenoEffs <- sort(allGenoEffs, decreasing=TRUE)
```

```{r}
lCumProbs <- list()
shapes <- data.frame(alpha=c(1, 1.3, 1.5, 2, 2.5, 3.5, 5),
                     beta=c( 3, 2.7, 2.4, 2, 1.7, 1.4, 1),
                     row.names=c("A1","A2","A3","B1","B2","B3","C"))#levSeries[1])
for(s in 1:nrow(shapes)){
  serie <- rownames(shapes)[s]
  tmp <- pbeta(seq(0,1,length.out=length(allGenoEffs)+1),
               shape1=shapes[serie,"alpha"],
               shape2=shapes[serie,"beta"])
  lCumProbs[[serie]] <- tmp
}
cumProbs <- as.data.frame(t(do.call(rbind, lCumProbs)))
cumProbs <- pivot_longer(cumProbs, cols=colnames(cumProbs), names_to="serie",
                      values_to="prob")
cumProbs <- as.data.frame(cumProbs)
cumProbs$serie <- factor(cumProbs$serie)
cumProbs$idx <- rep(1:(length(allGenoEffs)+1), each=nlevels(cumProbs$serie))
ggplot(cumProbs, aes(x=idx, y=prob, color=serie)) +
  labs(x="genetic effects sorted in decreasing order",
       y="cumulative probability to be sampled") +
  geom_line(size=1)

lProbs <- lapply(lCumProbs, diff)
probs <- as.data.frame(t(do.call(rbind, lProbs)))
probs <- pivot_longer(probs, cols=colnames(probs), names_to="serie",
                      values_to="prob")
probs <- as.data.frame(probs)
probs$serie <- factor(probs$serie)
probs$idx <- rep(1:length(allGenoEffs), each=nlevels(probs$serie))
ggplot(probs, aes(x=idx, y=prob, color=serie)) +
  labs(x="genetic effects sorted in decreasing order",
       y="probability to be sampled") +
  geom_line(size=1)
```

```{r}
lSerie2EffIdxs <- list()
idxToSampleFrom <- 1:length(allGenoEffs)
for(serie in levSeries){
  idx <- sample(x=idxToSampleFrom,
                size=nbGenosPerSerie,
                replace=FALSE,
                prob=lProbs[[serie]])
  lSerie2EffIdxs[[serie]] <- sort(idx)
  ## idxToSampleFrom <- idxToSampleFrom[! idxToSampleFrom %in% idx]
}

tmp <- sort(table(do.call(c, lSerie2EffIdxs)), decreasing=TRUE)
length(tmp)
table(tmp) # some genotypes were sampled multiple times
stopifnot(sum(table(tmp)) <= nbGenosInTrial)

lSerie2Effs <- lapply(lSerie2EffIdxs, function(idx){
  allGenoEffs[idx]
})
lSerie2Effs <- lapply(levSeries, function(serie){
  setNames(lSerie2Effs[[serie]], genosPerSerie[[serie]])
})
do.call(rbind, lapply(lSerie2Effs, summary))
names(lSerie2Effs) <- levSeries
```

```{r}
g <- c(do.call(c, lSerie2Effs), "g000"=0)
names(g) <- gsub("^.+\\.", "", names(g))
g <- scale(g, scale=F)[,1]
summary(g)
var(g)
sd(g)
stopifnot(all(names(g) == colnames(Z_g)))
```

#### Spatial

```{r}
rhoX <- 0.8
rhoY <- 0.5
corrMat_spat <- corrMatAR1(platformDims["X"], rhoX) %x% 
  corrMatAR1(platformDims["Y"], rhoY)
dim(corrMat_spat)
sigma2_ksi <- 50 # nugget
system.time(
    e_spat <- MASS::mvrnorm(n=1, mu=rep(0, ncol(Z_spat)),
                            Sigma=sigma2_ksi * corrMat_spat))
if(FALSE){
  st <- system.time(
      e_spat <- sparseMVN::rmvn.sparse(n=1, mu=rep(0, ncol(Z_spat)),
                                       CH=..., prec=TRUE))
  print(st)
}
summary(e_spat)
length(e_spat)
map$e_spat <- e_spat
```

```{r, fig.width=12}
pMap +
  labs(title="Map of the whole platform: spatial heterogeneity + errors") +
  geom_rect(aes(xmin=x-0.5, xmax=x+0.5,
                ymin=y-0.5, ymax=y+0.5, fill=e_spat),
            color="grey", size=0.2)

tmp <- list(col=min(map$x):max(map$x),
            row=min(map$y):max(map$y))
tmp$e_spat <- matrix(map$e_spat, byrow=FALSE,
                     nrow=length(tmp$row),
                     ncol=length(tmp$col),
                     dimnames=list(as.character(tmp$row), as.character(tmp$col)))
fields::image.plot(tmp$col, tmp$row, t(tmp$e_spat), col=topo.colors(100),
                   main="Map of the whole platform: spatial heterogeneity + errors",
                   xlab="x", ylab="y", las=1)
```

## Yield

```{r}
y <- X %*% beta + Z_g %*% g + Z_spat %*% e_spat
dat <- map
dat$yield <- y
```


# EDA

```{r, fig.width=10}
tmp <- tapply(dat$yield, dat$mngt, mean, na.rm=TRUE)
tmp <- data.frame(mngt=names(tmp),
                  mean=tmp)
tmp$mngt <- factor(tmp$mngt, levels=levels(dat$mngt))
tmp
ggplot(dat, aes(x=yield)) +
  labs(title="Yield", x="yield (in qt/ha)") +
  geom_histogram(bins=30) +
  geom_vline(aes(xintercept=mean, color=mngt), tmp, size=1.5)
```

```{r, fig.width=10}
ggplot(dat, aes(x=mngt, y=yield, group=mngt, color=mngt, fill=mngt)) +
  labs(title="Yield", x="", y="yield (in qt/ha)") +
  geom_violin(show.legend=FALSE)
```

```{r, fig.width=14}
ggplot(dat, aes(x=serie, y=yield, group=serie, color=serie, fill=serie)) +
  labs(title="Yield", x="", y="yield (in qt/ha)") +
  geom_violin(show.legend=FALSE) +
  facet_wrap(~ mngt)
```


# Per-trial analysis

```{r}
fitsSep <- lapply(trials, function(trial){
  subDat <- droplevels(dat[dat$trial == trial,])
  fit <- lm(yield ~ 1 + block + geno, data=subDat)
  if(FALSE){
    op <- par(mfrow=c(2,2))
    suppressWarnings(plot(fit, main=trial))
    par(op)
  }
  tav <- anova(fit)
  CV <- round((sqrt(tav["Residuals","Mean Sq"]) / mean(subDat$yield,na.rm=T)) * 100,
              dig=1)
  ETR <- round(sqrt(tav["Residuals","Mean Sq"]), dig=1)
  blockEffs <- emmeans(fit, ~ block)
  genoEffs <- emmeans(fit, ~ geno)
  list(fit=fit, tav=tav, CV=CV, ETR=ETR,
       blockEffs=blockEffs, genoEffs=genoEffs)
})
names(fitsSep) <- trials
```

```{r}
(tmp <- do.call(rbind,
                lapply(fitsSep, function(out){
                  c("CV"=out$CV, "ETR"=out$ETR)
                })
                ))
summary(tmp)
```


# Joint analysis

## Fit

```{r, fig.width=12, fig.height=10}
subDatSpats <- droplevels(dat[! dat$border,])
outF <- "corr-spat-het_SpATS.rds"
if(! file.exists(outF)){
  st <- system.time(
      fit <- SpATS(response="yield",
                   genotype="geno", genotype.as.random=TRUE,
                   spatial= ~ SAP(x, y, nseg=c(150, 15), pord=c(2,2), degree=c(3,3)),
                   fixed= ~ ferti + fungi + serie,
                   data=subDatSpats,
                   control=list(monitoring=0)))
  print(st)
  message("save file")
  saveRDS(fit, outF)
  print(tools::md5sum(outF))
} else{
  message("read file")
  print(tools::md5sum(outF))
  fit <- readRDS(outF)
}
sry <- summary(fit, which="variances")
(ETR <- sqrt(as.numeric(sry$p.table.vc["Residual", "Variance"])))
(H2 <- getHeritability(fit))
main <- paste0("Fit with SpATS:",
               " ETR=", round(ETR, 2), " H2=", H2)
plot(fit, main=main)
```

```{r}
pred_g <- predict(fit, which="geno")
g_blup <- setNames(pred_g$predicted.values, pred_g$geno)
g_blup <- g_blup[names(g)]
cor(g_blup, g, use="complete.obs")
plot(g_blup[names(g)], g, las=1)
fitg <- lm(g ~ g_blup[names(g)])
summary(fitg)$r.squared
abline(fitg, col="red")
```

## Extract spatial trend

```{r}
(minX <- min(map$x[! map$border]))
(maxX <- max(map$x[! map$border]))
spatTrend <- obtain.spatialtrend(fit, grid=c(maxX - minX + 1, platformDims["Y"]))
str(spatTrend$col.p)
range(spatTrend$col.p)
str(spatTrend$row.p)
range(spatTrend$row.p)
str(spatTrend$fit)
colnames(spatTrend$fit) <- as.character(spatTrend$col.p)
rownames(spatTrend$fit) <- as.character(spatTrend$row.p)
fields::image.plot(spatTrend$col.p, spatTrend$row.p, t(spatTrend$fit),
                   col=topo.colors(100), xlab="x", ylab="y", las=1)

if(FALSE){
  ## TODO: look inside plot.SpATS
  x.coord <- x$data[, xlab]
  y.coord <- x$data[, ylab]
  columns <- seq(min(x.coord), max(x.coord), by = min(diff(sort(unique(x.coord)))))
  rows <- seq(min(y.coord), max(y.coord), by = min(diff(sort(unique(y.coord)))))
  setNumericRounding(2)
  xy.coord <- data.table(expand.grid(columns = columns, rows = rows))
  setkeyv(xy.coord, c("rows", "columns"))
  ONE <- rep(1, length(x.coord))
  df <- data.table(columns = x.coord, rows = y.coord, response = response, 
                   fitted = fitted, residuals = residuals, geno.pred = geno.pred, 
                   weights = x$data$weights, ONE = ONE)
  setkeyv(df, c("rows", "columns"))
  df <- df[xy.coord]
  df <- df[order(df$columns, df$rows), ]
  p1 <- if (length(columns) > 100) 
          1
        else 100%/%length(columns) + 1
  p2 <- if (length(rows) > 100) 
          1
        else 100%/%length(rows) + 1
  fit.spatial.trend <- obtain.spatialtrend(x, grid = c(length(columns) * 
                                                       p1, length(rows) * p2))
  spatial.trend <- fit.spatial.trend$fit
  Mf <- kronecker(matrix(df$ONE, ncol = length(columns), nrow = length(rows)), 
                  matrix(1, p2, p1))
  colors <- topo.colors(100)
  fields::image.plot(fit.spatial.trend$col.p, fit.spatial.trend$row.p, 
                     t(spatial.trend * Mf), main = main.legends[4], col = colors, 
                     xlab = xlab, ylab = ylab, graphics.reset = TRUE, 
                     zlim = if (!is.null(dots$zlim)) 
                              dots$zlim
                            else range(spatial.trend))
}

tmp <- data.frame(x=rep(colnames(spatTrend$fit), each=nrow(spatTrend$fit)),
                  y=rownames(spatTrend$fit),
                  e_spat_pred=c(spatTrend$fit))
tmp$xy <- factor(paste0(tmp$x, "_", tmp$y))
tmp$e_spat <- map$e_spat[match(as.character(tmp$xy), as.character(map$xy))]
head(tmp)
plot(tmp$e_spat_pred, tmp$e_spat)
subDatSpats$residuals <- residuals(fit)
tmp$residuals <- subDatSpats$residuals[match(as.character(tmp$xy), as.character(subDatSpats$xy))]
plot(tmp$e_spat_pred, tmp$e_spat - tmp$residuals)

dat2 <- dat
dat2$e_spat_pred <- tmp$e_spat_pred[match(as.character(dat2$xy), as.character(tmp$xy))]
```

## Re-fit per trial

```{r}
fitsSep2 <- lapply(trials, function(trial){
  subDat <- droplevels(dat2[dat2$trial == trial,])
  fit <- lm(yield ~ 1 + block + geno, data=subDat)
  if(FALSE){
    op <- par(mfrow=c(2,2))
    suppressWarnings(plot(fit, main=trial))
    par(op)
  }
  tav <- anova(fit)
  CV <- round((sqrt(tav["Residuals","Mean Sq"]) / mean(dat2$yield,na.rm=T)) * 100,
              dig=1)
  ETR <- round(sqrt(tav["Residuals","Mean Sq"]), dig=1)
  blockEffs <- emmeans(fit, ~ block)
  genoEffs <- emmeans(fit, ~ geno)
  list(fit=fit, tav=tav, CV=CV, ETR=ETR, blockEffs=blockEffs, genoEffs=genoEffs)
})
names(fitsSep2) <- trials
```

```{r}
(tmp <- do.call(rbind,
                lapply(fitsSep2, function(out){
                  c("CV"=out$CV, "ETR"=out$ETR)
                })
                ))
summary(tmp)
```


# References

* Rodríguez-Álvarez MX, Boer MP, van Eeuwijk FA, Eilers PHC. 2018. Correcting for spatial heterogeneity in plant breeding experiments with P-splines. Spatial Statistics. 23:52–71. doi:10.1016/j.spasta.2017.10.003.



# Appendix

```{r info}
t1 <- proc.time()
t1 - t0
print(sessionInfo(), locale=FALSE)
```
