---
title: "Estimating heritability"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
```


# Overview

This document aims at estimating heritability (broad or narrow sense).

This document requires external packages to be available:
```{r load_pkg}
suppressPackageStartupMessages(library(mvtnorm)) # on the CRAN
suppressPackageStartupMessages(library(lme4)) # on the CRAN
suppressPackageStartupMessages(library(INLA)) # online
nb.threads <- 3 # 1
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Model

Let's take as an example multiple genotypes phenotyped several years in a row:

* $I$: number of genotypes

* $Q$: number of years of phenotyping

* $N = I \times Q$: number of phenotypic observations

Likelihood:

\[
\forall q \in \{1,\ldots,Q\}, i \in \{1,\ldots,I\}, \; y_{qi} = \beta_q + g_i + \epsilon_{qi}
\]

with the $\beta$'s being modeled as fixed effects, $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 \text{Id}_I)$ and $\boldsymbol{\epsilon} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\epsilon^2 \text{Id}_N)$

It can also be rewritten as $\boldsymbol{y} = X \boldsymbol{\beta} + Z \boldsymbol{g} + \boldsymbol{\epsilon}$ where $X$ and $Z$ are incidence matrices.

The main goal here is to estimate the (broad-sense) heritability $H^2 = \frac{\sigma_g^2}{\sigma_g^2 + \sigma_\epsilon^2}$ and quantify its uncertainty.


# Simulation

Set the seed:
```{r set_seed}
set.seed(1859)
```

Set the constants:
```{r set_constants}
I <- 200
Q <- 5
N <- I * Q
sigma.g <- 3
sigma.epsilon <- 3
sigma.g^2 / (sigma.g^2 + sigma.epsilon^2)
```

Simulate the explanatory and error variables:
```{r simul_var}
beta <- matrix(c(50, rmvnorm(n=1, mean=rep(0, Q-1), sigma=20 * diag(Q-1))))
g <- t(rmvnorm(n=1, mean=rep(0, I), sigma=sigma.g^2 * diag(I)))
epsilon <- t(rmvnorm(n=1, mean=rep(0, N), sigma=sigma.epsilon^2 * diag(N)))
```

Make a data frame with the appropriate structure:
```{r make_df}
geno.names <- sprintf("geno%03i", 1:I)
dat <- data.frame(geno=rep(geno.names, Q),
                  year=as.factor(rep(2010:(2010+Q-1), each=I)),
                  pheno=NA)
```

Make the incidence matrices:
```{r}
X <- model.matrix(~ 1 + year, data=dat)
Z <- model.matrix(~ -1 + geno, data=dat)
```

Make the phenotypes and fill the data frame:
```{r}
y <- X %*% matrix(beta) + Z %*% matrix(g) + epsilon
dat$pheno <- c(y)
str(dat)
```


# Data exploration

```{r}
tapply(dat$pheno, list(dat$year), summary)
```

```{r}
boxplot(pheno ~ year, data=dat, las=1, varwidth=TRUE,
        xlab="years", ylab="phenotypes",
        main="Simulated data")
abline(h=mean(dat$pheno), lty=2)
```


# Inference

## With lme4

### Model fitting

```{r lmer_fit}
system.time(
    fit.lmer.g <- lmer(formula=pheno ~ 1 + (1|geno),
                       data=dat, REML=TRUE))
system.time(
    fit.lmer.y.g <- lmer(formula=pheno ~ 1 + year + (1|geno),
                         data=dat, REML=TRUE))
```

### Model comparison and selection

```{r lmer_select}
extractAIC(fit.lmer.g)
extractAIC(fit.lmer.y.g)
fit.lmer <- fit.lmer.y.g
```

### Diagnostics

```{r lmer_orga}
beta.hat.lmer <- fixef(fit.lmer)
g.hat.lmer <- ranef(fit.lmer, condVar=TRUE, drop=TRUE)$geno
var.g.hat.lmer <- setNames(attr(g.hat.lmer, "postVar"),
                           names(g.hat.lmer))
sigma.epsilon.hat.lmer <- as.data.frame(VarCorr(fit.lmer))[2, "sdcor"]
sigma.g.hat.lmer <- as.data.frame(VarCorr(fit.lmer))[1, "sdcor"]
fit.lmer.all <- cbind(dat,
                      cond.res=residuals(fit.lmer),
                      scl.cond.res=residuals(fit.lmer) / sigma(fit.lmer),
                      fitted=fitted(fit.lmer))
str(fit.lmer.all, give.attr=FALSE)
```

```{r lmer_check}
(low.bound.norm.95 <- qnorm(p=0.025, mean=0, sd=1, lower.tail=TRUE))
x.lim <- max(abs(fit.lmer.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.lmer.all$scl.cond.res, y=fit.lmer.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.lmer.all$scl.cond.res)
qqnorm(y=fit.lmer.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.lmer.all$scl.cond.res, col="red")
x.lim <- max(abs(g.hat.lmer))
par(mar=c(5,6,4,1))
plot(x=g.hat.lmer, y=1:length(g.hat.lmer),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(g.hat.lmer), labels=names(g.hat.lmer),
     las=1)
abline(v=0, lty=2)
```

### Inference

```{r lmer_res}
summary(fit.lmer)
cbind(beta, beta.hat.lmer)
cor(g, g.hat.lmer)
c(sigma.epsilon, sigma.epsilon.hat.lmer)
c(sigma.g, sigma.g.hat.lmer)
system.time(
    prof <- profile(fitted=fit.lmer, signames=FALSE))
(ci.lmer <- confint(object=prof, level=0.95))
```

Confidence interval on heritability:

* see Table S2 from Bhatia et al (2014)

* how to implement the approach of Meyer and Houle (2013) with `lmer`?


## With INLA

### Model formulations

See [Holand *et al* (2013)](http://dx.doi.org/10.1534/g3.113.006700).

* model formulation 1 (MF1): used to compare and select models, and estimate $\sigma_\epsilon^2$

    1. likelihood: $y_{qi} | \eta_{qi} \sim \mathcal{N}(\eta_{qi}, \sigma_\epsilon^2)$
    
    2. latent field: $\eta_{qi} = \beta_q + g_i + e_{qi}$ with $\boldsymbol{\beta} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\beta^2 \text{Id}_Q)$, $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 \text{Id}_I)$, and $\boldsymbol{e} \sim \mathcal{N}(\boldsymbol{0}, \sigma_e^2 \text{Id}_N)$
    
    3. prior on hyperparameters: $\sigma_\beta^2$ is fixed at a large value ($2.2 \times 10^4 \approx 1/\exp(-10)$), $\sigma_e^2$ is fixed at a small value ($4.5 \times 10^{-5} \approx 1/\exp(10)$), $\sigma_g^2 \text{ and } \sigma_\epsilon^2 \sim \mathcal{IG}(0.5, 0.5)$
    
* model formulation 2 (MF2): used to estimate heritability

    1. likelihood: $y_{qi} | \eta_{qi} \sim \mathcal{N}(\eta_{qi}, \sigma_e^2)$
    
    2. latent field: $\eta_{qi} = \beta_q + g_i + \epsilon_{qi}$ with $\boldsymbol{\beta} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\beta^2 \text{Id}_Q)$, $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 \text{Id}_I)$, and $\boldsymbol{\epsilon} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\epsilon^2 \text{Id}_N)$
    
    3. prior on hyperparameters: same as MF1
    

### Model fitting

```{r inla_fit_prep_dat}
dat.inla <- cbind(dat, obs.id=paste(dat$geno, dat$year, sep="-"))
```

#### MF1

```{r inla_fit_MF1}
system.time(
    fit.inla.MF1.g <- inla(
        formula=pheno ~ 1 +
          f(geno, model="iid", constr=TRUE,
            hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(obs.id, model="iid", constr=TRUE,
            hyper=list(theta=list(initial=10, fixed=TRUE))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(param=c(0.5, 0.5)))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
system.time(
    fit.inla.MF1.y.g <- inla(
        formula=pheno ~ 1 +
          year +
          ## f(year, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(initial=-10, fixed=TRUE))) +
          f(geno, model="iid", constr=TRUE,
            hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(obs.id, model="iid", constr=TRUE,
            hyper=list(theta=list(initial=10, fixed=TRUE))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(param=c(0.5, 0.5)))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
```

#### MF2

```{r inla_fit_MF2}
system.time(
    fit.inla.MF2.g <- inla(
        formula=pheno ~ 1 +
          f(geno, model="iid", constr=TRUE,
            hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(obs.id, model="iid", constr=TRUE,
            hyper=list(theta=list(param=c(0.5, 0.5)))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(initial=10, fixed=TRUE))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
system.time(
    fit.inla.MF2.y.g <- inla(
        formula=pheno ~ 1 +
          year +
          ## f(year, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(initial=-10, fixed=TRUE))) +
          f(geno, model="iid", constr=TRUE,
            hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(obs.id, model="iid", constr=TRUE,
            hyper=list(theta=list(param=c(0.5, 0.5)))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(initial=10, fixed=TRUE))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
```

### Model comparison and selection

Posterior predictive checks: see Held, Schrödle and Rue (2010), and use `inla.pmarginal()`?

#### MF1

```{r inla_select_MF1}
c(fit.inla.MF1.g$waic$p.eff, fit.inla.MF1.g$waic$waic)
c(fit.inla.MF1.y.g$waic$p.eff, fit.inla.MF1.y.g$waic$waic)
fit.inla.MF1 <- fit.inla.MF1.y.g
```

#### MF2

As expected, using MF2 to compare models via likelihood-based metrics doesn't work:
```{r inla_select_MF2, eval=TRUE}
c(fit.inla.MF2.g$waic$p.eff, fit.inla.MF2.g$waic$waic)
c(fit.inla.MF2.y.g$waic$p.eff, fit.inla.MF2.y.g$waic$waic)
fit.inla.MF2 <- fit.inla.MF2.y.g
```

### Diagnostics

#### MF1

```{r inla_orga_MF1}
beta.hat.inla.MF1 <- setNames(fit.inla.MF1$summary.fixed[,"mean"],
                              rownames(fit.inla.MF1$summary.fixed))
g.hat.inla.MF1 <- setNames(fit.inla.MF1$summary.random$geno[,"mean"],
                           fit.inla.MF1$summary.random$geno[,"ID"])
idx <- grep("Gaussian observations", rownames(fit.inla.MF1$summary.hyperpar))
sigma.epsilon.hat.inla.MF1 <- 1 / sqrt(fit.inla.MF1$summary.hyperpar[idx, "mean"])
idx <- grep("geno", rownames(fit.inla.MF1$summary.hyperpar))
sigma.g.hat.inla.MF1 <- 1 / sqrt(fit.inla.MF1$summary.hyperpar[idx, "mean"])
fit.inla.MF1.all <- cbind(dat,
                          fitted=fit.inla.MF1$summary.linear.predictor[,"mean"],
                          cond.res=NA,
                          scl.cond.res=NA)
fit.inla.MF1.all$cond.res <- dat$pheno - fit.inla.MF1.all$fitted
fit.inla.MF1.all$scl.cond.res <- fit.inla.MF1.all$cond.res / rep(sigma.g.hat.inla.MF1, N)
str(fit.inla.MF1.all, give.attr=FALSE)
```

```{r inla_check_MF1}
(low.bound.norm.95 <- qnorm(p=0.025, mean=0, sd=1, lower.tail=TRUE))
x.lim <- max(abs(fit.inla.MF1.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.inla.MF1.all$scl.cond.res, y=fit.inla.MF1.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.inla.MF1.all$scl.cond.res)
qqnorm(y=fit.inla.MF1.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.inla.MF1.all$scl.cond.res, col="red")
x.lim <- max(abs(g.hat.inla.MF1))
par(mar=c(5,6,4,1))
plot(x=g.hat.inla.MF1, y=1:length(g.hat.inla.MF1),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(g.hat.inla.MF1), labels=names(g.hat.inla.MF1),
     las=1)
abline(v=0, lty=2)
```

#### MF2

```{r inla_orga_MF2}
beta.hat.inla.MF2 <- setNames(fit.inla.MF2$summary.fixed[,"mean"],
                              rownames(fit.inla.MF2$summary.fixed))
g.hat.inla.MF2 <- setNames(fit.inla.MF2$summary.random$geno[,"mean"],
                           fit.inla.MF2$summary.random$geno[,"ID"])
idx <- grep("obs.id", rownames(fit.inla.MF2$summary.hyperpar))
sigma.epsilon.hat.inla.MF2 <- 1 / sqrt(fit.inla.MF2$summary.hyperpar[idx, "mean"])
idx <- grep("geno", rownames(fit.inla.MF2$summary.hyperpar))
sigma.g.hat.inla.MF2 <- 1 / sqrt(fit.inla.MF2$summary.hyperpar[idx, "mean"])
fit.inla.MF2.all <- cbind(dat,
                          fitted=fit.inla.MF2$summary.linear.predictor[,"mean"],
                          cond.res=NA,
                          scl.cond.res=NA)
fit.inla.MF2.all$cond.res <- dat$pheno - fit.inla.MF2.all$fitted
fit.inla.MF2.all$scl.cond.res <- fit.inla.MF2.all$cond.res / rep(sigma.g.hat.inla.MF2, N)
str(fit.inla.MF2.all, give.attr=FALSE)
```

```{r inla_check_MF2}
(low.bound.norm.95 <- qnorm(p=0.025, mean=0, sd=1, lower.tail=TRUE))
x.lim <- max(abs(fit.inla.MF2.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.inla.MF2.all$scl.cond.res, y=fit.inla.MF2.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.inla.MF2.all$scl.cond.res)
qqnorm(y=fit.inla.MF2.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.inla.MF2.all$scl.cond.res, col="red")
x.lim <- max(abs(g.hat.inla.MF2))
par(mar=c(5,6,4,1))
plot(x=g.hat.inla.MF2, y=1:length(g.hat.inla.MF2),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(g.hat.inla.MF2), labels=names(g.hat.inla.MF2),
     las=1)
abline(v=0, lty=2)
```

### Inference

#### MF1

```{r inla_res_MF1}
summary(fit.inla.MF1)
cbind(beta, beta.hat.inla.MF1)
cor(g, g.hat.inla.MF1)
c(sigma.epsilon, sigma.epsilon.hat.inla.MF1)
c(sigma.g, sigma.g.hat.inla.MF1)
```

#### MF2

```{r inla_res_MF2}
summary(fit.inla.MF2)
cbind(beta, beta.hat.inla.MF2)
cor(g, g.hat.inla.MF2)
c(sigma.epsilon, sigma.epsilon.hat.inla.MF2)
c(sigma.g, sigma.g.hat.inla.MF2)
```

TODO: retrieve inference results for heritability


## With MCMCglmm

TODO, but potentially much slower



# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
