---
title: "Estimating heritability"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
```


# Overview

This document aims at estimating heritability (broad or narrow sense).

This document requires external packages to be available:
```{r load_pkg}
suppressPackageStartupMessages(library(mvtnorm)) # on the CRAN
suppressPackageStartupMessages(library(lme4)) # on the CRAN
suppressPackageStartupMessages(library(INLA)) # online
suppressPackageStartupMessages(library(breedR)) # on GitHub; 'master' >= 13/10/2017
nb.threads <- 3 # 1
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Model

## Notations

Assuming a plant species, let's take as an example multiple genotypes phenotyped for a single trait several years in a row on a single site (each genotype may be planted several times):

* $I$: number of genotypes ($\approx 200$);

* $J$: number of years of phenotyping ($\approx 5$);

* $K$: number of replicates of each genotype ($\approx 1-3$);

* $N = I \times J \ times K$: total number of phenotypic observations;

* $y_{ijk}$: phenotypic value of the $k$-th replicate of the $i$-th genotype the $j$-th year

* $\beta_j$: effect of the $j$-th year, modeled as a fixed effect

* $u_i$: genotypic value of the $i$-th genotype, modeled as a random variable

* $\epsilon_{ijk}$: error corresponding to the $k$-th replicate of the $i$-th genotype the $j$-th year

## Model 1

\[
Y = X B + Z U + E
\]

with:

* $U \sim \mathcal{MN}(\boldsymbol{0}, A, \Sigma_g)$;

* $E \sim \mathcal{MN}(\boldsymbol{0}, \text{Id}, \Sigma_\epsilon)$.

The main goal here is to estimate "heritabilities" ($\forall j, \; \frac{\Sigma_g^(j)}{\Sigma_g^(j) + \Sigma_\epsilon^(j)}$) and quantify their uncertainties.

## Model 2

Assume no covariance between years.

Likelihood:

\[
\mathcal{M}_1: \; \forall q \in \{1,\ldots,Q\}, i \in \{1,\ldots,I\}, \; y_{qi} = \beta_q + u_i + \epsilon_{qi}
\]

with:

* $\boldsymbol{u} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 A)$;

* $\boldsymbol{\epsilon} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\epsilon^2 \text{Id})$.

also rewritten as:

\[
\boldsymbol{y} = X \boldsymbol{\beta} + Z \boldsymbol{u} + \boldsymbol{\epsilon}
\]

where $X$ and $Z$ are appropriate incidence matrices.

The main goal here is to estimate heritability $\frac{\sigma_g^2}{\sigma_g^2 + \sigma_\epsilon^2}$ and quantify its uncertainty.


# Simulation

Set the seed:
```{r set_seed}
set.seed(1859)
```

Set the constants:
```{r set_constants}
I <- 200
J <- 5
K <- 2
N <- I * J * K
geno.names <- sprintf("geno%03i", 1:I)
year.ids <- paste0("Y", 2011:(2011+J-1))
rep.ids <- paste0("rep", 1:K)
```

Make a data frame with the appropriate structure:
```{r make_df}
dat <- data.frame(geno=rep(rep(geno.names, each=K), J),
                  year=rep(year.ids, each=I*K),
                  rep=rep(rep.ids, I*J),
                  pheno=NA)
```

Turn it into a matrix, with years in columns:
```{r}
mat <- matrix(data=NA, nrow=I*K, ncol=J)
colnames(mat) <- year.ids
rownames(mat) <- paste(dat$geno[dat$year == "Y2011"],
                       dat$rep[dat$year == "Y2011"], sep="_")
```

Make the incidence matrices:
```{r}
X <- model.matrix(~ 1 + year, data=dat)
colnames(X) <- year.ids
Z <- model.matrix(~ -1 + geno, data=dat)
colnames(Z) <- geno.names
```

Choose the (co)variances:
```{r}
Sigma.g <- diag(rep(c(5,2), length=J))
Sigma.epsilon <- diag(rep(3, length=J))
(h2 <- diag(Sigma.g) / (diag(Sigma.g) + diag(Sigma.epsilon)))
```

Simulate the explanatory and error variables, assuming for simplicity an identity matrix for $A$:
```{r simul_var}
B <- matrix(c(50, rmvnorm(n=1, mean=rep(0, Q-1), sigma=20 * diag(Q-1))))
A <- diag(I) # for simplicity
dimnames(A) <- list(geno.names, geno.names)
u <- t(rmvnorm(n=1, mean=rep(0, I), sigma=sigma.g^2 * A))
epsilon <- t(rmvnorm(n=1, mean=rep(0, N), sigma=sigma.epsilon^2 * diag(N)))
```

Make the phenotypes and fill the data frame:
```{r}
y <- X %*% matrix(beta) + Z %*% matrix(u) + epsilon
dat$pheno <- c(y)
str(dat)
```


# Data exploration

```{r}
tapply(dat$pheno, list(dat$year), summary)
```

```{r}
boxplot(pheno ~ year, data=dat, las=1, varwidth=TRUE,
        xlab="years", ylab="phenotypes",
        main="Simulated data")
abline(h=mean(dat$pheno), lty=2)
```

```{r}
mat <- do.call(cbind, lapply(unique(dat$year), function(year){
  dat$pheno[dat$year == year]
}))
colnames(mat) <- unique(dat$year)
rownames(mat) <- dat$geno[dat$year == unique(dat$year)[1]]
pairs(x=mat)
```


# Inference

```{r}
(low.bound.norm.95 <- qnorm(p=0.025, mean=0, sd=1, lower.tail=TRUE))
```

## With lme4

* can't provide confidence interval for heritability

* can handle genotype-year interactions

### Model fitting

```{r lmer_fit}
system.time(
    fit.lmer.g <- lmer(formula=pheno ~ 1 + (1|geno),
                       data=dat, REML=TRUE))
system.time(
    fit.lmer.y.g <- lmer(formula=pheno ~ 1 + year + (1|geno),
                         data=dat, REML=TRUE))
system.time(
    fit.lmer.yg <- lmer(formula=pheno ~ 1 + year + (1|geno) + (1|geno:year),
                        data=dat, REML=TRUE,
                        control=lmerControl(check.nobs.vs.nlev="ignore",
                                            check.nobs.vs.nRE="ignore")))
```

### Model comparison and selection

```{r lmer_select}
extractAIC(fit.lmer.g)
extractAIC(fit.lmer.y.g)
extractAIC(fit.lmer.yg)
fit.lmer <- fit.lmer.y.g
```

### Diagnostics

```{r lmer_orga}
beta.hat.lmer <- fixef(fit.lmer)
u.hat.lmer <- ranef(fit.lmer, condVar=TRUE, drop=TRUE)$geno
var.u.hat.lmer <- setNames(attr(u.hat.lmer, "postVar"),
                           names(u.hat.lmer))
sigma.epsilon.hat.lmer <- as.data.frame(VarCorr(fit.lmer))[2, "sdcor"]
sigma.u.hat.lmer <- as.data.frame(VarCorr(fit.lmer))[1, "sdcor"]
fit.lmer.all <- cbind(dat,
                      cond.res=residuals(fit.lmer),
                      scl.cond.res=residuals(fit.lmer) / sigma(fit.lmer),
                      fitted=fitted(fit.lmer))
str(fit.lmer.all, give.attr=FALSE)
```

```{r lmer_check}
x.lim <- max(abs(fit.lmer.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.lmer.all$scl.cond.res, y=fit.lmer.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.lmer.all$scl.cond.res)
qqnorm(y=fit.lmer.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.lmer.all$scl.cond.res, col="red")
x.lim <- max(abs(u.hat.lmer))
par(mar=c(5,6,4,1))
plot(x=u.hat.lmer, y=1:length(u.hat.lmer),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(u.hat.lmer), labels=names(u.hat.lmer),
     las=1)
abline(v=0, lty=2)
```

### Inference

```{r lmer_res}
summary(fit.lmer)
cbind(beta, beta.hat.lmer)
cor(g, u.hat.lmer)
c(sigma.epsilon, sigma.epsilon.hat.lmer)
c(sigma.g, sigma.u.hat.lmer)
system.time(
    prof <- profile(fitted=fit.lmer, signames=FALSE))
(ci.lmer <- confint(object=prof, level=0.95))
```

Confidence interval on heritability: it doesn't seem to currently be possible

* Bhatia et al (2014), in their Table S2, use an ad hoc procedure

* how to implement the approach of Meyer and Houle (2013) with `lmer`?


## With INLA

### Model formulations

See [Holand *et al* (2013)](http://dx.doi.org/10.1534/g3.113.006700).

* model formulation 1 (MF1): used to compare and select models, and estimate $\sigma_\epsilon^2$

    1. likelihood: $y_{qi} | \eta_{qi} \sim \mathcal{N}(\eta_{qi}, \sigma_\epsilon^2)$
    
    2. latent field: $\eta_{qi} = \beta_q + g_i + e_{qi}$ with $\boldsymbol{\beta} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\beta^2 \text{Id}_Q)$, $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 \text{Id}_I)$, and $\boldsymbol{e} \sim \mathcal{N}(\boldsymbol{0}, \sigma_e^2 \text{Id}_N)$
    
    3. prior on hyperparameters: $\sigma_\beta^2$ is fixed at a large value ($2.2 \times 10^4 \approx 1/\exp(-10)$), $\sigma_e^2$ is fixed at a small value ($4.5 \times 10^{-5} \approx 1/\exp(10)$), $\sigma_g^2 \text{ and } \sigma_\epsilon^2 \sim \mathcal{IG}(0.5, 0.5)$
    
* model formulation 2 (MF2): used to estimate heritability

    1. likelihood: $y_{qi} | \eta_{qi} \sim \mathcal{N}(\eta_{qi}, \sigma_e^2)$
    
    2. latent field: $\eta_{qi} = \beta_q + g_i + \epsilon_{qi}$ with $\boldsymbol{\beta} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\beta^2 \text{Id}_Q)$, $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 \text{Id}_I)$, and $\boldsymbol{\epsilon} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\epsilon^2 \text{Id}_N)$
    
    3. prior on hyperparameters: same as MF1
    

### Model fitting

```{r inla_fit_prep_dat}
dat.inla <- cbind(dat,
                  geno.id=as.numeric(gsub("geno", "", dat$geno)),
                  obs.id=paste(dat$geno, dat$year, sep="-"))
```

#### MF1

```{r inla_fit_MF1}
system.time(
    fit.inla.MF1.g <- inla(
        formula=pheno ~ 1 +
          ## f(geno, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(geno.id, model="generic0", constr=TRUE,
            Cmatrix=solve(A),
            hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(obs.id, model="iid", constr=TRUE,
            hyper=list(theta=list(initial=10, fixed=TRUE))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(param=c(0.5, 0.5)))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
system.time(
    fit.inla.MF1.y.g <- inla(
        formula=pheno ~ 1 +
          year +
          ## f(year, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(initial=-10, fixed=TRUE))) +
          ## f(geno, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(geno.id, model="generic0", constr=TRUE,
            Cmatrix=A,
            hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(obs.id, model="iid", constr=TRUE,
            hyper=list(theta=list(initial=10, fixed=TRUE))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(param=c(0.5, 0.5)))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
```

#### MF2

```{r inla_fit_MF2}
system.time(
    fit.inla.MF2.g <- inla(
        formula=pheno ~ 1 +
          ## f(geno, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(geno.id, model="generic2", constr=TRUE,
            Cmatrix=A,
            hyper=list(theta1=list(param=c(0.5, 0.5)),
                       theta2=list(param=c(0.5, 0.5)))),# +
          ## f(obs.id, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(param=c(0.5, 0.5)))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(initial=10, fixed=TRUE))),
        data=dat.inla,
        ## control.compute=list(waic=TRUE),
        control.compute=list(waic=FALSE),
        num.threads=nb.threads))
system.time(
    fit.inla.MF2.y.g <- inla(
        formula=pheno ~ 1 +
          year +
          ## f(year, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(initial=-10, fixed=TRUE))) +
          ## f(geno, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(param=c(0.5, 0.5)))) +
          f(geno.id, model="generic2", constr=TRUE,
            Cmatrix=A,
            hyper=list(theta1=list(param=c(0.5, 0.5)),
                       theta2=list(param=c(0.5, 0.5)))),# +
          ## f(obs.id, model="iid", constr=TRUE,
          ##   hyper=list(theta=list(param=c(0.5, 0.5)))),
        family="gaussian",
        control.family=list(hyper=list(theta=list(initial=10, fixed=TRUE))),
        data=dat.inla,
        control.compute=list(waic=TRUE),
        num.threads=nb.threads))
```

### Model comparison and selection

Posterior predictive checks: see Held, Schrödle and Rue (2010), and use `inla.pmarginal()`?

#### MF1

```{r inla_select_MF1}
c(fit.inla.MF1.g$waic$p.eff, fit.inla.MF1.g$waic$waic)
c(fit.inla.MF1.y.g$waic$p.eff, fit.inla.MF1.y.g$waic$waic)
fit.inla.MF1 <- fit.inla.MF1.y.g
```

#### MF2

As expected, using MF2 to compare models via likelihood-based metrics doesn't work:
```{r inla_select_MF2, eval=TRUE}
c(fit.inla.MF2.g$waic$p.eff, fit.inla.MF2.g$waic$waic)
c(fit.inla.MF2.y.g$waic$p.eff, fit.inla.MF2.y.g$waic$waic)
fit.inla.MF2 <- fit.inla.MF2.y.g
```

### Diagnostics

#### MF1

```{r inla_orga_MF1}
beta.hat.inla.MF1 <- setNames(fit.inla.MF1$summary.fixed[,"mean"],
                              rownames(fit.inla.MF1$summary.fixed))
u.hat.inla.MF1 <- setNames(fit.inla.MF1$summary.random$geno[,"mean"],
                           fit.inla.MF1$summary.random$geno[,"ID"])
idx <- grep("Gaussian observations", rownames(fit.inla.MF1$summary.hyperpar))
sigma.epsilon.hat.inla.MF1 <- 1 / sqrt(fit.inla.MF1$summary.hyperpar[idx, "mean"])
idx <- grep("geno", rownames(fit.inla.MF1$summary.hyperpar))
sigma.u.hat.inla.MF1 <- 1 / sqrt(fit.inla.MF1$summary.hyperpar[idx, "mean"])
fit.inla.MF1.all <- cbind(dat,
                          fitted=fit.inla.MF1$summary.linear.predictor[,"mean"],
                          cond.res=NA,
                          scl.cond.res=NA)
fit.inla.MF1.all$cond.res <- dat$pheno - fit.inla.MF1.all$fitted
fit.inla.MF1.all$scl.cond.res <- fit.inla.MF1.all$cond.res / rep(sigma.u.hat.inla.MF1, N)
str(fit.inla.MF1.all, give.attr=FALSE)
```

```{r inla_check_MF1}
x.lim <- max(abs(fit.inla.MF1.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.inla.MF1.all$scl.cond.res, y=fit.inla.MF1.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.inla.MF1.all$scl.cond.res)
qqnorm(y=fit.inla.MF1.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.inla.MF1.all$scl.cond.res, col="red")
x.lim <- max(abs(u.hat.inla.MF1))
par(mar=c(5,6,4,1))
plot(x=u.hat.inla.MF1, y=1:length(u.hat.inla.MF1),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(u.hat.inla.MF1), labels=names(u.hat.inla.MF1),
     las=1)
abline(v=0, lty=2)
```

#### MF2

```{r inla_orga_MF2}
beta.hat.inla.MF2 <- setNames(fit.inla.MF2$summary.fixed[,"mean"],
                              rownames(fit.inla.MF2$summary.fixed))
u.hat.inla.MF2 <- setNames(fit.inla.MF2$summary.random$geno.id[,"mean"],
                           fit.inla.MF2$summary.random$geno.id[,"ID"])
idx <- grep("obs.id", rownames(fit.inla.MF2$summary.hyperpar))
sigma.epsilon.hat.inla.MF2 <- 1 / sqrt(fit.inla.MF2$summary.hyperpar[idx, "mean"])
idx <- grep("Precision-cmatrix for geno.id",
            rownames(fit.inla.MF2$summary.hyperpar))
sigma.u.hat.inla.MF2 <- 1 / sqrt(fit.inla.MF2$summary.hyperpar[idx, "mean"])
idx <- grep("h2 for geno.id",
            rownames(fit.inla.MF2$summary.hyperpar))
h2.hat.inla.MF2 <- fit.inla.MF2$summary.hyperpar[idx, "mean"]
fit.inla.MF2.all <- cbind(dat,
                          fitted=fit.inla.MF2$summary.linear.predictor[,"mean"],
                          cond.res=NA,
                          scl.cond.res=NA)
fit.inla.MF2.all$cond.res <- dat$pheno - fit.inla.MF2.all$fitted
fit.inla.MF2.all$scl.cond.res <- fit.inla.MF2.all$cond.res / rep(sigma.u.hat.inla.MF2, N)
str(fit.inla.MF2.all, give.attr=FALSE)
```

```{r inla_check_MF2}
x.lim <- max(abs(fit.inla.MF2.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.inla.MF2.all$scl.cond.res, y=fit.inla.MF2.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.inla.MF2.all$scl.cond.res)
qqnorm(y=fit.inla.MF2.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.inla.MF2.all$scl.cond.res, col="red")
x.lim <- max(abs(u.hat.inla.MF2))
par(mar=c(5,6,4,1))
plot(x=u.hat.inla.MF2, y=1:length(u.hat.inla.MF2),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(u.hat.inla.MF2), labels=names(u.hat.inla.MF2),
     las=1)
abline(v=0, lty=2)
```

### Inference

#### MF1

```{r inla_res_MF1}
summary(fit.inla.MF1)
cbind(beta, beta.hat.inla.MF1)
cor(g, u.hat.inla.MF1)
c(sigma.epsilon, sigma.epsilon.hat.inla.MF1)
c(sigma.g, sigma.u.hat.inla.MF1)
```

#### MF2

```{r inla_res_MF2}
summary(fit.inla.MF2)
cbind(beta, beta.hat.inla.MF2)
## cor(g, u.hat.inla.MF2)
c(sigma.epsilon, sigma.epsilon.hat.inla.MF2)
c(sigma.g, sigma.u.hat.inla.MF2)
c(h2, h2.hat.inla.MF2)
```


## With breedR

```{r breedR_fit_prep_dat}
dat.breedR <- cbind(dat,
                    gxy=factor(dat$geno:dat$year))
str(dat.breedR)
```

### Model fitting

```{r breedR_fit}
system.time(
    fit.breedR.g <- remlf90(
        fixed=pheno ~ 1,
        generic=list(geno=list(incidence=Z, covariance=A, var.ini=2)),
        data=dat,
        var.ini=list(resid=10),
        progsf90.options=paste("se_covar_function h2",
                               "G_2_2_1_1/(G_2_2_1_1+R_1_1)"),
        method="ai"))
system.time(
    fit.breedR.y.g <- remlf90(
        fixed=pheno ~ 1 + year,
        generic=list(geno=list(incidence=Z, covariance=A, var.ini=2)),
        data=dat,
        var.ini=list(resid=10),
        progsf90.options=paste("se_covar_function h2",
                               "G_2_2_1_1/(G_2_2_1_1+R_1_1)"),
        method="ai"))
system.time(
    fit.breedR.yg <- remlf90(
        fixed=pheno ~ 1 + year,
        generic=list(geno=list(incidence=Z, covariance=A, var.ini=2)),
        data=dat,
        var.ini=list(resid=10),
        progsf90.options=paste("se_covar_function h2",
                               "G_2_2_1_1/(G_2_2_1_1+R_1_1)"),
        method="ai"))
```

### Model comparison and selection

```{r breedR_select}
extractAIC(fit.breedR.g)
extractAIC(fit.breedR.y.g)
extractAIC(fit.breedR.yg)
fit.breedR <- fit.breedR.y.g
```

### Diagnostics

```{r breedR_orga}
beta.hat.breedR <- setNames(as.vector(fixef(fit.breedR)$year),
                            names(fixef(fit.breedR)$year))
u.hat.breedR <- setNames(as.vector(ranef(fit.breedR)$geno),
                         names(ranef(fit.breedR)$geno))
var.u.hat.breedR <- attr(ranef(fit.breedR)$geno, "se")^2
sigma.epsilon.hat.breedR <- sqrt(fit.breedR$var["Residual",
                                                "Estimated variances"])
sigma.u.hat.breedR <- sqrt(fit.breedR$var["geno",
                                          "Estimated variances"])
h2.hat.breedR <- fit.breedR$funvars["sample mean", "h2"]
h2.hat.low.breedR <- fit.breedR$funvars["sample mean", "h2"] -
  abs(low.bound.norm.95) * fit.breedR$funvars["sample sd", "h2"]
h2.hat.high.breedR <- fit.breedR$funvars["sample mean", "h2"] +
  abs(low.bound.norm.95) * fit.breedR$funvars["sample sd", "h2"]
fit.breedR.all <- cbind(dat,
                        cond.res=residuals(fit.breedR),
                        scl.cond.res=residuals(fit.breedR) /
                          sigma.epsilon.hat.breedR,
                        fitted=fitted(fit.breedR))
str(fit.breedR.all, give.attr=FALSE)
```

```{r breedR_check}
x.lim <- max(abs(fit.breedR.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.breedR.all$scl.cond.res, y=fit.breedR.all$fitted, las=1,
     xlim=c(-x.lim, x.lim))
abline(v=0, lty=2)
abline(v=c(low.bound.norm.95, abs(low.bound.norm.95)), lty=3)
shapiro.test(fit.breedR.all$scl.cond.res)
qqnorm(y=fit.breedR.all$scl.cond.res, main="Normal Q-Q plot of scl.cond.res")
qqline(y=fit.breedR.all$scl.cond.res, col="red")
x.lim <- max(abs(u.hat.breedR))
par(mar=c(5,6,4,1))
plot(x=u.hat.breedR, y=1:length(u.hat.breedR),
     xlim=c(-x.lim, x.lim),
     yaxt="n", ylab="")
axis(side=2, at=1:length(u.hat.breedR), labels=names(u.hat.breedR),
     las=1)
abline(v=0, lty=2)
```

### Inference

```{r breedR_res}
summary(fit.breedR)
cbind(beta, beta.hat.breedR)
cor(g, u.hat.breedR)
c(sigma.epsilon, sigma.epsilon.hat.breedR)
c(sigma.g, sigma.u.hat.breedR)
fit.breedR$var[, "S.E."]
c(h2, h2.hat.breedR, h2.hat.low.breedR, h2.hat.high.breedR)
```


## With MCMCglmm

TODO, but potentially (much) slower



# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
