---
title: "GBLUP with population structure"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
```


# Overview

This document aims at using GBLUP on a sample showing population structure, following [Janss et al (2012)](http://www.genetics.org/content/192/2/693.long).

This document requires external packages to be available:
```{r load_pkg}
suppressPackageStartupMessages(library(scrm)) # on the CRAN
suppressPackageStartupMessages(library(coda)) # on the CRAN
suppressPackageStartupMessages(library(rutilstimflutre)) # on GitHub
stopifnot(compareVersion("0.150.0",
                         as.character(packageVersion("rutilstimflutre")))
          != 1)
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Simulate data

## Genotypes

Set the seed:
```{r simul_genos_set_seed}
set.seed(1859)
```

Simulate haplotypes and genotypes in three populations (the first two closer to each other than with the third):
```{r simul_haplos_genos}
nb.pops <- 3
nb.genos <- 3 * 10^2
nb.chrs <- 3
chr.len.phy <- 10^4     # chromosome physical length in base pairs
mu <- 10^(-7)           # neutral mutation rate in events / base / generation
u <- mu * chr.len.phy   #  in events / chrom / gen
chr.len.gen <- 10^(-1)  # chromosome genetic length in Morgans
c.r <- chr.len.gen / chr.len.phy # recomb rate in events / base / gen
r <- c.r * chr.len.phy  #  in events / chrom / gen
m <- 1 * 10^(-4)        # fraction of a pop replaced from other pops each gen
Ne <- 10^4              # effective population size
theta <- 4 * Ne * u     # scaled neutral mutation rate in events / chrom
rho <- 4 * Ne * r       # scaled recomb rate in events / chrom
M <- 4 * Ne * m         # scaled migration rate in events
system.time(
    genomes <- simulCoalescent(nb.inds=nb.genos,
                               nb.reps=nb.chrs,
                               pop.mut.rate=theta,
                               pop.recomb.rate=rho,
                               chrom.len=chr.len.phy,
                               nb.pops=nb.pops,
                               other=paste0("-m 1 2 ", M,
                                            " -m 1 3 ", 1.7 * M,
                                            " -m 2 3 ", 1.7 * M),
                               get.alleles=TRUE))
nb.snps <- ncol(genomes$genos)
afs.pop <- estimSnpAf(X=genomes$genos)
mafs.pop <- estimSnpMaf(afs=afs.pop)
mrk2chr <- setNames(genomes$snp.coords$chr, rownames(genomes$snp.coords))
```


## Phenotypes

Set the seed:
```{r simul_phenos_set_seed}
set.seed(1859)
```

```{r simul_phenos}
mu <- 50
h2 <- 0.6
phenos <- simulBvsr(Q=1, mu=mu, X=genomes$genos, m1=FALSE, ctr=TRUE,
                    ## std=TRUE,
                    std=FALSE,
                    pi=1, pve=h2)
phenos$sigma2
phenos$sigma.a2
(sigma.g.a2 <- phenos$sigma.a2 * 2 * sum(afs.pop * (1 - afs.pop)))
sigma.g.a2 / (sigma.g.a2 + phenos$sigma2) # close to the specified h2
```

Note that one should set `std=FALSE` in order to get the specified `pve`.


# Explore the data

## Genotypes

Check the number of genotypes per population:
```{r explo_geno_inds_per_pop}
table(inds.per.pop <- kmeans(genomes$genos, nb.pops)$cluster)
```

Look at allele frequencies:
```{r explo_geno_allel_freq}
plotHistAllelFreq(afs=afs.pop, main="Allele frequencies")
plotHistMinAllelFreq(mafs=mafs.pop, main="Minor allele frequencies")
```

PCA can be used to look at population structure.

As in [Engelhart and Stephens (2010)](http://dx.doi.org/10.1371/journal.pgen.1001117), the equivalent of admixture proportions can be plotted (note also the effect of centering, as in their figure 3):
```{r explo_geno_pca_centering}
pca.X <- pca(X=genomes$genos, ct=FALSE, sc=FALSE, ES10=TRUE)
par(mfrow=c(2,2))
plot(1:nb.genos, pca.X$Lambda[,1], ylim=c(-0.5, 0.5), main="Lambda 1", ylab="")
plot(1:nb.genos, pca.X$Lambda[,2], main="Lambda 2", ylab=""); abline(h=0, lty=2)
plot(1:nb.genos, pca.X$Lambda[,3], main="Lambda 3", ylab=""); abline(h=0, lty=2)
plot(1:nb.genos, pca.X$Lambda[,4], main="Lambda 4", ylab=""); abline(h=0, lty=2)
```

The relative magnitude of the eigenvalues can give some indication about the number of populations represented in the sample:
```{r explo_geno_pca_eigenvalues}
par(mfrow=c(1,2))
barplot(pca.X$prop.vars[1:10], main="pca(X) w/o centering")
barplot(pca.X$prop.vars[1:10], main="zoom-in", ylim=c(0,0.01))
```

[Shriner (2011)](http://dx.doi.org/10.1038/hdy.2011.26) proposed a method based on average squared partial correlations to choose the number of PCs to keep, but it doesn't seem to be very efficient here:
```{r explo_geno_pca_Shriner}
getNbPCsMinimAvgSqPartCor(t(genomes$genos))
```

The `MASS::parcoord` function can be helpful in visualizing the coordinates of each individual in the PC system:
```{r explo_geno_parcoord}
MASS::parcoord(pca.X$rot.dat[,1:6], col=inds.per.pop, main="pca(X)")
```

Different estimators exist to estimate additive relationships between individuals:
```{r explo_geno_add_rel}
A.cs <- estimGenRel(X=genomes$genos, afs=afs.pop, method="center-std")
A.vr <- estimGenRel(X=genomes$genos, afs=afs.pop, method="vanraden1")
pca.A.cs <- pca(S=A.cs)
pca.A.vr <- pca(S=A.vr)
par(mfrow=c(1,2))
barplot(pca.A.cs$prop.vars[1:10], main="pca(A.cs)")
barplot(pca.A.vr$prop.vars[1:10], main="pca(A.vr)")
plotPca(pca.A.vr$rot.dat, prop.vars=pca.A.vr$prop.vars,
        idx.x=1, idx.y=2, cols=inds.per.pop, main="A.vr")
plotPca(pca.A.vr$rot.dat, prop.vars=pca.A.vr$prop.vars,
        idx.x=2, idx.y=3, cols=inds.per.pop, main="A.vr")
MASS::parcoord(pca.A.vr$rot.dat[,1:6], col=inds.per.pop, main="pca(A.vr)")
```

But the one from [VanRaden (2008)](http://dx.doi.org/10.3168/jds.2007-0980) may be preferred as it is on the same scale as the one calculated from the pedigree, as shown in [Toro et al (2011)](http://www.gsejournal.org/content/43/1/27):
```{r check_simul_genos}
imageWithScale(A.vr, main="Additive genetic relationships")
summary(diag(A.vr))
summary(A.vr[upper.tri(A.vr)])
par(mfrow=c(1,2))
hist(diag(A.vr), col="grey", border="white")
hist(A.vr[upper.tri(A.vr)], col="grey", border="white")
```

Whatsoever, Janss et al (2012) center and standardize the SNP genotype matrix, and then take its cross product divided by the number of genotypes.
```{r explo_geno_pca_fig4}
W <- scale(x=genomes$genos, center=TRUE, scale=TRUE)
G <- tcrossprod(W) / ncol(W)
pca.G <- pca(S=G)
which(sapply(pca.G$eigen.values, function(x){
  all.equal(0, x) == TRUE
})) # one eigenvalue is 0 (the latest) due to the centering of W
barplot(pca.G$prop.vars[1:10], main="pca(G)")
plotPca(pca.G$rot.dat, pca.G$prop.vars, idx.x=1, idx.y=2, main="pca(G)",
        cols=inds.per.pop)
```

Plot figure 4 from Janss et al on the simulated data:
```{r explo_geno_pca_fig4}
plot(x=0:nb.genos, y=c(0, cumsum(pca.G$eigen.values)) / nb.genos,
     ylim=c(0,1), type="l", las=1,
     xlab="number of eigenvalues", ylab="cumulative sum", main="pca(G)")
abline(h=1, a=0, b=1/nb.genos, lty=2)
```

## Phenotypes

```{r explo_pheno}
phenos$dat$pop <- as.factor(paste0("pop", inds.per.pop))
hist(phenos$Y[,1], col="grey", border="white", las=1,
     main="Simulated phenotypes", xlab="")
boxplot(response1 ~ pop, data=phenos$dat, notch=TRUE, las=1,
        main="Simulated phenotypes")
abline(h=median(phenos$dat$response1), lty=2)
```



# Perform inference

### Classical LMM

```{r}
fit.g <- lmer(response1 ~ 1 + (1|geno), data=phenos$dat,
              control=lmerControl(check.nobs.vs.nlev="ignore",
                                  check.nobs.vs.nRE="ignore"))
fit.p.g <- lmer(response1 ~ 1 + pop + (1|geno), data=phenos$dat,
                control=lmerControl(check.nobs.vs.nlev="ignore",
                                    check.nobs.vs.nRE="ignore"))
extractAIC(fit.g)
extractAIC(fit.p.g)
```

### As in Janss et al

\[
\boldsymbol{y} = \boldsymbol{1} \mu + W \boldsymbol{b} + \boldsymbol{e} \text{ with } \boldsymbol{b} \sim \mathcal{N}(\boldsymbol{0}, \sigma_b^2 \text{Id}) \text{ and } \boldsymbol{e} \sim \mathcal{N}(\boldsymbol{0}, \sigma_e^2 \text{Id})
\]

* $n$: number of individuals

* $m$: number of markers

* $W$: $n \times m$ matrix of marker genotypes, centered and scaled

* $\boldsymbol{b}$: $m \times 1$ matrix of marker effects

* $\boldsymbol{g}$: $n \times 1$ matrix of genomic values such that $\boldsymbol{g} = W \boldsymbol{b}$ such that $\mathbb{V}(\boldsymbol{g}|W) = W W' \sigma_b^2 = \frac{1}{m} W W' \sigma_g^2$

Factorization: $W W' = U D U'$

\[
\boldsymbol{y} = \boldsymbol{1} \mu + U \boldsymbol{\alpha} + \boldsymbol{e} \text{ with } \boldsymbol{\alpha} \sim \mathcal{N}(\boldsymbol{0}, \sigma_b^2 D)
\]

* $\boldsymbol{\alpha}$: $n \times 1$ matrix of principal components of $WW'$ (indeed: $\boldsymbol{g} = U \boldsymbol{\alpha} \Leftrightarrow \boldsymbol{\alpha} = U' \boldsymbol{g}$)

### Simplified implementation

```{r implem_simplif, eval=FALSE}
##' 
##'
##' Run a Gibbs sampler
##' @param y vector of responses of length n
##' @param K list of lists (one per prior)
##' @param XF n x pF incidence matrix for fixed effects
##' @param df0 degrees of freedom for the prior variances
##' @param S0 scale for the prior variances
##' @param weights optional vector
##' @param nIter number of iterations
##' @param burnIn burn-in
##' @param thin thinning
##' @param saveAt if not NULL, results will be saved in this file
##' @param verbose verbosity level (0/1/2)
##' @return list
##' @author Luc Janss [aut], Gustavo de los Campos [aut], Timothee Flutre [ctb]
##' @export
gibbsJanss2012simple <- function(y, K, XF= NULL, df0=0, S0=0, weights=NULL,
                                 nIter=110, burnIn=10, thin=10, saveAt=NULL,
                                 verbose=0){
  stopifnot(is.numeric(y),
            is.list(K))
  if(! is.null(saveAt))
    if(file.exists(saveAt))
      file.remove(saveAt)

  if(verbose > 0)
    write("prepare inputs...", stdout())
  iter <- 0
  n <- length(y)
  whichNa <- which(is.na(y))
  nNa <- length(whichNa)
  hasWeights <- ! is.null(weights)
  if(! hasWeights)
    weights <- rep(1, n)
  sumW2 <- sum(weights^2)
  mu <- weighted.mean(x=y, w=weights, na.rm=TRUE)
  yStar <- y * weights
  yStar[whichNa] <- mu * weights[whichNa]
  e <- (yStar - weights * mu)
  varE <- as.numeric(var(e, na.rm=TRUE) / 2)
  sdE <- sqrt(varE)
  nK <- length(K)
  postMu <- 0
  postVarE <- 0
  postYHat <- rep(0, n)
  postYHat2 <- rep(0, n)
  postLogLik <- 0
  hasXF <- ! is.null(XF)
  if(hasXF){
    for(i in 1:n)
      XF[i, ] <- weights[i] * XF[i, ]
    SVD.XF <- svd(XF)
    SVD.XF$Vt <- t(SVD.XF$v)
    SVD.XF <- SVD.XF[-3]
    pF0 <- length(SVD.XF$d)
    pF <- ncol(XF)
    bF0 <- rep(0, pF0)
    bF <- rep(0, pF)
    namesBF <- colnames(XF)
    post_bF <- bF
    post_bF2 <- bF
    rm(XF)
  }
  for(k in 1:nK){
    if(hasWeights & is.null(K[[k]]$K))
      stop("if weights are provided, K is needed")
    if(hasWeights){
      T <- diag(weights)
      K[[k]]$K <- T %*% K[[k]]$K %*% T
    }
    if(is.null(K[[k]]$V)){
      K[[k]]$K <- as.matrix(K[[k]]$K)
      tmp <- eigen(K[[k]]$K)
      K[[k]]$V <- tmp$vectors
      K[[k]]$d <- tmp$values
    }
    if(is.null(K[[k]]$tolD))
      K[[k]]$tolD <- 1e-12

    tmp <- K[[k]]$d > K[[k]]$tolD
    K[[k]]$levelsU <- sum(tmp)
    K[[k]]$d <- K[[k]]$d[tmp]
    K[[k]]$V <- K[[k]]$V[, tmp]
    if(is.null(K[[k]]$df0))
      K[[k]]$df0 <- 1
    if(is.null(K[[k]]$S0))
      K[[k]]$S0 <- 1e-15
    K[[k]]$varU <- varE / nK / 2
    K[[k]]$u <- rep(0, n)
    K[[k]]$uStar <- rep(0, K[[k]]$levelsU)
    K[[k]]$postVarU <- 0
    K[[k]]$postUStar <- rep(0, K[[k]]$levelsU)
    K[[k]]$postU <- rep(0, n)
    K[[k]]$postCumMSa<-rep(0, K[[k]]$levelsU)
    K[[k]]$postH1<-rep(0, K[[k]]$levelsU)
  }

  if(verbose > 0)
    write("perform inference...", stdout())
  if(! is.null(saveAt)){
    tmp <- c("logLik", "mu", "varE", paste0("varU", 1:nK))
    if(hasXF)
      tmp <- c(tmp, paste0("bF", 1:pF))
    write(tmp, ncol=length(tmp), file=saveAt, append=FALSE,  sep=" ")
  }
  if(verbose > 0){
    tmpOut <- paste0("iter: ", 0,
                     "; time: ", round(0, 4),
                     "; varE: ", round(varE, 3),
                     "; mu: ", round(mu, 3), "\n")
    cat(tmpOut)
  }
  time <- proc.time()[3]
  for(i in 1:nIter){
    if(hasXF){
      sol <- (crossprod(SVD.XF$u, e) + bF0)
      tmp <- sol + rnorm(n=pF0, sd=sqrt(varE))
      bF <- crossprod(SVD.XF$Vt, tmp / SVD.XF$d)
      e <- e + SVD.XF$u %*% (bF0 - tmp)
      bF0 <- tmp
    }
    for(k in 1:nK){
      e <- e + K[[k]]$u
      rhs <- crossprod(K[[k]]$V, e) / varE
      varU <- K[[k]]$varU * K[[k]]$d
      C <- as.numeric(1 / varU + 1 / varE)
      SD <- 1 / sqrt(C)
      sol <- rhs / C
      tmp <- rnorm(n=K[[k]]$levelsU, sd=SD, mean=sol)
      K[[k]]$uStar <- tmp
      K[[k]]$u <- as.vector(K[[k]]$V %*% tmp)
      e <- e - K[[k]]$u
      tmp <- K[[k]]$uStar / sqrt(K[[k]]$d)
      S <- as.numeric(crossprod(tmp)) + K[[k]]$S0
      df <- K[[k]]$levelsU + K[[k]]$df0
      K[[k]]$varU <- S / rchisq(n=1, df=df)
    }
    e <- e + weights * mu
    rhs <- sum(weights * e) / varE
    C <- sumW2 / varE
    sol <- rhs / C
    mu <- rnorm(n=1, sd=sqrt(1 / C), mean=sol)
    e <- e - weights * mu
    df <- n + df0
    SS <- as.numeric(crossprod(e)) + S0
    varE <- SS / rchisq(n=1, df=df)
    yHat <- yStar - e
    if (nNa > 0) {
      yStar[whichNa] <- yHat[whichNa] + rnorm(n=nNa,  sd=sdE)
      e[whichNa] <- yStar[whichNa] - yHat[whichNa]
    }
    tmpE <- e / weights
    tmpSD <- sqrt(varE) / weights
    if(nNa > 0){
      tmpE <- tmpE[-whichNa]
      tmpSD <- tmpSD[-whichNa]
    }
    logLik <- sum(dnorm(tmpE, sd=tmpSD, log=TRUE))
    if ((i > burnIn) & (i %% thin == 0)) {
      iter <- iter + 1
      constant <- (iter - 1) / (iter)
      postMu <- postMu * constant + mu / iter
      postVarE <- postVarE * constant + varE / iter
      postYHat <- postYHat * constant + yHat / iter
      for(k in 1:nK){
        K[[k]]$postVarU <- K[[k]]$postVarU * constant +  K[[k]]$varU / iter
        K[[k]]$postUStar <- K[[k]]$postUStar * constant + K[[k]]$uStar / iter
        K[[k]]$postU <- K[[k]]$postU * constant + K[[k]]$u / iter
        tmp <- cumsum(K[[k]]$uStar^2) / K[[k]]$levelsU
        K[[k]]$postCumMSa <- K[[k]]$postCumMSa * constant + tmp / iter
        tmp <- K[[k]]$uStar^2 > mean(K[[k]]$uStar^2)
        K[[k]]$postH1 <- K[[k]]$postH1 * constant + tmp / iter 
      }
      postLogLik <- postLogLik * constant + logLik / iter
      if(hasXF){
        post_bF <- post_bF * constant + bF / iter
        post_bF2 <- post_bF2 * constant + (bF^2) / iter
      }
    }
    tmp <- i %% thin == 0
    if(! is.null(saveAt) & i > burnIn & tmp){
      tmp <- c(logLik, mu, varE)
      for(k in 1:nK)
        tmp <- c(tmp, K[[k]]$varU)
      if(hasXF)
        tmp <- bF
      write(tmp, ncol=length(tmp), file=saveAt, append=TRUE,  sep=" ")
    }
    tmp <- proc.time()[3]
    if(verbose > 0){
      if(verbose == 1 & i %% thin != 0)
        next
      tmpOut <- paste0("iter: ", i,
                       "; time: ", round(tmp - time, 4),
                       "; varE: ", round(varE, 3), "\n")
      cat(tmpOut)
    }
    time <- tmp
  }

  if(verbose > 0)
    write("prepare the output...", stdout())
  out <- list(mu=postMu, fit=list(), varE=postVarE,
              yHat=as.numeric(postYHat / weights),
              weights=weights, K=list(), whichNa=whichNa, df0=df0, 
              S0=S0, nIter=nIter, burnIn=burnIn, saveAt=saveAt)
  out$fit$postMeanLogLik <- postLogLik
  tmpE <- (yStar - postYHat) / weights
  tmpSD <- sqrt(postVarE) / weights
  if(nNa > 0){
    tmpE <- tmpE[-whichNa]
    tmpSD <- tmpSD[-whichNa]
  }
  out$fit$logLikAtPostMean <- sum(dnorm(tmpE, sd=tmpSD, log=TRUE))
  out$fit$pD <- -2 * (out$fit$postMeanLogLik - out$fit$logLikAtPostMean)
  out$fit$DIC <- out$fit$pD - 2 * out$fit$postMeanLogLik
  if(hasXF){
    out$bF <- as.vector(post_bF)
    out$SD.bF <- as.vector(sqrt(post_bF2 - post_bF^2))
    names(out$bF) <- namesBF
    names(out$SD.bF) <- namesBF
  }
  for(k in 1:nK){
    out$K[[k]] <- list()
    out$K[[k]]$u <- K[[k]]$postU
    out$K[[k]]$uStar <- K[[k]]$postUStar
    out$K[[k]]$varU <- K[[k]]$postVarU
    out$K[[k]]$cumMSa <- K[[k]]$postCumMSa
    out$K[[k]]$probH1 <- K[[k]]$postH1
    out$K[[k]]$df0 <- K[[k]]$df0
    out$K[[k]]$S0 <- K[[k]]$S0
    out$K[[k]]$tolD <- K[[k]]$tolD
  }
  return(out)
}
```

### Run Gibbs sampler

```{r run_gibbs_simulated_data, eval=TRUE}
nb.iters <- 2 * 10^4
burnin <- 2 * 10^3
thin <- 10
seed <- 1859
fit <- gibbsJanss2012(y=phenos$Y[,1],
                      XF=NULL,
                      K=list(list(V=pca.G$rot.dat,
                                  d=pca.G$eigen.values,
                                  df0=5, S0=3/2)),
                      nIter=nb.iters, burnIn=burnin, thin=thin,
                      saveAt="gibbsJanss2012_samples.txt", verbose=0)
fit$mu
fit$varE
fit$K[[1]]$varU
fit$fit$DIC
posteriors <- mcmc(data=read.table(fit$saveAt, header=TRUE))
posteriors <- mcmc(cbind(posteriors,
                         h2=posteriors[,"varU1"] /
                           (posteriors[,"varU1"] + posteriors[,"varE"])))
summary(posteriors)
snpeffs <- t(W) %*% pca.G$rot.dat %*%
  diag(1 / pca.G$eigen.values) %*% fit$K[[1]]$u
```

Now let's shuffle the rows of W to reproduce the equivalent of figure 2 of Janss et al:
```{r run_gibbs_simulated_data_shuffle, eval=TRUE}
shuf.idx <- sample.int(n=nrow(W))
G.shuf <- tcrossprod(W[shuf.idx,]) / ncol(W)
dimnames(G.shuf) <- dimnames(G)
pca.G.shuf <- pca(S=G.shuf)
fit.shuf <- gibbsJanss2012(y=phenos$Y[,1],
                           XF=NULL,
                           K=list(list(V=pca.G.shuf$rot.dat,
                                       d=pca.G.shuf$eigen.values,
                                       df0=5, S0=3/2)),
                           nIter=nb.iters, burnIn=burnin, thin=thin,
                           saveAt="gibbsJanss2012_samples.txt", verbose=0)
fit.shuf$mu
fit.shuf$varE
fit.shuf$K[[1]]$varU
fit.shuf$fit$DIC
posteriors.shuf <- mcmc(data=read.table(fit.shuf$saveAt, header=TRUE))
snpeffs.shuf <- t(W) %*% pca.G.shuf$rot.dat %*%
  diag(1 / pca.G.shuf$eigen.values) %*% fit.shuf$K[[1]]$u
## reproduce figure 2
plot(x=1:ncol(W), y=abs(snpeffs), pch=20, col="black")
points(x=1:ncol(W), y=abs(snpeffs.shuf), pch=20, col="grey")
```

Now let's reproduce the equivalent of figures 5 and 6 of Janss et al:
```{r run_gibbs_simulated_data, fig.height=15, eval=TRUE}
nb.PCs <- 0
nb.iters <- 2 * 10^4
burnin <- 2 * 10^3
thin <- 10
seed <- 1859
if(nb.PCs == 0){
  fm <- RKHS(y=phenos$Y[,1],
             XF=NULL,
             K=list(list(V=pca.G$rot.dat, d=pca.G$eigen.values,
                         df0=5, S0=3/2)),
             nIter=nb.iters, burnIn=burnin, thin=thin)
  fit <- gibbsJanss2012(y=phenos$Y[,1],
                        XF=NULL,
                        K=list(list(V=pca.G$rot.dat, d=pca.G$eigen.values,
                                    df0=5, S0=3/2)),
                        nIter=nb.iters, burnIn=burnin, thin=thin,
                        saveAt="gibbsJanss2012_samples.txt", verbose=0)
} else
  fit <- gibbsJanss2012(y=phenos$Y[,1],
                        XF=pca.G$rot.dat[, 1:nb.PCs, drop=FALSE],
                        K=list(list(V=pca.G$rot.dat, d=pca.G$eigen.values,
                                    df0=5, S0=3/2)),
                        nIter=nb.iters, burnIn=burnin, thin=thin,
                        saveAt=getwd(), verbose=0)

fm$mu # 50
fm$varE # 150
fm$K[[1]]$varU # 139
fm$fit$DIC # 2463

fit$mu # 50
fit$varE # 155
fit$K[[1]]$varU # 133
fit$fit$DIC # 2469

post.samples <- mcmc(data=read.table(fit$saveAt, header=TRUE))
dim(post.samples)
post.samples <- mcmc(cbind(post.samples,
                           h2=post.samples[,"varU1"] /
                             (post.samples[,"varU1"] + post.samples[,"varE"])))
head(post.samples)
plot(post.samples, ask=FALSE)
summary(post.samples)
HPDinterval(post.samples)
```

Clean:
```{r clean}
for(f in Sys.glob("gibbsJanss2012_*.txt"))
  file.remove(f)
```



# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
