---
title: "Imputation of SNP genotypes in a population with fastPHASE"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
```


# Overview

This document requires external packages to be available:
```{r load_pkg}
suppressPackageStartupMessages(library(scrm))
suppressPackageStartupMessages(library(rutilstimflutre))
stopifnot(file.exists(Sys.which("fastphase")))
suppressPackageStartupMessages(library(parallel))
nb.cores <- detectCores() - 1
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Simulate genetic data at HWE

Set the seed:
```{r set_seed}
set.seed(1859)
```

Simulate haplotypes and genotypes in a single population:
```{r simul_haplos_genos}
nb.genos <- 1*10^3
nb.chrs <- 6
Ne <- 10^4
chrom.len <- 1*10^6
mu <- 10^(-8)
c.rec <- 10^(-8)
genomes <- simulCoalescent(nb.inds=nb.genos,
                           nb.reps=nb.chrs,
                           pop.mut.rate=4 * Ne * mu * chrom.len,
                           pop.recomb.rate=4 * Ne * c.rec * chrom.len,
                           chrom.len=chrom.len,
                           get.alleles=TRUE)
afs.hwe <- estimSnpAf(X=genomes$genos)
plotHistAllelFreq(afs=afs.hwe,
                  main="Allele frequencies at HWE")
mafs.hwe <- estimSnpMaf(afs=afs.hwe)
plotHistMinAllelFreq(mafs=mafs.hwe,
                     main="Minor allele frequencies at HWE")
A.vr.hwe <- estimGenRel(X=genomes$genos, afs=afs.hwe, method="vanraden1")
imageWithScale(A.vr.hwe,
               main="Additive genetic relationships at HWE")
summary(diag(A.vr.hwe))
hist(diag(A.vr.hwe), col="grey", border="white")
summary(A.vr.hwe[upper.tri(A.vr.hwe)])
hist(A.vr.hwe[upper.tri(A.vr.hwe)], col="grey", border="white")
```


# Mimick RAD-seq

Discard SNP genotypes according to a "RAD-seq" design, i.e. all genotypes and SNPs have NA, but with a skewed distribution (30% of the sites with 0-0.05% NA, 17% of the sites with 0.05-0.1% NA, 8% with 0.1-0.15% NA, and the remaining sites with a uniformly-increasing amount of NA):
```{r radseq}
genos.tokeep <- sample(rownames(genomes$genos), size=300)
dim(X <- genomes$genos[genos.tokeep,])
X.na <- X
bin.width <- 0.05
max.prop <- 0.9
bin.mids <- seq(from=bin.width/2, to=max.prop, by=bin.width)
bin.heights <- c(0.3, 0.17, 0.08,
                 rep((1-(0.3+0.17+0.08))/(length(bin.mids) - 3),
                     length(bin.mids) - 3))
stopifnot(sum(bin.heights) == 1)
assigned.bins <- cut(x=1:ncol(X),
                     breaks=floor(ncol(X) * cumsum(c(0, bin.heights))),
                     labels=FALSE)
assigned.bins <- sample(assigned.bins) # shuffle the bins among SNPs
table(assigned.bins)
bin.probs <- seq(from=0, to=max.prop, length.out=length(bin.mids))
## for each SNP, sample genotype indices to set as missing
idx1 <- sapply(1:ncol(X), function(j){
  sample.int(n=nrow(X), size=round(bin.probs[assigned.bins[j]] * nrow(X)))
})
idx2 <- do.call(c, lapply(1:length(idx1), function(j){
  ((j-1) * nrow(X)) + idx1[[j]]
}))
X.na[idx2] <- NA
sum(is.na(X.na))
sum(is.na(X.na)) / length(X.na)
```

Plot grid of missing SNP genotypes:
```{r}
plotGridMissGenos(X=X.na)
```

Plot histogram of NA proportions (same as fig S4 of [Chan et al, 2016](http://dx.plos.org/10.1371/journal.pone.0160733)):
```{r}
miss.snps <- calcFreqMissSnpGenosPerSnp(X=X.na)
hist(miss.snps, breaks=seq(0,1,0.05), col="grey", border="white",
     main="Distribution of the proportion of missing data per biallelic SNP",
     xlab="proportion of missing data at a given SNP",
     ylab="number of SNPs")
```


# Perform imputation

```{r imputation}
system.time(
    X.imp <- runFastphase(X.na)
```


# Assess accuracy

```{r assess_accuracy}
length(X)
sum(X.na == X, na.rm=TRUE)
sum(is.na(X.na))
sum(X.imp == X)
sum(X.imp != X)
sum(X.imp != X) / sum(is.na(X.na))
```

Check additive genetic relationships:
```{r}
A.vr.imp <- estimGenRel(X=X.imp, afs=afs.hwe)
summary(diag(A.vr.imp))
summary(A.vr.imp[upper.tri(A.vr.imp)])
```


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
