---
title: "Genotype imputation"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
```


# Overview

This document requires external packages to be available:
```{r load_pkg}
library(rutilstimflutre)
library(scrm)
stopifnot(file.exists(Sys.which("FImpute")))
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Simulate genetic data at HWE

Set the seed:
```{r, set_seed}
set.seed(1859)
```

Simulate haplotypes and genotypes in a single population:
```{r simul_haplos_genos}
nb.genos <- 1*10^3
nb.chrs <- 2
Ne <- 10^4
chrom.len <- 1*10^6
mu <- 10^(-8)
c.rec <- 10^(-8)
genomes <- simulCoalescent(nb.inds=nb.genos,
                           nb.reps=nb.chrs,
                           pop.mut.rate=4 * Ne * mu * chrom.len,
                           pop.recomb.rate=4 * Ne * c.rec * chrom.len,
                           chrom.len=chrom.len,
                           get.alleles=TRUE)
afs.hwe <- estimSnpAf(X=genomes$genos)
plotHistAllelFreq(afs=afs.hwe,
                  main="Allele frequencies at HWE")
mafs.hwe <- estimSnpMaf(afs=afs.hwe)
plotHistMinAllelFreq(mafs=mafs.hwe,
                     main="Minor allele frequencies at HWE")
A.vr.hwe <- estimGenRel(X=genomes$genos, afs=afs.hwe, method="vanraden1")
imageWithScale(A.vr.hwe,
               main="Additive genetic relationships at HWE")
summary(diag(A.vr.hwe))
hist(diag(A.vr.hwe), col="grey", border="white")
summary(A.vr.hwe[upper.tri(A.vr.hwe)])
hist(A.vr.hwe[upper.tri(A.vr.hwe)], col="grey", border="white")
```


# Perform controlled crosses

Choose 4 individuals at random as grand-parents:
```{r, choose_gdparents}
(idx.gdparents <- sample.int(n=nb.genos, size=4))
genos.gdparents <- genomes$genos[idx.gdparents,]
names.gdparents <- rownames(genos.gdparents)
haplos.gdparents <- getHaplosInds(haplos=genomes$haplos,
                                  ind.names=names.gdparents)
```

Make 2 pairs and cross each pair once, to make parents:
```{r, crosse_to_make_parents}
name.par1 <- paste0(idx.gdparents[1], "x", idx.gdparents[2], "-1")
name.par2 <- paste0(idx.gdparents[3], "x", idx.gdparents[4], "-1")
(crosses.par <- data.frame(parent1=c(names.gdparents[1], names.gdparents[3]),
                           parent2=c(names.gdparents[2], names.gdparents[4]),
                           child=c(name.par1, name.par2),
                           stringsAsFactors=FALSE))
loc.crovers.par <- drawLocCrossovers(crosses=crosses.par,
                                     nb.snps=sapply(haplos.gdparents, ncol))
haplos.par <- makeCrosses(haplos=haplos.gdparents, crosses=crosses.par,
                          loc.crossovers=loc.crovers.par)
genos.par <- segSites2allDoses(seg.sites=haplos.par,
                               ind.ids=getIndNamesFromHaplos(haplos.par),
                               snp.ids=rownames(genomes$snp.coords))
```

Cross them several times to make offsprings:
```{r, crosse_to_make_offsprings}
nb.offs <- 200
names.offs <- paste0("off-",
                     sprintf(fmt=paste0("%0", floor(log10(nb.offs))+1, "i"),
                             1:nb.offs))
head(crosses.off <- data.frame(parent1=rep(name.par1, nb.offs),
                               parent2=rep(name.par2, nb.offs),
                               child=names.offs,
                               stringsAsFactors=FALSE))
loc.crovers.off <- drawLocCrossovers(crosses=crosses.off,
                                     nb.snps=sapply(haplos.par, ncol))
haplos.offs <- makeCrosses(haplos=haplos.par, crosses=crosses.off,
                           loc.crossovers=loc.crovers.off)
genos.offs <- segSites2allDoses(seg.sites=haplos.offs,
                                ind.ids=getIndNamesFromHaplos(haplos.offs),
                                snp.ids=rownames(genomes$snp.coords))
```

Plot pedigree:
```{r, plot_ped}
ped <- data.frame(ind=c(names.gdparents,
                        crosses.par$child,
                        crosses.off$child),
                  mother=c(rep(NA, length(names.gdparents)),
                           crosses.par$parent1,
                           crosses.off$parent1),
                  father=c(rep(NA, length(names.gdparents)),
                           crosses.par$parent2,
                           crosses.off$parent2),
                  gen=c(rep(-1, length(names.gdparents)),
                        rep(0, nrow(crosses.par)),
                        rep(1, nrow(crosses.off))),
                  stringsAsFactors=FALSE)
ped.tmp <- rbind(ped[1:7,],
                 c(ind="off-..", ped[7, -1]),
                 c(ind="off-...", ped[7, -1]),
                 c(ind="off-....", ped[7, -1]),
                 ped[nrow(ped),])
plotPedigree(inds=ped.tmp$ind, mothers=ped.tmp$mother, fathers=ped.tmp$father,
             generations=ped.tmp$gen, main="Pedigree of the controlled crosses")
```

Check additive genetic relationships:
```{r, add_gen_rel}
A.vr.cross <- estimGenRel(X=rbind(genos.gdparents, genos.par, genos.offs),
                         afs=afs.hwe, method="vanraden1")
A.t.cross <- estimGenRel(X=rbind(genos.gdparents, genos.par, genos.offs),
                         afs=afs.hwe, method="toro2011_eq10")
cor(c(A.vr.cross), c(A.t.cross))
imageWithScale(A.vr.cross, main="Additive genetic relationships of crosses")
imageWithScale(A.vr.cross[1:10, 1:10],
               main="Additive genetic relationships of crosses (subset)",
               idx.rownames=1:10, idx.colnames=1:10)
summary(diag(A.vr.cross))
summary(A.vr.cross[upper.tri(A.vr.cross)])
summary(A.vr.cross[name.par1, grep("off", colnames(A.vr.cross))])
summary(A.vr.cross[name.par2, grep("off", colnames(A.vr.cross))])
summary(A.t.cross[name.par1, grep("off", colnames(A.t.cross))])
summary(A.t.cross[name.par2, grep("off", colnames(A.t.cross))])
```

As expected, the additive genetic relationships between all parent-child pairs are around 0.5, corresponding to a coancestry coefficient of 1/4.

Impact of not using the allele frequencies estimated from the original population at HWE, on the estimation of additive genetic relationships:
```{r, add_gen_rel_allfreq}
A.vr.cross <- estimGenRel(X=rbind(genos.gdparents, genos.par, genos.offs),
                          method="vanraden1")
A.t.cross <- estimGenRel(X=rbind(genos.gdparents, genos.par, genos.offs),
                         method="toro2011_eq10")
cor(c(A.vr.cross), c(A.t.cross))
summary(A.vr.cross[name.par1, grep("off", colnames(A.vr.cross))])
summary(A.vr.cross[name.par2, grep("off", colnames(A.vr.cross))])
summary(A.t.cross[name.par1, grep("off", colnames(A.t.cross))])
summary(A.t.cross[name.par2, grep("off", colnames(A.t.cross))])
```

In that case, we don't retrieve the theoretical coancestry coefficient of 1/4.
But note that for predicting BLUPs of breeding values, it should not matter (see Stranden and Christensen, 2011).


# Mimick genotyping by sequencing

Discard SNP genotypes according to a "GBS" design, i.e. all genotypes and SNPs have NA, but with a skewed distribution (30% of the sites with 0-0.05% NA, 17% of the sites with 0.05-0.1% NA, 8% with 0.1-0.15% NA, and the remaining sites with a uniformly-increasing amount of NA):
```{r, gbs}
dim(X <- rbind(genos.gdparents, genos.par, genos.offs))
X.na <- X
bin.width <- 0.05
max.prop <- 0.9
bin.mids <- seq(from=bin.width/2, to=max.prop, by=bin.width)
bin.heights <- c(0.3, 0.17, 0.08,
                 rep((1-(0.3+0.17+0.08))/(length(bin.mids) - 3),
                     length(bin.mids) - 3))
stopifnot(sum(bin.heights) == 1)
assigned.bins <- cut(x=1:ncol(X),
                     breaks=floor(ncol(X) * cumsum(c(0, bin.heights))),
                     labels=FALSE)
assigned.bins <- sample(assigned.bins) # shuffle the bins among SNPs
table(assigned.bins)
bin.probs <- seq(from=0, to=max.prop, length.out=length(bin.mids))
## for each SNP, sample genotype indices to set as missing
idx1 <- sapply(1:ncol(X), function(j){
  sample.int(n=nrow(X), size=round(bin.probs[assigned.bins[j]] * nrow(X)))
})
idx2 <- do.call(c, lapply(1:length(idx1), function(j){
  ((j-1) * nrow(X)) + idx1[[j]]
}))
X.na[idx2] <- NA
sum(is.na(X.na))
sum(is.na(X.na)) / length(X.na)
```

Plot grid of missing SNP genotypes:
```{r}
plotGridMissGenos(X=X.na)
```

Plot histogram of NA proportions (same as fig S4 of Chan et al, 2016):
```{r}
miss.snps <- calcFreqMissSnpGenosPerSnp(X=X.na)
hist(miss.snps, breaks=seq(0,1,0.05), col="grey", border="white",
     main="Distribution of the proportion of missing data per biallelic SNP",
     xlab="proportion of missing data at a given SNP",
     ylab="number of SNPs")
```


# Perform imputation

```{r, imputation}
snp.coords <- genomes$snp.coords[colnames(X.na),]
snp.coords$chr <- as.numeric(sub("chr", "", snp.coords$chr))
system.time(
    out.FI <- runFimpute(X=X.na, snp.coords=snp.coords, clean="all"))
names(out.FI)
X.imp.FI <- out.FI$genos.imp
```

Use the pedigree:
```{r, imputation_ped}
ped.FI <- data.frame(id=ped$ind,
                     father=ped$father,
                     mother=ped$mother,
                     sex=NA,
                     stringsAsFactors=FALSE)
ped.FI$sex[1:6] <- c("F","M","F","M", "F","M")
ped.FI$sex[-c(1:6)] <- sample(c("F","M"), size=nrow(ped.FI)-6, replace=TRUE)
head(ped.FI, 10)
system.time(
    out.FI.ped <- runFimpute(X=X.na, snp.coords=snp.coords,
                             ped=ped.FI, clean="all"))
X.imp.FI.ped <- out.FI.ped$genos.imp
```


# Assess accuracy

Accuracy of imputation without the pedigree:
```{r, assess_accuracy}
length(X)
sum(X.na == X, na.rm=TRUE)
sum(is.na(X.na))
sum(X.imp.FI == X)
sum(X.imp.FI != X)
sum(X.imp.FI != X) / sum(is.na(X.na))
```

Check additive genetic relationships:
```{r}
A.vr.cross.imp <- estimGenRel(X=X.imp.FI, afs=afs.hwe)
summary(A.vr.cross.imp[name.par1, grep("off", colnames(A.vr.cross.imp))])
summary(A.vr.cross.imp[name.par2, grep("off", colnames(A.vr.cross.imp))])
```

Accuracy of imputation with the pedigree:
```{r, assess_accuracy_ped}
sum(X.imp.FI.ped == X)
sum(X.imp.FI.ped != X)
sum(X.imp.FI.ped != X) / sum(is.na(X.na))
```

Comparison:
```{r}
length(idx.pedwrong <- which(X.imp.FI == X & X.imp.FI.ped != X))
length(idx.pedright <- which(X.imp.FI.ped == X & X.imp.FI != X))
```


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
