---
title: "Introduction to the multivariate 'animal model'"
author: "Timothée Flutre (INRAE)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  rmarkdown::html_vignette:
    toc: true
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
---

<!--
This R chunk is used to set up important options.
-->
```{r, echo=FALSE}
if(! "params" %in% ls())
  params <- list(root.dir=NULL)
knitr::opts_chunk$set(fig.align="center")
```


# Preamble

License: [CC BY-SA 4.0](http://creativecommons.org/licenses/by-sa/4.0/)

References:

* [Lynch & Walsh (1998)](http://www.worldcat.org/isbn/0878934812)

* [Henderson & Quaas (1976)](http://www.journalofanimalscience.org/content/43/6/1188.abstract)

* [Magnus & Neudecker (2007)](http://www.worldcat.org/isbn/0471986321)

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Set-up

Load the required external packages:
```{r load_pkg}
options(digits=5)
suppressPackageStartupMessages(library(scrm))
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(rutilstimflutre))
```

Set the default values of the parameters:
```{r}
inferPkg <- ""
```

Get the settings from the command-line:
```{r}
if(! is.null(params$rmd.params))
  for(i in seq_along(params$rmd.params))
    assign(names(params$rmd.params)[i], params$rmd.params[[i]])
```

Check the parameters:
```{r}
stopifnot(inferPkg %in% c("", "lme4", "sommer", "breedR", "rstan", "jags"))
if(! inferPkg == ""){
  suppressPackageStartupMessages(library(package=inferPkg, character.only=TRUE))
}
```


# Statistical model

Notations:

* $I$: number of genotypes

* $T$: number of traits

* $Q$: number of replicates

* $N$: number of phenotypes per trait (here $I \times Q$)

* $P$: number of SNPs

* $Y$: $N \times T$ matrix of phenotypes

* $W$: $N \times Q$ design matrix relating phenotypes to replicates

* $C$: $Q \times T$ matrix of "fixed-effect" parameters

* $Z$: $N \times I$ design matrix relating phenotypes to genotypes

* $G_A$: $I \times T$ matrix of "random" additive genotypic values, so-called "breeding values"

* $A$: $I \times I$ matrix of additive genetic relationships

* $X$: $I \times P$ matrix of bi-allelic SNP genotypes encoded as allele dose in ${0,1,2}$

* $V_{G_A}$: $T \times T$ matrix of genetic variances and covariances

* $E$: $N \times T$ matrix of errors

* $V_E$: $T \times T$ matrix of error variances and covariances

Model:

\[
Y = W C + Z G_A + E
\]

with:

* $G_A \sim \mathcal{MN}(0, A, V_{G_A})$;

* $E \sim \mathcal{MN}(0, Id, V_E)$.

where $\mathcal{MN}_{n \times d}(M, U, V)$ represents a [matrix Normal distribution](https://en.wikipedia.org/wiki/Matrix_normal_distribution) with mean matrix $M$ of dimension $n \times d$, and variance-covariance matrices $U$ and $V$ of respective dimensions $n \times n$ and $d \times d$.

Given that any matrix Normal distribution is equivalent to a multivariate Normal distribution, $\text{MVN}(0, V \otimes U)$ of dimension $nd \times 1$, it is useful to re-write the likelihood using the [$vec$ operator](https://en.wikipedia.org/wiki/Vectorization_(mathematics)) and [Kronecker product](https://en.wikipedia.org/wiki/Kronecker_product) ($\otimes$):

\[
vec(Y) = (Id_T \otimes W) vec(C) + (Id_T \otimes Z) vec(G_A) + vec(E)
\]

with:

* $vec(G_A) \sim \text{MVN}(0, V_{G_A} \otimes A)$;

* $vec(E) \sim \text{MVN}(0, V_E \otimes Id_N)$.


# Simulate data

To allow reproducibility, set the seed of the generator of pseudo-random numbers:
```{r set_seed}
set.seed(1859)
```

Set dimensions:
```{r set_dim}
T <- 2     # number of traits (don't change in this vignette)
I <- 100   # number of genotypes
Q <- 10    # number of replicates per genotype
N <- I * Q # number of phenotypes per trait
```

## Genotypes

Simulate haplotypes via the [coalescent](https://en.wikipedia.org/wiki/Coalescent_theory) with recombination, and encode the corresponding bi-allelic SNP genotypes additively as allele doses in ${0,1,2}$:
```{r simul_genos}
Ne <- 10^4
chrom.len <- 10^5
mu <- 10^(-8)
c <- 10^(-8)
genomes <- simulCoalescent(nb.inds=I,
                           pop.mut.rate=4 * Ne * mu * chrom.len,
                           pop.recomb.rate=4 * Ne * c * chrom.len,
                           chrom.len=chrom.len)
X <- genomes$genos
(P <- ncol(X))
```

Estimate additive genetic relationships:
```{r estim_A}
A <- estimGenRel(X, relationships="additive", method="vanraden1", verbose=0)
summary(diag(A)) # under HWE, average should be 1
summary(A[upper.tri(A)]) # under HWE, average should be 0
```

## Phenotypes

Choose the parameters:
```{r}
mu <- c(50, 20) # global means
mean.C <- 5
sd_C <- 2
var_G_1 <- 2 # additive genotypic variance of trait 1
var_G_2 <- 4 # additive genotypic variance of trait 2
cor_G <- 0.7 # additive genotypic correlation between traits 1 and 2
cov.G <- cor_G * sqrt(var_G_1) * sqrt(var_G_2)
(V_G_A <- matrix(c(var_G_1, cov.G, cov.G, var_G_2), nrow=2, ncol=2))
kappa(V_G_A)
h2_1 <- 0.7 # var_G_1 / (var_G_1 + var_E_1 / Q)
h2_2 <- 0.3 # var_G_2 / (var_G_2 + var_E_2 / Q)
## var_E_1 <- TODO
## var_E_2 <- TODO
var_E_1 <- 5
var_E_2 <- 4
cor_E <- 0 #-0.2
cov_E <- cor_E * sqrt(var_E_1) * sqrt(var_E_2)
(V_E <- matrix(c(var_E_1, cov_E, cov_E, var_E_2), nrow=2, ncol=2))
kappa(V_E)
```

Simulate phenotypes:
```{r simul_phenos}
model <- simulAnimalModel(T=T, Q=Q, mu=mu, mean.C=mean.C, sd.C=sd_C,
                          A=A, V.G.A=V_G_A, V.E=V_E)
str(model$data)
head(model$data)
regplot(model$G.A[,1], model$G.A[,2],
        xlab="G_A of trait1", ylab="G_A of trait2", main="Genotypic covariation",
        show.crit=c("corP", "R2adj"), legend.x="bottomright")
abline(v=0, h=0, col="lightgrey")
```

Save the truth:
```{r}
truth <- list(T=T, I=I, Q=Q, N=N,
              A=A)
truth <- append(truth, model)
truthFile <- "intro-mvAM_truth.rds"
saveRDS(object=truth, file=truthFile)
tools::md5sum(truthFile)
```


# Explore the data

Data in the "wide" format:
```{r}
dat <- model$data
str(dat)
head(dat)
```

Reformat in the "long" format:
```{r}
datLong <- data.frame(geno=rep(dat$geno, times=2),
                      trait=rep(c("response1","response2"), each=nrow(dat)),
                      year=rep(dat$year, times=2),
                      value=c(dat$response1, dat$response2),
                      stringsAsFactors=TRUE)
str(datLong)
head(datLong)
```

Summary statistics:
```{r smry_stats}
do.call(rbind, tapply(dat$response1, dat$year, summary))
do.call(rbind, tapply(dat$response2, dat$year, summary))
```

Plots:
```{r plots_data}
hist(dat$response1, breaks="FD", col="grey", border="white", las=1,
     main="Trait 1")
hist(dat$response2, breaks="FD", col="grey", border="white", las=1,
     main="Trait 2")
regplot(dat$response1, dat$response2, asp=1,
        xlab="Trait1", ylab="Trait2", main="Phenotypic covariation",
        show.crit=c("corP", "R2adj"), legend.x="bottomright")
abline(v=mu[1], h=mu[2], col="lightgrey")
op <- par(mar=c(8,4,4,2)+0.1)
boxplot(value ~ year + trait, data=datLong, las=2,
        main="Phenotypes per trait and year", xlab="")
par(op)
```


# Fit the model to data

## Vie ReML (`lme4`)

[Trick](https://mac-theobio.github.io/QMEE/lectures/MultivariateMixed.notes.html)
```{r fit_lmer}
if(inferPkg == "lme4"){
  st <- system.time(
      fit.lmer <- lmer(value ~ -1 + trait + year + trait:year + (trait-1|geno),
                       data=datLong,
                       control=lmerControl(optimizer="bobyqa",
                                           check.nobs.vs.nlev="ignore",
                                           check.nobs.vs.nRE="ignore")))
  print(st)
}
```

## Via ReML (`breedR`)

```{r fit_remlf90}
if(inferPkg == "breedR"){
  st <- system.time(
      fit.remlf90 <- remlf90(fixed=cbind(response1, response2) ~ year,
                             generic=list(add=list(model$Z, A)),
                             data=dat))
  print(st)
}
```

## Via ReML (`sommer`)

```{r fit_mmer}
if(inferPkg == "sommer"){
  st <- system.time(
      fit.mmer <- mmer(fixed=cbind(response1, response2) ~ year,
                       random= ~ vsr(geno, Gu=A, Gtc=unsm(T)),
                       rcov= ~ vsr(units, Gtc=unsm(T)),
                       data=dat,
                       method="NR"))
  print(st)
}
```

## Via Gibbs (`rjags`)

See code in the appendix of https://arxiv.org/abs/1507.08638

```{r, fit_jags}
tmp.file.rjags <- "intro-mvAM_fit-rjags.rds"
if(inferPkg == "rjags"){
  if(! file.exists(tmp.file.rjags)){
    ## TODO
    ## forJags <- list(N=N, m=m, d=d,
    ##                 y=y[1:d,1:N], x=x[1:m,],
    ##                 b0=rep(0, m), B0=diag(0.0001, m),
    ##                 Ku=diag(1, d), Ke=diag(1, d),
    ##                 Z=rep(0, d), r=r)
    st <- system.time(
        fit.rjags <- jagsAMmv(data=tmp, relmat=list(geno.add=A),
                              nb.chains=nb.chains, nb.iters=nb.iters,
                              burnin=burnin, thin=thin,
                              rm.jags.file=TRUE, verbose=1))
    print(st)
    saveRDS(fit.rjags, tmp.file.rjags)
  } else{
    print("load tmp file...")
    fit.rjags <- readRDS(tmp.file.rjags)
  }
}
```

## Via HMC (`rstan`)

```{r fit_stan}
tmp.file.rstan <- "intro-mvAM_fit-rstan.rds"
if(inferPkg == "rstan"){
  if(! file.exists(tmp.file.rstan)){
    nb.chains <- 1
    burnin <- 1 * 10^3
    thin <- 1
    nb.usable.iters <- 1 * 10^3
    nb.iters <- ceiling(burnin + (nb.usable.iters * thin) / nb.chains)
    tmp <- dat
    colnames(tmp)[colnames(tmp) == "geno"] <- "geno.add"
    st <- system.time(
        fit.rstan <- stanAMmv(data=tmp, relmat=list(geno.add=A),
                              nb.chains=nb.chains, nb.iters=nb.iters,
                              burnin=burnin, thin=thin,
                              task.id="intro-mvAM",
                              verbose=1,
                              rm.stan.file=TRUE, rm.sm.file=TRUE))
    print(st)
    saveRDS(fit.rstan, tmp.file.rstan)
  } else{
    print("load tmp file...")
    fit.rstan <- readRDS(tmp.file.rstan)
  }
}
```


# Evaluate parameter estimates

## From ReML (`lme4`)

```{r eval_lmer}
if(inferPkg == "lme4"){
  print(summary(fit.lmer))
  print(VarCorr(fit.lmer))
  print(as.data.frame(VarCorr(fit.lmer)))
  print(truth$V.G.A)
  print(truth$V.E)
}
```

## From ReML (`breedR`)

```{r eval_remlf90}
if(inferPkg == "breedR"){
  print(summary(fit.remlf90))
}
```

## From ReML (`sommer`)

```{r eval_mmer}
if(inferPkg == "sommer"){
  names(fit.mmer)
  print(fit.mmer$sigma)
  print(truth[c("V.G.A","V.E")])

  ## rmse(c(fit.mmer$var.comp$add) - c(V_G_A))
  ## rmse(fit.mmer$var.comp$add[1,1] - V_G_A[1,1])
  ## rmse(fit.mmer$var.comp$add[2,2] - V_G_A[2,2])
  ## rmse(fit.mmer$var.comp$add[1,2] - V_G_A[1,2])

  ## rmse(c(fit.mmer$var.comp$Residual) - c(V_E))
  ## rmse(fit.mmer$var.comp$Residual[1,1] - V_E[1,1])
  ## rmse(fit.mmer$var.comp$Residual[2,2] - V_E[2,2])
  ## rmse(fit.mmer$var.comp$Residual[1,2] - V_E[1,2])

  rmse(fit.mmer$u.hat$add[,1] - model$G.A[,1])
  rmse(fit.mmer$u.hat$add[,2] - model$G.A[,2])
  cor(model$G.A[,1], fit.mmer$u.hat$add[,1])
  cor(model$G.A[,2], fit.mmer$u.hat$add[,2])
  regplot(model$G.A[,1], fit.mmer$u.hat$add[,1], xlab="true G_A[,1]",
          ylab="BLUP G_A[,1]")
  regplot(model$G.A[,2], fit.mmer$u.hat$add[,2], xlab="true G_A[,2]",
          ylab="BLUP G_A[,2]")
}
```

## From HMC (`rstan`)

Assess convergence:
```{r fit_stan_cvg, eval=TRUE}
## tmp <- grep("^corr[GE]\\[", names(fit.rstan), value=TRUE)
tmp <- c("corrG[1,1]", "corrG[1,2]", "corrG[2,2]",
         "corrE[1,1]", "corrE[1,2]", "corrE[2,2]")
rstan::traceplot(fit.rstan, pars=tmp)
print(fit.rstan, pars=tmp)
```


# Conclusions

The ReML procedure is fast, but it does not quantify the uncertainty in variance components.
It hence is required to use an additional procedures to do this, e.g. the bootstrap, which can then be computationally costly.


# Appendix

```{r info}
t1 <- proc.time()
t1 - t0
print(sessionInfo(), locale=FALSE)
```
