---
title: "Introduction to the Gibbs sampler"
author: "Timothée Flutre (Inra)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
output:
  rmarkdown::html_vignette:
    toc: true
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{Intro to Gibbs sampler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
setwd("~/src/rutilstimflutre/vignettes/")

library(devtools)
build_vignettes()

library(rmarkdown)
render("intro-Gibbs-sampler.Rmd", "html_document")
-->

# Preamble

License: [CC BY-SA 4.0](http://creativecommons.org/licenses/by-sa/4.0/)

References:

* Gelman *et al*, (2004), Bayesian Data Analysis

* http://galton.uchicago.edu/~eichler/stat24600/Handouts/s09.pdf


# Example based on the Normal model

## Data

Let us imagine we measured height of $N$ plants from the same wheat variety.
As the data display some variation, we will use $N$ random variables, $Y_1,\ldots,Y_N$, as a useful abstraction to capture such a variation.
For each of them, we have a single realization, noted $y_1,\ldots,y_N$.
The data set hence is:

\[
\mathcal{D} = \{y_1, \ldots, y_N \}
\]

Let us assume they all come from the same Normal distribution, specified by its mean, $\mu$, and variance, $\sigma^2$.
The parameter set hence is:

\[
\Theta = \{ \mu, \sigma^2 \}
\]

Furthermore, we also assume that all observations are exchangeable, i.e. conditionally independent given the parameters:

\[
\forall i \in \{1,\ldots,N\}, \; y_i | \mu, \sigma^2 \overset{\text{i.i.d.}}{\sim} \mathcal{N}(\mu, \sigma^2)
\]

The likelihood hence can be written as:

\begin{align}
\mathcal{L}(\Theta; \mathcal{D}) &= p(\mathcal{D} \, | \, \Theta) \\
&= \prod_{i=1}^N p(y_i \, | \, \mu, \sigma^2) \\
&= \prod_{i=1}^N \frac{1}{\sigma \sqrt{2 \pi}} \exp \left( - \frac{1}{2} \frac{(y_i - \mu)^2}{\sigma^2}\right)
\end{align}


## Simulation

```{r normal-model_simul, fig.width=7, fig.height=5}
set.seed(1859)
N <- 200
mu <- 52
sigma <- 4
y <- rnorm(n=N, mean=mu, sd=sigma)
hist(y, breaks="FD", col="grey", border="white", las=1)
```


## Inference

### Conjuguate prior assuming unknown mean but known variance

See section 2.6 "Estimating the mean of a normal distribution with known variance", from Gelman *et al* (2004), page 46.

The likelihood of a single data point, say $y_i$, is Normal, as the exponential of a quadratic form.
A conjuguate prior hence also is Normal: $\mu \sim \mathcal{N}(\mu_0, \sigma_0^2)$.
And so is the posterior: $\mu | y_i \sim \mathcal{N}(\mu_1, \sigma_1^2)$ with $\frac{1}{\sigma_1^2} = \frac{1}{\sigma_0^2} + \frac{1}{\sigma^2}$ and $\mu_1 = \frac{\frac{1}{\sigma_0^2} \mu_0 + \frac{1}{\sigma^2} y_i}{\frac{1}{\sigma_1^2}}$.

Now with all data points:

\begin{align}
p(\mu | \boldsymbol{y}) &= p(\mu) \prod_i p(y_i |\mu) \\
&\propto \exp \left( -\frac{1}{2} \left[ \frac{1}{\sigma_0^2} (\mu - \mu_0)^2 + \frac{1}{\sigma^2} \sum_i (y_i - \mu)^2 \right] \right)
\end{align}

After algebraic manipulations: $\mu | \boldsymbol{y} \sim \mathcal{N}(\mu_N, \sigma_N^2)$ with $\frac{1}{\sigma_N^2} = \frac{1}{\sigma_0^2} + \frac{N}{\sigma^2}$ and $\mu_N = \frac{\frac{1}{\sigma_0^2} \mu_0 + \frac{N}{\sigma^2} \bar{y}}{\frac{1}{\sigma_1^2}}$.

When assuming unknown mean but known variance, we can see that, with $N$ fixed, as $\sigma_0^2 \rightarrow + \infty$, then $\mu|\boldsymbol{y} \sim \mathcal{N}(\bar{y}, \sigma^2 / N)$.


### Conjuguate prior assuming unknown variance but known mean

See section 2.7 "Other standard single-parameter models", from Gelman *et al* (2004), page 49.

The likelihood is: $p(\boldsymbol{y} | \sigma^2) \propto (\sigma^2)^{-N/2} \exp \left( - \frac{N}{2 \sigma^2} v \right)$, where $v = \frac{1}{N} \sum_i (y_i - \mu)^2$.

The corresponding conjuguate prior is: $p(\sigma^2) \propto (\sigma^2)^{-(\nu_0+1)} \exp \left( - \frac{\sigma_0^2}{\sigma^2} \right)$.
It is a scaled inverse-$\chi^2$ distribution with $\nu_0$ degrees of freedom and scale $\sigma_0^2$.
It is equivalent to an inverse gamma distribution, $\text{Inv}-\Gamma$, with shape $\nu_0 / 2$ and scale $(\nu_0 / 2) \times \sigma_0^2$ (see page 574 of Gelman *et al*).

The resulting posterior is: $\sigma^2 | \boldsymbol{y} \sim \text{Inv}-\chi^2 \left( \nu_0 + N, \frac{\nu_0 \sigma_0^2 + N v}{\nu_0 + N} \right)$.

If the prior degrees of freedom $\nu_0$ is small relative to the data degrees of freedom $N$, then the posterior is approximately as if $\nu_0 = 0$, that is: $\sigma^2 | \boldsymbol{y} \sim \text{Inv}-\chi^2 (N, v)$.
This limiting form of the posterior also corresponds to the prior $p(\sigma^2) \propto 1 / \sigma^2$.


### Non-informative prior assuming unknown mean and variance

See section 2.9 "Noninformative prior distributions", from Gelman *et al* (2004), page 61.

Realizing that $\mu$ is a pure *location parameter* and aiming at a non-informative prior implies a Uniform distribution on $\mu$, that is $p(\mu) \propto \text{constant}$ over the range $]-\infty, +\infty[$.

Realizing that $\sigma$ is a pure *scale parameter* and aiming at a non-informative prior implies a Uniform distribution on $\log(\sigma)$, that is $p(\log(\sigma)) \propto 1$, or $p(\sigma^2) \propto 1 / \sigma^2$.

Here, we will thus choose a so-called "non-informative" prior using these Uniform distributions. See section 3.2 "Normal data with a noninformative prior distribution", from Gelman *et al* (2004), page 74.

\begin{align}
p(\Theta) &= p(\mu, \sigma^2) \\
&= p(\mu) \times p(\sigma^2) \\
&\propto 1 / \sigma^2
\end{align}

With this prior, the joint posterior is:

\begin{align}
p(\mu, \sigma^2 | \boldsymbol{y}) &\propto \sigma^{-N-2} \exp \left( - \frac{1}{2 \sigma^2} \sum_i (y_i - \mu)^2 \right) \\
&= \sigma^{-N-2} \exp \left( - \frac{1}{2 \sigma^2} [(N-1)s^2 + N(\bar{y} - \mu)^2] \right)
\end{align}

where $s^2 = \frac{1}{N-1} \sum_i (y_i - \bar{y}) ^2$.

However, we are mostly interested in $\mu$, and $\sigma^2$ is seen as a nuisance parameter.
Assuming $\sigma^2$ known, it is easy to see that the conditional posterior for $\mu$ is: $\mu | \sigma^2, \boldsymbol{y} \sim \mathcal{N}(\bar{y}, \sigma^2 / N)$.
Now, the marginal posterior for $\sigma^2$ is: $\sigma^2 | \boldsymbol{y} \sim \text{Inv}-\chi^2 (N-1, s^2)$.

As it rarely happens, for this model, we can also derive the marginal posterior distribution for $\mu$ in closed form, which is Student' $t$: $\mu | \boldsymbol{y} \sim t(\bar{y}, s^2 / N)$.


### Implementation

Even though we can here directly sample from the marginal posterior of $\mu$, this is rarely the case in practice.
But having closed form for $\sigma^2 | \boldsymbol{y}$ and $\mu | \sigma^2, \boldsymbol{y}$ allows to use the Gibbs sampler.
See section 11.3 "The Gibbs sampler" from Gelman *et al* (2004), page 287.

```{r normal-model_implem}
##' Gibbs sampler
##'
##' Perform inference with the Gibbs sampelr for the Normal model with non-informative priors for the mean and variance. Caution, the goal of the function is pedagogy, not efficiency.
##' @param y vector of data; missing data should be encoded as NA and will be ignored
##' @param seeds vector of seeds for the pseudo-random number generator; one per chain (see \code{\link[base]{set.seed}})
##' @param init list with initial parameters' values; same for all chains; if NULL, default values will be used
##' @param nb.chains number of chains
##' @param nb.iters number of iterations per chain; same for all chains
##' @param nb.cores number of cores to use (see \code{\link[parallel]{mclapply}))
##' @param verbose verbosity level (0/1)
##' @return list with one matrix per chain, with iterations in rows and parameters in columns
##' @author Timothee Flutre
##' @export 
gibbsSamplerNormalModel <- function(y, seeds=NULL, init=NULL, nb.chains=2,
                                    nb.iters=10^3, nb.cores=1, verbose=1){
  stopifnot(is.vector(y))
  if(! is.null(seeds))
    stopifnot(is.vector(seeds),
              length(seeds) == nb.chains)
  if(! is.null(init))
    stopifnot(is.list(init),
              ! is.null(names(init)),
              "sigma2" %in% names(init))

  ## discard the missing data and compute the sufficient statistics
  y <- y[stats::complete.cases(y)]
  N <- length(y)
  y.bar <- mean(y)
  s2 <- (1 / (N - 1)) * sum((y - y.bar)^2)
  if(verbose > 0){
    msg <- paste0("N=", N,
                  " y.bar=", format(y.bar, digits=2),
                  " s2=", format(s2, digits=2))
    write(msg, stdout())
  }

  ## prepare the requirements for the sampler
  if(is.null(seeds))
    seeds <- sample.int(n=nb.chains, size=10^6)
  rinvgamma <- function(n, shape, scale){
    stopifnot(shape > 0, scale > 0)
    1 / stats::rgamma(n=n, shape=shape, scale=1 /scale)
  }
  rscalinvchisq <- function(n, df, scale){
    stopifnot(df > 0, scale > 0)
    rinvgamma(n=n, shape=df / 2, scale=(df / 2) * scale)
  }

  ## run the Gibbs sampler
  post.samples <- parallel::mclapply(1:nb.chains, function(c){
    if(verbose > 0){
      write(paste("chain", c), stdout())
      pb <- txtProgressBar(style=3)
    }

    set.seed(seeds[c])
    out <- matrix(data=NA, nrow=nb.iters, ncol=2)
    colnames(out) <- c("sigma2", "mu")

    if(! is.null(init)){
      sigma2 <- init$sigma2
    } else
      sigma2 <- rscalinvchisq(n=1, df=N - 1, scale=s2)
    for(i in 1:nb.iters){
      if(verbose > 0)
        setTxtProgressBar(pb, i)
      mu <- stats::rnorm(n=1, mean=y.bar, sd=sqrt(sigma2 / N))
      sigma2 <- rscalinvchisq(n=1, df=N - 1, scale=s2)
      out[i, "mu"] <- mu
      out[i, "sigma2"] <- sigma2
    }

    close(pb)

    return(out)
  }, mc.silent=ifelse(verbose == 0, TRUE, FALSE), mc.cores=nb.cores)

  return(post.samples)
}
```


### Execution

```{r normal-model_exec, eval=TRUE}
out <- gibbsSamplerNormalModel(y=y, seeds=c(1859, 9581), nb.chains=2,
                               nb.iters=10^3, nb.cores=1, verbose=1)
```


## Evaluation

### Convergence

Requires the [coda](https://cran.r-project.org/web/packages/coda/) package:

```{r normal-model_eval, eval=TRUE, fig.width=7, fig.height=5}
if(require(coda)){
  out <- coda::as.mcmc.list(lapply(out, as.mcmc))
  for(c in 1:length(out)){
    print(paste("chain", c))
    rutilstimflutre::plotMcmcChain(out[[c]], "mu", pe=mu)
    rutilstimflutre::plotMcmcChain(out[[c]], "sigma2", pe=sigma^2)
  }
  print(gelman.diag(out))
  print(geweke.diag(out))
  print(heidel.diag(out))
  print(raftery.diag(out, q=0.5, r=0.05)  )
}
```

Based on the diagnostics, we can assume that convergence has occurred.
Moreover, a small number of burn-in iterations can be discarded, but no thinning is necessary.


### Posteriors

```{r normal-model_posteriors, fig.width=7, fig.height=5}
if(require(coda)){
  burnin <- 100
  thin <- 1
  out <- window(x=out, start=1 + burnin, end=nrow(out[[1]]), thin=thin)
  print(summary(out))
  print(effectiveSize(out))
  plot(as.vector(out[[1]][,"mu"]), as.vector(out[[1]][,"sigma2"]),
       xlab="mu", ylab="sigma2", las=1,
       main="Joint posterior distribution")
  points(as.vector(out[[1]][,"mu"]), as.vector(out[[1]][,"sigma2"]))
}
```


# Example based on the multiple linear regression

## Data

...

## Simulation

...

## Inference

...

## Implementation

...

## Evaluation

...


# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
