---
title: "Introduction to the statistical analysis of a multi-environment trial"
author: "Timothée Flutre (INRAE)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 6
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
---

```{r knitr_setup, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=6, fig.height=5, fig.align="center",
                      global.par=TRUE)
```


# Preamble

Dependencies:
```{r load_pkg}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(Matrix))
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(nlme))
suppressPackageStartupMessages(library(MM4LMM))
```

Execution time (see the appendix):
```{r time_0}
t0 <- proc.time()
```


# Model

The goal of this document is to introduce linear models to analyze trials with *genotype-environment interactions* allowing for *heteroscedasticity* in the errors.

## Overview

A *multi-environment trial* (MET) consists in $I$ varieties grown in $J$ environments.

* At the scale of the whole trial, not all varieties are necessarily grown in all environments, in which case the MET is *incomplete*.
However, it is still much advised that all varieties are grown in the same number of environments, and well distributed among them, so that the MET is *balanced*.

* At the scale of a given environment, not all varieties grown in it are necessarily replicated.
But at least some of them should be, e.g., to deal with spatial heterogeneity.

When the design is complete and with the same replicate number of varieties in each environment, one may choose to model the variety and environment main effects as *fixed*.
Otherwise, one may also choose to model both effects as *random*, or one as fixed and the other as random.
When a main effect is modeled as fixed, a *contrast* needs to be chosen, typically the "sum" contrast.

## Formulas

Yield of the $k$-th replicate of variety $i$ in environment $j$:
\[
y_{ijk} = \mu + \alpha_i + \theta_j + \delta_{ij} + \epsilon_{ijk}
\]
where $\mu$ is the overall intercept, $\alpha_i$ is the main effect for variety $i$, $\theta_j$ is the main effect for environment $j$, $\delta_{ij}$ is the interaction effect for variety $i$ and environment $j$, and $\epsilon_{ijk}$ is the residual term for replicate $k$ of this variety-by-environment combination.

Let $n_j$ denotes the number of plots in environment $j$, hence the MET sample size, i.e., the total number of plots, is defined as $n := \sum_j n_j$.

In vector form:
\begin{align}
\boldsymbol{y} = \boldsymbol{\text{1}} \mu +  M_\alpha \, \boldsymbol{\alpha} + M_\theta \, \boldsymbol{\theta} + M_\delta \, \boldsymbol{\delta} + \boldsymbol{\epsilon} \text{ with } \epsilon_{ijk} \sim \mathcal{N}(0, \sigma_j^2)
\end{align}
where $\boldsymbol{y}$ and $\boldsymbol{\epsilon}$ are vectors of length $n$, the $M$'s are design matrices, and the lengths of vectors $\boldsymbol{\alpha}$, $\boldsymbol{\theta}$ and $\boldsymbol{\delta}$ depend on whether these effects are modeled as fixed or random.

Such a model can be generically written like this:
\[
\boldsymbol{y} = X \boldsymbol{\beta} + \sum_{k=1}^K Z_k \boldsymbol{u}_k
\]
where $\boldsymbol{\beta}$ are fixed effects with $X$ as design matrix, the $\boldsymbol{u}_k$'s are random effects (including the error terms) with $Z_k$ as design matrix and $\boldsymbol{u}_k \sim \mathcal{N}(\boldsymbol{0}, \sigma_k^2 R_k)$ where $\sigma_k^2$ is a variance component and $R_k$ a correlation matrix.


# Simulation

```{r}
set.seed(12345)
```

Dimensions of the MET assuming that it is complete and that each environment has a randomized complete block design with the same number of blocks:
```{r}
I <- 20 # nb of varieties
nbSites <- 4; nbSeasons <- 2
J <- nbSites * nbSeasons # nb of environments
K <- 5 # nb of blocks; assumed
```

MET sample size:
```{r}
list_n_j <- lapply(1:J, function(j){
  I * K
})
(n <- sum(do.call(c, list_n_j)))
```

## Identifiers and global data frame

Genotype identifiers:
```{r}
genos <- sprintf("g%03i", 1:I)
```

Environment identifiers:
```{r}
sites <- sort(unique(sapply(1:nbSites, function(i){
  paste(sample(LETTERS, 4), collapse="")})))
seasons <- seq(2015, 2015 + nbSeasons - 1)
envs <- apply(cbind(rep(seasons, times=nbSites),
                    rep(sites, each=nbSeasons)),
              1, paste, collapse="")
envs
names(list_n_j) <- envs
```

Block identifiers:
```{r}
blocks <- LETTERS[1:K]
```

Global data frame:
```{r}
dat <- data.frame(env=rep(envs, each=I*K),
                  geno=rep(genos, times=K*J),
                  gxe=NA,
                  block=rep(rep(blocks, each=I), times=J),
                  yield=NA,
                  stringsAsFactors=TRUE)
dat$env <- factor(as.character(dat$env), levels=envs)
dat$gxe <- apply(dat, 1, function(x){paste0(x["geno"], x["env"], collapse="-")})
## let the levels of the "gxe" column be ordered as varieties within environments:
tmp <- cbind("geno"=rep(genos, times=J),
             "env"=rep(envs, each=I))
tmp <- as.data.frame(tmp)
tmp$gxe <- apply(tmp, 1, function(x){paste0(x["geno"], x["env"], collapse="-")})
dat$gxe <- factor(dat$gxe, levels=tmp$gxe)
str(dat)
stopifnot(nrow(dat) == n,
          nlevels(dat$geno) == I,
          nlevels(dat$env) == J,
          nlevels(dat$gxe) == I * J)
head(dat)
```

## $X$

We will model here the environment main effects as fixed.

Design matrix of the fixed effects, assuming an intercept and a "sum" contrast:
```{r}
X <- model.matrix(~ 1 + env, data=dat, contrasts.arg=list(env="contr.sum"))
stopifnot(nrow(X) == n,
          ncol(X) == 1 + (J - 1))
dim(X)
head(X)
tail(X)
image(t(X)[,nrow(X):1], main="X", col=c("grey","white","black"),
      xaxt="n", yaxt="n")
```

## $\beta$

Overall intercept:
```{r}
mu <- 50
```

Contrasts of the main environment effects modeled as fixed:
```{r}
var_theta <- 20
theta <- mvrnorm(n=1, mu=rep(0, J-1), Sigma=var_theta * diag(J-1))
summary(theta)
```

Vector of fixed effects of interest:
```{r}
beta <- c(mu, theta)
length(beta)
stopifnot(length(beta) == 1 + (J - 1))
```

Retrieve the main environment effects (not their contrasts):
```{r}
coding_matrix <- cbind(intercept=1,
                       contr.sum(levels(dat$env)))
dim(coding_matrix)
(trueEnvEffs <- (coding_matrix %*% beta)[,1])
```

TODO: add block effects (nested in the environments)

## $Z$

Design matrix of the random effects:
```{r}
Z_g <- model.matrix(~ 0 + geno, data=dat)
Z_gxe <- model.matrix(~ 0 + geno:env, data=dat)
Z <- cbind(Z_g, Z_gxe)
stopifnot(nrow(Z) == n,
          ncol(Z) == I + (I * J))
dim(Z)
head(Z)[,1:6]
tail(Z)[,tail(colnames(Z))]
image(t(Z)[,nrow(Z):1], main="Z", col=c("white","black"),
      xaxt="n", yaxt="n")
```

## $u$

Main variety effects modeled as random, assuming no correlation between genotypic values:
```{r}
var_alpha <- 10
R_alpha <- diag(I) # could be replaced by a kinship matrix
alpha <- mvrnorm(n=1, mu=rep(0, I), Sigma=var_alpha * R_alpha)
names(alpha) <- levels(dat$geno)
summary(alpha)
```

Variety-environment effects:
```{r}
var_delta <- 3
R_delta <- diag(I*J)
delta <- mvrnorm(n=1, mu=rep(0, I*J), Sigma=var_delta * R_delta)
names(delta) <- levels(dat$gxe)
summary(delta)
```

Vector of random effects:
```{r}
u <- c(alpha, delta)
length(u)
stopifnot(length(u) == I + (I * J))
```

## $\epsilon$

Error terms:
```{r}
## type_err_var <- "homoscedasticity"
type_err_var <- "heteroscedasticity"
list_var_j <- lapply(1:J, function(j){
  if(type_err_var == "homoscedasticity"){
    2
  } else if(type_err_var == "heteroscedasticity")
    runif(n=1, min=2, max=8)
})
names(list_var_j) <- envs
unlist(list_var_j)
summary(unlist(list_var_j))
list_e_j <- lapply(1:J, function(j){
  n_j <- sum(dat$env == envs[j])
  rnorm(n=n_j, mean=0, sd=sqrt(list_var_j[[j]]))
})
epsilon <- do.call(c, list_e_j)
length(epsilon)
stopifnot(length(epsilon) == n)
```

## $y$

```{r}
y <- X %*% beta + Z %*% u + epsilon
length(y)
stopifnot(length(y) == n)
```

```{r}
dat$yield <- y[,1]
```

## Save

```{r}
outF <- "intro-het-GxE_dat.csv"
write.table(dat, outF, sep="\t", row.names=FALSE)
tools::md5sum(outF)
```


# Exploration

```{r}
str(dat)
summary(dat)
```

## Overall

```{r}
breaks <- pretty(range(dat$yield),
                 n = nclass.FD(dat$yield),
                 min.n = 1)
p <- ggplot(dat, aes(x=yield)) +
  labs(title="Simulation", x="yield") +
  geom_histogram(color="white", breaks=breaks)
p
```

## Per environment

```{r}
plotPerEnv <- function(dat){
  ggplot() +
    labs(title="Simulation", x="environment", y="yield") +
    geom_violin(data=dat, aes(x=env, y=yield, fill=env)) +
    geom_point(data=dat, aes(x=env, y=yield, fill=env),
               position=position_jitterdodge(seed=1, dodge.width=0.9),
               show.legend=FALSE) +
    theme(axis.text.x = element_text(angle = 90))
}
```

```{r, fig.width=12}
plotPerEnv(dat)
```

## Per genotype

```{r, fig.width=12}
p <- ggplot(dat, aes(x=geno, y=yield, fill=geno)) +
  labs(title="Simulation", x="genotype", y="yield") +
  geom_violin() +
  geom_point(position=position_jitterdodge(seed=1, dodge.width=0.9),
             show.legend=FALSE)
p
```


# Statistical analysis

## Two-stage

### `lm` then `coef`

#### Run

```{r}
fits1 <- lapply(1:J, function(j){
  sub_dat <- droplevels(subset(dat, env == envs[j]))
  ## lm(yield ~ 0 + geno + block, data=sub_dat)
  lm(yield ~ 0 + geno, data=sub_dat)
})
names(fits1) <- envs
coefs1 <- lapply(seq_along(fits1), function(j){
  env <- envs[j]
  fit <- fits1[[env]]
  tmp <- coef(fit)
  tmp <- tmp[grep("^geno", names(tmp))]
  names(tmp) <- sub("^geno", "", names(tmp))
  data.frame(env=env, geno=names(tmp), estim=tmp, stringsAsFactors=TRUE)
})
df_coefs1 <- do.call(rbind, coefs1)
rownames(df_coefs1) <- NULL
head(df_coefs1)
```

Sort the environments per mean yield:
```{r}
meanEnvs <- sort(tapply(df_coefs1$estim, df_coefs1$env, mean))
summary(meanEnvs)
df_coefs1$env <- factor(as.character(df_coefs1$env), names(meanEnvs))
```

#### Check

##### Error variances

```{r}
(tmp <- data.frame(true=unlist(list_var_j),
                   estim=sapply(fits1, sigma)^2,
                   row.names=envs))
p <- ggplot(tmp, aes(x=true, y=estim)) +
  labs(title=paste0("Error variances",
                    " (cor=", round(cor(tmp$true, tmp$estim), 2), ")"),
         x="true value", y="estimate") +
    geom_point() +
  geom_abline(intercept=0, slope=1)
p
```

##### Genotype effects per environment

```{r}
for(j in 1:J){
  tmp <- data.frame(true=trueEnvEffs[envs[j]] + alpha +
                      delta[grep(envs[j], names(delta))],
                    estim=coefs1[[j]]$estim)
  p <- ggplot(tmp, aes(x=true, y=estim)) +
    labs(title=paste0("Environment ", envs[j],
                      " (R2=", round(summary(fits1[[j]])$r.squared, 3), ")"),
         x="true value", y="estimate") +
    geom_point() +
    geom_abline(intercept=0, slope=1)
  print(p)
}
```

#### Plot

```{r, fig.width=12}
p <- ggplot(df_coefs1, aes(x=env, y=estim, group=geno)) +
  labs(title="Simulation",
       x="environment", y="marginal mean of genotype yield") +
  geom_line(aes(color=geno)) + geom_point(aes(color=geno)) +
  theme(axis.text.x = element_text(angle = 90))
p
```

### `lm` then `emmeans`

#### Run

```{r}
fits2 <- lapply(1:J, function(j){
  sub_dat <- droplevels(subset(dat, env == envs[j]))
  ## lm(yield ~ 1 + block + geno, data=sub_dat)
  lm(yield ~ 1 + geno, data=sub_dat)
})
names(fits2) <- envs
coefs2 <- lapply(seq_along(fits2), function(j){
  env <- envs[j]
  fit <- fits1[[env]]
  tmp <- summary(emmeans(object=fit, specs="geno"))
  tmp <- setNames(tmp$emmean, tmp$geno)
  data.frame(env=env, geno=names(tmp), estim=tmp, stringsAsFactors=TRUE)
})
df_coefs2 <- do.call(rbind, coefs2)
rownames(df_coefs2) <- NULL
head(df_coefs2)
```

Sort the environments per mean yield:
```{r}
meanEnvs <- sort(tapply(df_coefs2$estim, df_coefs2$env, mean))
summary(meanEnvs)
df_coefs2$env <- factor(as.character(df_coefs2$env), names(meanEnvs))
```

#### Check

##### Error variances

```{r}
(tmp <- data.frame(true=unlist(list_var_j),
                   estim=sapply(fits2, sigma)^2,
                   row.names=envs))
p <- ggplot(tmp, aes(x=true, y=estim)) +
  labs(title=paste0("Error variances",
                    " (R2=", round(summary(fits2[[j]])$r.squared, 3), ")"),
       x="true value", y="estimate") +
  geom_point() +
  geom_abline(intercept=0, slope=1)
p
```

##### Genotype effects per environment

```{r}
for(j in 1:J){
  tmp <- data.frame(true=trueEnvEffs[envs[j]] + alpha +
                      delta[grep(envs[j], names(delta))],
                    estim=coefs2[[j]]$estim)
  p <- ggplot(tmp, aes(x=true, y=estim)) +
    labs(title=paste0("Environment ", envs[j],
                      " (R2=", round(summary(fits2[[j]])$r.squared, 3), ")"),
         x="true value", y="estimate") +
    geom_point() +
    geom_abline(intercept=0, slope=1)
  print(p)
}
```

#### Plot

```{r, fig.width=12}
p <- ggplot(df_coefs2, aes(x=env, y=estim, group=geno)) +
  labs(title="Simulation",
       x="environment", y="marginal mean of genotype yield") +
  geom_line(aes(color=geno)) + geom_point(aes(color=geno)) +
  theme(axis.text.x = element_text(angle = 90))
p
```

## One-stage

### `lm`

The `lm` function from the `stats` package performs inference via ordinary least squares assuming homoscedasticity and cannot handle random variables.

#### Fit

```{r}
fits_lm <- list()
```

##### Without GxE

```{r}
## fit <- lm(model=yield ~ 1 + geno + env + geno:env + env/block, data=dat)
system.time(
    fit0 <- lm(yield ~ 1 + geno + env, data=dat,
               contrasts=list(geno="contr.sum", env="contr.sum")))
fits_lm[["woGxE"]] <- fit0
fixef0 <- coef(fit0)
err_var0 <- sigma(fit0)^2
scaled_residuals0 <- residuals(fit0) / err_var0
y_hat0 <- fitted(fit0)
in_diag0 <- cbind(dat,
                 "scaled_residuals"=scaled_residuals0,
                 "y_hat"=y_hat0)
```

##### With GxE

```{r}
## fit <- lm(model=yield ~ 1 + geno + env + geno:env + env/block, data=dat)
system.time(
    fit <- lm(yield ~ 1 + geno + env + geno:env, data=dat,
              contrasts=list(geno="contr.sum", env="contr.sum")))
fits_lm[["wGxE"]] <- fit
fixef <- coef(fit)
err_var <- sigma(fit)^2
scaled_residuals <- residuals(fit) / err_var
y_hat <- fitted(fit)
in_diag <- cbind(dat,
                 "scaled_residuals"=scaled_residuals,
                 "y_hat"=y_hat)
```

#### Model comparison

```{r}
AIC(fit0, fit)
BIC(fit0, fit)
```

```{r}
anova(fit0, fit)
```

#### Diagnostics

```{r}
par(mfrow=c(2,2))
plot(fit0, ask=FALSE)
```

```{r}
par(mfrow=c(2,2))
plot(fit, ask=FALSE)
```

#### Inference
      
##### Intercept

```{r}
data.frame("true"=mu,
           "estim"=summary(fit)$coefficients["(Intercept)", "Estimate"],
           "se"=summary(fit)$coefficients["(Intercept)", "Std. Error"])
```

##### Environment main effects

Contrasts:
```{r}
idx <- grep("^env", rownames(summary(fit)$coefficients))
(tmp <- cbind("true"=theta,
              "estim"=summary(fit)$coefficients[idx, "Estimate"],
              "se"=summary(fit)$coefficients[idx, "Std. Error"]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (lm)", pch=19); abline(a=0, b=1, lty=2)
```

Estimated marginal means:
```{r}
estimEnvEffs <- as.data.frame(emmeans(fit, "env"))$emmean
(tmp <- cbind("true"=trueEnvEffs,
              "estim"=estimEnvEffs))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (lm)", pch=19); abline(a=0, b=1, lty=2)
```

##### Genotype main effects

Estimated marginal means:
```{r}
estimGenoEffs <- as.data.frame(emmeans(fit, "geno"))$emmean
(tmp <- cbind("true"=mu + alpha,
              "estim"=estimGenoEffs))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Genotype main effects (lm)", pch=19); abline(a=0, b=1, lty=2)
```

##### Error variances

```{r}
(tmp <- cbind("true"=unlist(list_var_j),
              "estim"=err_var))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Error variances (lm)", pch=19); abline(a=0, b=1, lty=2)
```

##### PVE

```{r}
(tav <- anova(fit))
```

##### Plots

```{r, fig.width=12}
p <- plotPerEnv(dat)
p + geom_segment(data=data.frame(x=1:J - 0.3,
                                 xend=1:J + 0.3,
                                 y=estimEnvEffs,
                                 yend=estimEnvEffs),
                 aes(x=x, xend=xend, y=y, yend=yend), size=3)
```

### `gls` from `nlme`

The `gls` function from the `nlme` package performs inference via generalized least squares, but cannot handle random variables.

#### Fit

```{r}
fits_gls <- list()
fixefs <- list()
errVars <- list()
in_diags_gls <- list()
```

##### Assuming homoscedasticity

```{r}
options(contrasts=c("contr.sum", "contr.poly")) # gls() has no arg "contrasts"
## fit <- gls(model=yield ~ 1 + geno + env + geno:env + env/block, data=dat)
system.time(
    fit <- gls(model=yield ~ 1 + geno + env + geno:env, data=dat))
fits_gls[["hom"]] <- fit
fixefs[["hom"]] <- coef(fit)
errVars[["hom"]] <- fit$sigma^2
scaled_residuals_hom <- residuals(fit) / errVars[["hom"]]
y_hat_hom <- fitted(fit)
in_diags_gls[["hom"]] <- cbind(dat,
                               "scaled_residuals"=scaled_residuals_hom,
                               "y_hat"=y_hat_hom)
```

##### Assuming heteroscedasticity

```{r}
options(contrasts=c("contr.sum", "contr.poly")) # gls() has no arg "contrasts"
## fit <- gls(model=yield ~ 1 + geno + env + geno:env + env/block, data=dat,
system.time(
    fit <- gls(model=yield ~ 1 + geno + env + geno:env, data=dat,
               weights=varIdent(form=~1|env)))
fits_gls[["het"]] <- fit
fixefs[["het"]] <- coef(fit)
errVars[["het"]] <- (c(1, exp(coef(fit$modelStruct$varStruct))) *
  fit$sigma)^2
scaled_residuals_het <- residuals(fit) / rep(errVars[["het"]], each=I * K)
y_hat_het <- fitted(fit)
in_diags_gls[["het"]] <- cbind(dat,
                               "scaled_residuals"=scaled_residuals_het,
                               "y_hat"=y_hat_het)
```

#### Model comparison

```{r}
AIC(fits_gls[["hom"]], fits_gls[["het"]])
BIC(fits_gls[["hom"]], fits_gls[["het"]])
```

```{r}
anova(fits_gls[["hom"]], fits_gls[["het"]])
```

#### Diagnostics

```{r, fig.width=10}
plots <- lapply(names(in_diags_gls), function(hyp){
  in_diag <- in_diags_gls[[hyp]]
  p <- ggplot(in_diag, aes(x=y_hat, y=scaled_residuals, color=env)) +
    labs(title=paste0("Error variance: ", hyp),
         x="fitted values", y="scaled residuals") +
    geom_hline(yintercept=0) +
    geom_hline(yintercept=c(-2,2), linetype="dashed") +
    geom_point(size=2)
  p
})
plots[[1]]
plots[[2]]
```

```{r, fig.width=10}
sortEnvsPerVar <- names(sort(do.call(c, list_var_j)))
plots <- lapply(names(in_diags_gls), function(hyp){
  in_diag <- in_diags_gls[[hyp]]
  in_diag$env <- factor(as.character(in_diag$env),
                        levels=sortEnvsPerVar)
  p <- ggplot(in_diag, aes(x=env, y=scaled_residuals)) +
    labs(title=paste0("Error variance: ", hyp),
         x="fitted values", y="scaled residuals") +
    geom_hline(yintercept=0) +
    geom_hline(yintercept=c(-2,2), linetype="dashed") +
    geom_violin(aes(fill=env, color=env), trim=FALSE) +
    geom_boxplot(width=0.1)
  p
})
plots[[1]]
plots[[2]]
```

#### Inference

##### Assuming homoscedasticity

```{r}
hyp <- "hom"
```

###### Intercept

```{r}
data.frame("true"=mu,
           "estim"=summary(fits_gls[[hyp]])$tTable["(Intercept)", "Value"],
           "se"=summary(fits_gls[[hyp]])$tTable["(Intercept)", "Std.Error"])
```

###### Environment main effects

Contrasts:
```{r}
idx <- grep("^env", rownames(summary(fits_gls[[hyp]])$tTable))
(tmp <- cbind("true"=theta,
              "estim"=summary(fits_gls[[hyp]])$tTable[idx, "Value"],
              "se"=summary(fits_gls[[hyp]])$tTable[idx, "Std.Error"]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (gls hom)", pch=19); abline(a=0, b=1, lty=2)
```

Estimated marginal means:
```{r}
estimEnvEffs <- as.data.frame(emmeans(fits_gls[[hyp]], "env"))$emmean
(tmp <- cbind("true"=trueEnvEffs,
              "estim"=estimEnvEffs))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (gls hom)", pch=19); abline(a=0, b=1, lty=2)
```

###### Genotype main effects

Estimated marginal means:
```{r}
estimGenoEffs <- as.data.frame(emmeans(fits_gls[[hyp]], "geno"))$emmean
(tmp <- cbind("true"=mu + alpha,
              "estim"=estimGenoEffs))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Genotype main effects (gls hom)", pch=19); abline(a=0, b=1, lty=2)
```

###### Error variances

```{r}
(tmp <- cbind("true"=unlist(list_var_j),
              "estim"=errVars[[hyp]]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Error variances (gls hom)", pch=19); abline(a=0, b=1, lty=2)
```

##### PVE

```{r}
(tav_hom <- anova(fits_gls[[hyp]]))
```

##### Assuming heteroscedasticity

```{r}
hyp <- "het"
```

###### Intercept

```{r}
data.frame("true"=mu,
           "estim"=summary(fits_gls[[hyp]])$tTable["(Intercept)", "Value"],
           "se"=summary(fits_gls[[hyp]])$tTable["(Intercept)", "Std.Error"])
```

###### Environment main effects

Contrasts:
```{r}
idx <- grep("^env", rownames(summary(fits_gls[[hyp]])$tTable))
(tmp <- cbind("true"=theta,
              "estim"=summary(fits_gls[[hyp]])$tTable[idx, "Value"],
              "se"=summary(fits_gls[[hyp]])$tTable[idx, "Std.Error"]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (gls het)", pch=19); abline(a=0, b=1, lty=2)
```

Estimated marginal means:
```{r}
estimEnvEffs <- as.data.frame(emmeans(fits_gls[[hyp]], "env"))$emmean
(tmp <- cbind("true"=trueEnvEffs,
              "estim"=estimEnvEffs))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (gls het)", pch=19); abline(a=0, b=1, lty=2)
```

###### Genotype main effects

Estimated marginal means:
```{r}
estimGenoEffs <- as.data.frame(emmeans(fits_gls[[hyp]], "geno"))$emmean
(tmp <- cbind("true"=mu + alpha,
              "estim"=estimGenoEffs))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Genotype main effects (gls het)", pch=19); abline(a=0, b=1, lty=2)
```

###### Error variances

```{r}
(tmp <- cbind("true"=unlist(list_var_j),
              "estim"=errVars[[hyp]]))
cor(tmp[,"true"], tmp[,"estim"])
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Error variances (gls het)", pch=19); abline(a=0, b=1, lty=2)
```

### `MMEst` from `MM4LMM`

The `MMEst` from the `MM4LMM` package performs inference via ReML, and can handle both random variables and heteroscedasticity.

#### Fit

Output variables:
```{r}
fits_mm <- list()
blups_mm <- list()
in_diags_mm <- list()
```

Common design matrix of fixed effects:
```{r}
## X_C <- model.matrix(~ 1 + env + env/block, data=dat)
X_C <- model.matrix(~ 1 + env, data=dat,
                    contrasts.arg=list(env="contr.sum"))
dim(X_C)
head(X_C)
tail(X_C)
image(t(X_C)[,nrow(X_C):1], main="X_C", col=c("grey","white","black"),
      xaxt="n", yaxt="n")
```

##### Assuming homoscedasticity

Make design matrices:
```{r, fig.width=14}
hyp <- "hom"
lZ_hom <- list("geno"=model.matrix(~ 0 + geno, data=dat),
               "gxe"=model.matrix(~ 0 + gxe, data=dat),
               "err"=diag(nrow(dat)))
sapply(lZ_hom, dim)
par(mfrow=c(1,3))
for(k in seq_along(lZ_hom))
  image(t(lZ_hom[[k]])[,nrow(lZ_hom[[k]]):1],
        main=paste0("Z_", names(lZ_hom)[k], " (", nrow(lZ_hom[[k]]), " x ",
                    ncol(lZ_hom[[k]]), ")"),
        col=c("white","black"), xaxt="n", yaxt="n")
```

Make var-cov matrices:
```{r, fig.width=14}
lV_hom <- list("geno"=diag(nlevels(dat$geno)),
               "gxe"=diag(nlevels(dat$gxe)),
               "err"=diag(nrow(dat)))
stopifnot(all(names(lV_hom) == names(lZ_hom)))
par(mfrow=c(1,3))
for(k in seq_along(lV_hom))
  image(t(lV_hom[[k]])[,nrow(lV_hom[[k]]):1],
        main=paste0("V_", names(lV_hom)[k], " (", nrow(lV_hom[[k]]), " x ",
                    ncol(lV_hom[[k]]), ")"),
        col=c("white","black"), xaxt="n", yaxt="n")
```

Fit:
```{r}
system.time(
    fits_mm[[hyp]] <- MMEst(Y=dat$y, Cofactor=X_C, VarList=lV_hom, ZList=lZ_hom,
                            Method="Reml"))
blups_mm[[hyp]] <- MMBlup(Y=dat$y, Cofactor=X_C, VarList=lV_hom, ZList=lZ_hom,
                          ResMM=fits_mm[[hyp]])
y_hat_hom <- X_C %*% fits_mm[[hyp]]$NullModel$Beta +
  cbind(lZ_hom$geno, lZ_hom$gxe) %*% c(blups_mm[[hyp]]$geno, blups_mm[[hyp]]$gxe)
in_diags_mm[[hyp]] <- cbind(dat,
                            "y_hat"=y_hat_hom,
                            "residuals"=dat$y - y_hat_hom,
                            "scaled_residuals"=NA)
```

##### Assuming heteroscedasticity

Make design matrices:
```{r, fig.width=15}
hyp <- "het"
mkZerr <- function(env, dat){
  isEnv <- dat$env == env
  nbReps <- sum(isEnv)
  out <- matrix(0, nrow(dat), ncol=nbReps)
  out[isEnv] <- diag(nbReps)
  return(out)
}
lZerr <- lapply(levels(dat$env), mkZerr, dat)
names(lZerr) <- paste0("err", levels(dat$env))
lZ_het <- append(list("geno"=model.matrix(~ 0 + geno, data=dat),
                      "gxe"=model.matrix(~ 0 + gxe, data=dat)),
                 lZerr)
par(mfrow=c(2,5))
for(k in seq_along(lZ_het))
  image(t(lZ_het[[k]])[,nrow(lZ_het[[k]]):1],
        main=paste0("Z_", names(lZ_het)[k], " (", nrow(lZ_het[[k]]), " x ",
                    ncol(lZ_het[[k]]), ")"),
        col=c("white","black"), xaxt="n", yaxt="n")
```

Make var-cov matrices:
```{r}
lVerr <- lapply(table(dat$env), diag)
names(lVerr) <- paste0("err", names(lVerr))
lV_het <- append(list("geno"=diag(nlevels(dat$geno)),
                      "gxe"=diag(nlevels(dat$gxe))),
                 lVerr)
stopifnot(all(names(lV_het) == names(lZ_het)))
```

Fit:
```{r}
system.time(
    fits_mm[[hyp]] <- MMEst(Y=dat$y, Cofactor=X_C, VarList=lV_het, ZList=lZ_het,
                            Method="Reml"))
blups_mm[[hyp]] <- MMBlup(Y=dat$y, Cofactor=X_C, VarList=lV_het, ZList=lZ_het,
                          ResMM=fits_mm[[hyp]])
y_hat_het <- X_C %*% fits_mm[[hyp]]$NullModel$Beta +
  cbind(lZ_het$geno, lZ_het$gxe) %*% c(blups_mm[[hyp]]$geno, blups_mm[[hyp]]$gxe)
in_diags_mm[[hyp]] <- cbind(dat,
                            "y_hat"=y_hat_het,
                            "residuals"=dat$y - y_hat_het,
                            "scaled_residuals"=NA)
```

#### Model comparison

Likelihood ratio test statistic:
\begin{align}
LRT &= -2 \, \log \left( \frac{sup_{\theta \in \Theta_0} \mathcal{L}(\theta)}{sup_{\theta \in \Theta_1} \mathcal{L}(\theta)} \right) \\
    &= 2 \, ( l(\hat{\theta}_{H_1}) - l(\hat{\theta}_{H_0}) )
\end{align}

Asymptotic distribution for testing that *one variance* equals zero in a linear mixed model with one single random effect (case 6 of Self and Liang, 1987): $LRT \rightarrow \frac{1}{2} \chi^2(0) + \frac{1}{2} \chi^2(1)$

```{r}
(loglik_hom <- fits_mm[["hom"]]$NullModel$LogLik)
(loglik_het <- fits_mm[["het"]]$NullModel$LogLik)
(lrt <- 2 * (loglik_het - loglik_hom))
```

A simple Chi2 test should be conservative:
```{r}
h1 <- hist(rchisq(n=1000, df=1), breaks="FD", plot=F)
h2 <- hist(0.5 * rchisq(n=1000, df=0) + 0.5 * rchisq(n=1000, df=1),
           breaks="FD", plot=F)
plot(h1, col=rgb(0,0,1,1/4), xlim=c(0,12), main="Chi2 distributions",
     xlab="support", las=1)
plot(h2, col=rgb(1,0,0,1/4), xlim=c(0,12), add=T)
legend("topright", legend=c("Chi2(1)", "\n0.5 Chi2(0)\n+ 0.5 Chi2(1)"), bty="n",
       fill=c(rgb(0,0,1,1/4), rgb(1,0,0,1/4)))
if(lrt <= 12){
  abline(v=lrt, lwd=2)
  legend("right", legend="LRT", lty=1, lwd=2, bty="n")
}
(pval <- pchisq(lrt, df=1, lower.tail=FALSE))
(pval <- 0.5 * pchisq(lrt, df=0, lower.tail=FALSE) +
   0.5 * pchisq(lrt, df=1, lower.tail=FALSE))
```

Attention, ici on a 7 contraintes d'égalité de var, mais on va considérer un test avec 7 nullités.
Eventuellement simuler sous homoscédasticité, faire le test, vérif que p val ~ unif (mais attention, on ne sait pas exactement à quoi à ressemble la distrib sous H0: une unif + éventuellement une Dirac; ça peut notament poser pbm pour control tests multiples...).

```{r}
##' From the TcGSA package available at Bioconductor.
##' @param quant value of the test statistic
##' @param s number of fixed effects to test
##' @param q number of random effects to test
pchisqmix <- function(quant, s, q, lower.tail=TRUE){
  if(q > 0){
    mixprobs <- numeric(q + 1)
    for(k in s:(q + s)){
      mixprobs[k - s + 1] <- choose(q, k - s) * 2^(-q)
    }
    mix <- mixprobs
  } else{
    mix = 1
  }
  res <- numeric(length(mix))
  for(k in (s:(q + s))){
    res[k - s + 1] <- mix[k - s + 1] * stats::pchisq(quant, df = k,
                                                     lower.tail = lower.tail)
  }
  return(sum(res) / sum(mix))
}

pchisqmix(lrt, 0, 7, FALSE)
```

TODO: Conditionnal AIC (Liang et al, 2008)
```{r}
## (AICc <- )
```

TODO: Hybrid BIC (Delattre et al, 2014) but used to select fixed effects
```{r}
## (BIChyb_hom <- -2 * loglik_hom +
##    length(ncol(X_C)) * log(nrow(dat)) +
##    length(lZ_hom) * log())
```

#### Diagnostics

TODO

#### Inference

##### Assuming homoscedasticity

```{r}
hyp <- "hom"
```

###### Intercept

```{r}
data.frame("true"=mu,
           "estim"=fits_mm[[hyp]]$NullModel$Beta[1],
           "se"=sqrt(diag(fits_mm[[hyp]]$NullModel$VarBeta)[1]))
```

###### Environment main effects

Contrasts:
```{r}
(tmp <- cbind("true"=theta,
              "estim"=fits_mm[[hyp]]$NullModel$Beta[2:J],
              "se"=sqrt(diag(fits_mm[[hyp]]$NullModel$VarBeta)[2:J])))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (MMEst hom)", pch=19); abline(a=0, b=1, lty=2)
```

###### Genotype main effects

```{r}
(tmp <- cbind("true"=alpha,
              "estim"=blups_mm[[hyp]]$geno[,1]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Genotype main effects (MMEst hom)", pch=19); abline(a=0, b=1, lty=2)
```

###### Error variances

```{r}
(tmp <- cbind("true"=unlist(list_var_j),
              "estim"=rep(fits_mm[[hyp]]$NullModel$Sigma2["err"], J)))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Error variances (MMEst hom)", pch=19); abline(a=0, b=1, lty=2)
```

###### Genetic variances

```{r}
c(var_alpha, var_delta)
fits_mm[[hyp]]$NullModel$Sigma2[1:2]
```

##### Assuming heteroscedasticity

```{r}
hyp <- "het"
```

###### Intercept

```{r}
data.frame("true"=mu,
           "estim"=fits_mm[[hyp]]$NullModel$Beta[1],
           "se"=diag(fits_mm[[hyp]]$NullModel$VarBeta)[1])
```

###### Environment main effects

Contrasts:
```{r}
(tmp <- cbind("true"=theta,
              "estim"=fits_mm[[hyp]]$NullModel$Beta[2:J],
              "se"=diag(fits_mm[[hyp]]$NullModel$VarBeta)[2:J]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Environment main effects (MMEst het)", pch=19); abline(a=0, b=1, lty=2)
```

###### Genotype main effects

```{r}
(tmp <- cbind("true"=alpha,
              "estim"=blups_mm[[hyp]]$geno[,1]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Genotype main effects (MMEst het)", pch=19); abline(a=0, b=1, lty=2)
```

###### Error variances

```{r}
idx <- grep("^err", names(fits_mm[[hyp]]$NullModel$Sigma2))
(tmp <- cbind("true"=unlist(list_var_j),
              "estim"=fits_mm[[hyp]]$NullModel$Sigma2[idx]))
plot(tmp[,"estim"], tmp[,"true"], las=1, asp=1, xlab="estimated value", ylab="true value", main="Error variances (MMEst het)", pch=19); abline(a=0, b=1, lty=2)
```

###### Genetic variances

```{r}
c(var_alpha, var_delta)
fits_mm[[hyp]]$NullModel$Sigma2[1:2]
```


# Acknowledgments

T. Mary-Huard


# Appendix

```{r info}
t1 <- proc.time()
t1 - t0
print(sessionInfo(), locale=FALSE)
```
