---
title: "Parentage analysis"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
##`
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
opts_knit$set(progress=TRUE, verbose=TRUE)
```


# Overview

This document aims at presenting how to perform parentage analysis.

This document requires external packages:
```{r load_pkg}
suppressPackageStartupMessages(library(scrm)) # on the CRAN
suppressPackageStartupMessages(library(rmetasim)) # on the CRAN
suppressPackageStartupMessages(library(kinship2)) # on the CRAN
suppressPackageStartupMessages(library(related)) # github.com/timothyfrasier/related
suppressPackageStartupMessages(library(rcolony)) # github.com/jonesor/rcolony
suppressPackageStartupMessages(library(rutilstimflutre)) # on GitHub
stopifnot(compareVersion("0.159.1",
                         as.character(packageVersion("rutilstimflutre")))
          != 1)
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```



# Simulate data

More details [here](https://cran.r-project.org/package=rmetasim).

* species: perennial plant

* mating system: [monoecious](https://en.wikipedia.org/wiki/Monoecious) (a tree has female and male flowers) with no [geitonogamy](https://en.wikipedia.org/wiki/Geitonogamy), i.e. "every individual can be a mother and a father, but none can be both the mother and father of the same offspring" (from A. Strand)

* one population/habitat: with a given carrying capacity

    * initially, only juveniles (half females and half males)

* life cycle: four-stage (juvenile and adult for both female and male)

    * male and female are distinguished because `kinship2` requires sex

* markers: several SSRs (each with more than 2 alleles)

* time interval: interpreted as being a single year

## Setup the `rmetasim` object

Set the seed:
```{r set_seed}
set.seed(1859)
```

Set the input parameters:
```{r}
nb.habitats <- 1
nb.stages <- 4
nb.generations <- 1000
nb.locus <- 5
nb.alleles <- 7
nb.fem.juv <- 2
nb.mal.juv <- 2
carrying.cap <- 2000
```

### Set up the object

```{r}
land <- landscape.new.empty()
land <- landscape.new.intparam(land, h=nb.habitats, s=nb.stages,
                               totgen=nb.generations)
## mp=1: every ovule of a given mother could be fertilized by a different father
land <- landscape.new.switchparam(land, mp=1)
land <- landscape.new.floatparam(land, s=0)
```

### Set up the local demography

```{r}
## stages: female juvenile (1) and adult (2), male juvenile (3) and adult (4)
## Survival during a given time interval [t, t']: same for female and male
##   Pr(juvenile at t' | juvenile at t) = 0
##   Pr(adult at t' | juvenile at t) = 0.95
##   Pr(dead at t' | juvenile at t) = 0.05
##   Pr(juvenile at t' | adult at t) = 0
##   Pr(adult at t' | adult at t) = 0.7
##   Pr(dead at t' | adult at t) = 0.3
loc.survival <- matrix(c(0, 0.95, 0, 0,     # column 1: from juvenile female to ...
                         0, 0.70, 0, 0,     # column 2: from adult female to ...
                         0, 0,    0, 0.95,  # column 3: from juvenile male to ...
                         0, 0,    0, 0.70), # column 4: from adult male to ...
                       nrow=nb.stages, ncol=nb.stages)
## Female reproduction during a given time interval [t, t']:
##   E[#offsprings | juvenile x juvenile] = 0
##   E[#offsprings | juvenile x adult] = 0
##   E[#offsprings | adult x juvenile] = 0
##   E[#offsprings female | adult x adult] = 2
##   E[#offsprings male | adult x adult] = 2
loc.reprod.fem <- matrix(c(0, 0, 0, 0,
                           2, 0, 2, 0,
                           0, 0, 0, 0,
                           0, 0, 0, 0),
                         nrow=nb.stages, ncol=nb.stages)
## Male reproduction during a given time interval [t, t']:
##   Pr(cross with adult female | adult male) = 1
loc.reprod.mal <- matrix(c(0, 0, 0, 0,
                           0, 0, 0, 0,
                           0, 0, 0, 0,
                           0, 1, 0, 0),
                         nrow=nb.stages, ncol=nb.stages)
land <- landscape.new.local.demo(land, S=loc.survival,
                                 R=loc.reprod.fem,
                                 M=loc.reprod.mal)
```

To be changed if there is more than one habitat:
```{r}
lmat <- matrix(0, nb.stages*nb.habitats, nb.stages*nb.habitats)
land <- landscape.new.epoch(land, S=lmat, R=lmat, M=lmat,
                            carry=c(carrying.cap))
```

### Set up the markers

```{r}
for(i in 1:nb.locus)
  land <- landscape.new.locus(land, type=1, ploidy=2,
                              mutationrate=0.005,
                              numalleles=nb.alleles,
                              frequencies=NULL)
```

### Set up the initial individuals

```{r}
nb.fem.adu <- 0
nb.mal.adu <- 0
land <- landscape.new.individuals(land, c(nb.fem.juv, nb.fem.adu,
                                          nb.mal.juv, nb.mal.adu))
ind.col.names <- c("stage", "notused", "birth",
                   "id", "matid", "patid",
                   paste0(rep(paste0("loc", 1:land$intparam$locusnum), each=2),
                          ".", 1:2))
colnames(land$individuals) <- ind.col.names
stage2sex <- setNames(c(2, 2, 1, 1), # for kinship2: female=2 1=male
                      c(1, 2, 3, 4)) # <- stages
```

## Extract the pedigree

```{r}
ped <- cbind(land$individuals[, c(4:6, 3)],
             land$individuals[, 1] %% land$intparam$stages)
colnames(ped) <- c("ind", "mother", "father", "gen", "sex")
ped[ped[,"sex"] %in% c(0,1), "sex"] <- 2 # female for kinship2
ped[ped[,"sex"] %in% c(2,3), "sex"] <- 1 # male for kinship2
ped[ped[,"mother"] == 0, "gen"] <- 0
ped[, c("mother","father")] <- NA
nrow(ped)
head(ped)
tail(ped)
table(ped[,"gen"])
table(ped[,"sex"])
```

## Extract the SSR genotypes

```{r}
genos <- land$individuals[, 7:ncol(land$individuals)]
colnames(genos) <- paste0(rep(paste0("loc", 1:land$intparam$locusnum), each=2),
                          ".", 1:2)
rownames(genos) <- land$individuals[, 4]
dim(genos)
genos[1:4, 1:min(ncol(genos),6)]
table(genos[,1:2])
```

## Useful functions

```{r}
handleRmetasimPedigree <- function(ped.prev, land.new=NULL, stage2sex=NULL,
                                   gens.tokeep=NULL, prop=1){
  stopifnot(is.matrix(ped.prev),
            ncol(ped.prev) == 5,
            ! is.null(colnames(ped.prev)),
            all(colnames(ped.prev) == c("ind","mother","father","gen","sex")),
            xor(is.null(land.new), is.null(gens.tokeep)),
            all(prop >= 0, prop <= 1))
  if(! is.null(land.new))
    stopifnot(rmetasim::is.landscape(land.new),
              ! is.null(stage2sex),
              length(stage2sex) == land$intparam$stages)
  if(! is.null(gens.tokeep))
    stopifnot(is.vector(gens.tokeep),
              is.numeric(gens.tokeep))

  ped.out <- ped.prev

  ## add new individuals to the previous pedigree
  if(! is.null(land.new) & is.null(gens.tokeep) &
     ! all(land.new$individuals[,4] %in% ped.out[,"ind"])){
    ped.new <- land.new$individuals[, c(4:6,3)]
    colnames(ped.new) <- colnames(ped.prev)[1:ncol(ped.new)]
    ## add stage
    ped.new <- cbind(ped.new,
                     stage=(land.new$individuals[,1] %% land.new$intparam$stages) + 1)
    ## convert stage to sex
    ped.new <- cbind(ped.new,
                     sex=ped.new[,"stage"])
    stage2idx <- lapply(names(stage2sex), function(stage){
      which(ped.new[,"sex"] == stage)
    })
    names(stage2idx) <- names(stage2sex)
    for(stage in names(stage2idx)){
      idx <- stage2idx[[stage]]
      if(length(idx) > 0)
        ped.new[idx, "sex"] <- stage2sex[stage]
    }
    ped.new <- ped.new[, - grep("stage", colnames(ped.new))]
    idx <- which(! ped.new[,"ind"] %in% ped.out[,"ind"])
    if(length(idx) > 0)
      ped.out <- rbind(ped.out,
                       ped.new[idx,])
  }

  ## remove duplicated individuals
  is.dup <- duplicated(ped.out[,"ind"])
  if(any(is.dup))
    ped.out <- ped.out[! is.dup,]

  ## extract specific generations, with their founders
  if(! is.null(gens.tokeep)){
    ped.sub <- ped.out[ped.out[,"gen"] %in% gens.tokeep,]
    if(prop < 1){ # subsampling
      idx <- sample.int(n=nrow(ped.sub),
                        size=round(prop * nrow(ped.sub)))
      ped.sub <- ped.sub[idx,]
    }
    par1 <- ped.sub[,"mother"]
    par1 <- par1[! is.na(par1)]
    par2 <- ped.sub[,"father"]
    par2 <- par2[! is.na(par2)]
    parents <- sort(unique(c(par1, par2)))
    is.abs <- ! parents %in% ped.sub[,"ind"]
    if(any(is.abs)){
      par.abs <- parents[is.abs]
      stopifnot(all(par.abs %in% ped.out[,"ind"]))
      ped.par.abs <- ped.out[ped.out[,"ind"] %in% par.abs,]
      ped.par.abs[,c("mother","father")] <- NA
      ped.out <- rbind(ped.par.abs, ped.sub)
    }
  }

  return(ped.out)
}

handleRmetasimGenos <- function(genos.prev, land=NULL){
  if(! is.null(land))
    stopifnot(rmetasim::is.landscape(land))

  genos.new <- land$individuals[, 7:ncol(land$individuals)]
  rownames(genos.new) <- land$individuals[,4]

  genos.out <- rbind(genos.prev, genos.new)
  genos.out <- genos.out[! duplicated(rownames(genos.out)),]
  return(genos.out)
}
```

## First iteration

The `landscape.simulate` function wraps other functions, and it may be better to call them directly because the time interval is incremented inside `landscape.survive`.

Run:
```{r}
land <- landscape.reproduce(land)
ped <- handleRmetasimPedigree(ped, land, stage2sex)
genos <- handleRmetasimGenos(genos, land)
stopifnot(nrow(genos) == nrow(ped))
stopifnot(all(ped[,"ind"] %in% as.numeric(rownames(genos))))
land <- landscape.survive(land)
land <- landscape.carry(land)
land <- landscape.advance(land)
```

Because initially there are only juveniles, no new individual appeared.
However, the initial juveniles became adults.

```{r}
nrow(land$individuals)
table(floor(land$individuals[,1] / land$intparam$stages)) # habitat
table(land$individuals[,1] %% land$intparam$stages) # stage
table(land$individuals[,3]) # generation of birth
```

## Second iteration

Run:
```{r}
land <- landscape.reproduce(land)
ped <- handleRmetasimPedigree(ped, land, stage2sex)
genos <- handleRmetasimGenos(genos, land)
stopifnot(nrow(genos) == nrow(ped))
stopifnot(all(ped[,"ind"] %in% as.numeric(rownames(genos))))
land <- landscape.survive(land)
land <- landscape.carry(land)
land <- landscape.advance(land)
```

```{r}
nrow(land$individuals)
table(floor(land$individuals[,1] / land$intparam$stages)) # habitat
table(land$individuals[,1] %% land$intparam$stages) # stage
table(land$individuals[,3]) # generation of birth

plotPedigree(ped[,"ind"], ped[,"mother"], ped[,"father"], ped[,"gen"])
nrow(ped)
table(ped[,"gen"])
table(ped[,"sex"])

nrow(genos)
stopifnot(nrow(genos) == nrow(ped))
stopifnot(all(ped[,"ind"] %in% as.numeric(rownames(genos))))
```

## More iterations

```{r}
nb.iters <- 50
for(i in 1:nb.iters){
  land <- landscape.reproduce(land)
  ped <- handleRmetasimPedigree(ped, land, stage2sex)
  genos <- handleRmetasimGenos(genos, land)
  stopifnot(nrow(genos) == nrow(ped))
  stopifnot(all(ped[,"ind"] %in% as.numeric(rownames(genos))))
  land <- landscape.survive(land)
  land <- landscape.carry(land)
  land <- landscape.advance(land)
}
```

```{r}
## plotPedigree(ped[,"ind"], ped[,"mother"], ped[,"father"], ped[,"gen"],
##              xmin=-2.5, xmax=2.5)
hist(ped[,"gen"], main="Number of individuals per generation",
     col="grey", border="white", xlab="generations", ylab="", las=1)
table(genos[,1:2])
table(genos[,3:4])
```

## Save

```{r}
p2f <- "pedigree_rmetasim.tsv"
write.table(ped, file=p2f, quote=FALSE, sep="\t", row.names=FALSE)
```

```{r}
p2f <- "genos_rmetasim.tsv"
write.table(genos, file=p2f, quote=FALSE, sep="\t", col.names=FALSE)
```



# Infer genetic relationships

Only keep the last three generations, and subsample:
```{r}
length(all.gens <- sort(unique(ped[,"gen"])))
(gens.tokeep <- all.gens[(length(all.gens)-2):length(all.gens)])
ped.tmp <- handleRmetasimPedigree(ped, land=NULL, stage2sex=NULL,
                                  gens.tokeep, 0.02)
nrow(ped.tmp)
table(ped.tmp[,"gen"])
table(ped.tmp[,"sex"])

plotPedigree(ped.tmp[,"ind"], ped.tmp[,"mother"], ped.tmp[,"father"], ped.tmp[,"gen"], verbose=1, xmin=-2, xmax=2)

genos.tmp <- genos[as.character(ped.tmp[,"ind"]),]
dim(genos.tmp)
```

## From pedigree

Compute the expected additive genetic relationships from the pedigree:
```{r}
ped.obj <- pedigree(id=ped.tmp[,"ind"], dadid=ped.tmp[,"father"],
                    momid=ped.tmp[,"mother"], sex=ped.tmp[,"sex"])
A.ped <- kinship(ped.obj) * 2
```

Plot:
```{r}
imageWithScale(A.ped, main="Expected additive genetic relationships from the pedigree")
```

Example of a mother-father-offspring trio:
```{r}
tail(ped.tmp)
(off <- as.character(ped.tmp[nrow(ped.tmp), "ind"]))
(mother <- as.character(ped.tmp[nrow(ped.tmp), "mother"]))
(father <- as.character(ped.tmp[nrow(ped.tmp), "father"]))
A.ped[c(mother, father, off), c(mother, father, off)]
```

TODO: Example of two full-siblings:

TODO: Example of two half-siblings:

## From SSRs

More details [here](https://github.com/timothyfrasier/related).

Format the data:
```{r}
p2f <- "genos_related.tsv"
tmp <- genos.tmp + 1
write.table(tmp, file=p2f, quote=FALSE, sep="\t", col.names=FALSE)
genodat <- readgenotypedata(p2f)
file.remove(p2f)
names(genodat)
dim(genodat$gdata)
genodat$nloci
genodat$nalleles
genodat$ninds
genodat$freqs
```

TODO: use allele frequencies from the whole data set

Compute the MLEs:
```{r}
p2f <- "parentage-analysis_coanc.RData"
if(! file.exists(p2f)){
  system.time(
      coanc <- coancestry(genotype.data=genodat$gdata,
                          error.rates=0,
                          dyadml=2, allow.inbreeding=TRUE))
  save(coanc, file=p2f)
  print(tools::md5sum(path.expand(p2f)))
} else{
  print(tools::md5sum(path.expand(p2f)))
  load(p2f)
}
A.ssr <- coancestry2relmat(x=coanc, estim.coancestry="dyadml",
                           estim.inbreeding="LR", rel.type="relationships")
```

Plot:
```{r}
imageWithScale(A.ssr, main="Additive genetic relationships from the SSRs")
```

```{r}
head(coanc$relatedness)
tail(coanc$relatedness)
i <- 1
(pair <- c(coanc$relatedness$ind1.id[i], coanc$relatedness$ind2.id[i]))
A.ped[pair, pair]
A.ssr[pair, pair]
```

Example of a mother-father-offspring trio:
```{r}
tail(ped.tmp)
(off <- as.character(ped.tmp[nrow(ped.tmp), "ind"]))
(mother <- as.character(ped.tmp[nrow(ped.tmp), "mother"]))
(father <- as.character(ped.tmp[nrow(ped.tmp), "father"]))
A.ped[c(mother, father, off), c(mother, father, off)]
```

TODO: Example of two full-siblings:

TODO: Example of two half-siblings:

## Comparison

```{r}
A.ped[1:3,1:3]
A.ssr[1:3,1:3]
```

```{r}
plot(diag(A.ped), diag(A.ssr))
plot(A.ped[upper.tri(A.ped)], A.ssr[upper.tri(A.ssr)])
```



# Infer parentages

TODO: see `rcolony`



# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
