---
title: "Analysis of a biparental design for perennial fruit plants"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: TRUE
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
opts_knit$set(progress=TRUE, verbose=TRUE)
```


# Overview

Many field experiments in perennial fruit plants are set up so that two parents and their offsprings are planted at a single location and traits are phenotyped for several years.
This document aims at presenting how to perform the analysis of such experiments when production is negatively correlated from one year to the next.

This document requires external packages:
```{r load_pkg}
suppressPackageStartupMessages(library(parallel)) # on the CRAN
nb.cores <- max(1, detectCores() - 1)
suppressPackageStartupMessages(library(scrm)) # on the CRAN
suppressPackageStartupMessages(library(MASS)) # on the CRAN
suppressPackageStartupMessages(library(beanplot)) # on the CRAN
suppressPackageStartupMessages(library(lattice)) # on the CRAN
suppressPackageStartupMessages(library(MuMIn)) # on the CRAN
suppressPackageStartupMessages(library(boot)) # on the CRAN
suppressPackageStartupMessages(library(lme4)) # on the CRAN
suppressPackageStartupMessages(library(nlme)) # on the CRAN
suppressPackageStartupMessages(library(lmerTest)) # on the CRAN
suppressPackageStartupMessages(library(rutilstimflutre)) # on GitHub
stopifnot(compareVersion("0.166.0",
                         as.character(packageVersion("rutilstimflutre")))
          != 1)
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Statistical model

## Notations

* $G$: number of genotypes

* $R$: number of replicates per genotype

* $T$: number of years

* $N$: number of phenotypes ($N = G \times R \times T$)

* $\boldsymbol{y}$: $N$-vector of phenotypes

* $X$: $N \times T$ design matrix relating phenotypes to years

* $\boldsymbol{\beta}$: $T$-vector of year effects (will be modeled as "fixed")

* $Z$: $N \times G$ design matrix relating phenotypes to genotypes

* $\boldsymbol{u}$: $G$-vector of genotypic values (will be modeled as "random")

* $\sigma_g^2$: variance component of the genotypic values

* $\boldsymbol{\epsilon}$: $N$-vector of errors

* $\sigma^2$: variance component of the errors

* $\rho$: absolute value of the correlation between errors from one year to the next for a given replicate

## Likelihood

\[
\boldsymbol{y} = X \boldsymbol{\beta} + Z \boldsymbol{u} + \boldsymbol{\epsilon}
\]

with:

* $\boldsymbol{u} \sim \mathcal{N}(0, \sigma_u^2 \, \text{Id})$;

* $\boldsymbol{\epsilon} \sim \mathcal{N}(0, \Sigma)$.

For each replicate, $T$ errors are drawn at once from a multivariate Normal such that its correlation matrix is Toeplitz:

\[
\begin{pmatrix}
1      & -\rho  & \rho   & -\rho  & \ldots \\
-\rho  & 1      & -\rho & \rho   & \ldots \\
\vdots & \ldots & 1     & \ldots & \vdots
\end{pmatrix}
\]


# Data simulation

## Inputs

```{r}
mean.pheno <- 100
min.pheno <- 20
prop.var.fix <- 0.1
prop.var.geno <- 0.1
(var.pheno <- ((mean.pheno - min.pheno) / 3.5)^2)
(var.fix <- prop.var.fix * var.pheno)
(var.geno <- prop.var.geno * var.pheno)
(cv.geno <- sqrt(var.geno) / mean.pheno)
(var.error <- var.pheno - (var.fix + var.geno))
var.error / var.pheno
(H2.ind <- var.geno / var.pheno)
H2.means <- 0.8
nb.years <- 10
(nb.reps <- round(((H2.means/nb.years) * var.error) /
                  (var.geno - H2.means * var.geno)))
G <- 100
R <- nb.reps
T <- nb.years
rho <- 0.5
```

## Genotypes

```{r set_seed_genos}
set.seed(1859)
```

### Base population

Simulate haplotypes of a base population via the sequential coalescent with recombination, and encode the corresponding bi-allelic SNP genotypes additively as allele doses in ${0,1,2}$:
```{r simul_genos}
nb.genos <- 500 # base population from which parents will be sampled later on
nb.chrs <- 10
chr.len.phy <- 10^5
Ne <- 10^4
mu <- 10^(-8)
c.rec <- 10^(-8)
genomes <- simulCoalescent(nb.inds=nb.genos, nb.reps=nb.chrs,
                           pop.mut.rate=4 * Ne * mu * chr.len.phy,
                           pop.recomb.rate=4 * Ne * c.rec * chr.len.phy,
                           chrom.len=chr.len.phy,
                           get.alleles=TRUE)
(P <- ncol(genomes$genos))
afs.pop <- estimSnpAf(X=genomes$genos)
mafs.pop <- estimSnpMaf(afs=afs.pop)
A.vr.pop <- estimGenRel(X=genomes$genos, afs=afs.pop, method="vanraden1")
```

Look at some visual checks:
```{r check_simul_genos}
plotHistAllelFreq(afs=afs.pop,
                  main=paste0("Allele frequencies of ", P, " SNPs"))
plotHistMinAllelFreq(mafs=mafs.pop,
                     main=paste0("Minor allele frequencies of ", P, " SNPs"))
summary(diag(A.vr.pop)) # under HWE, average should be 1
hist(diag(A.vr.pop), col="grey", border="white")
summary(A.vr.pop[upper.tri(A.vr.pop)]) # under HWE, average should be 0
hist(A.vr.pop[upper.tri(A.vr.pop)], col="grey", border="white")
```

### Controlled crosses

Choose two individuals as parents:
```{r choose_parents}
(names.parents <- sample(x=rownames(genomes$genos), size=2))
A.vr.pop[names.parents, names.parents]
genos.parents <- genomes$genos[names.parents,]
haplos.parents <- getHaplosInds(haplos=genomes$haplos,
                                ind.names=names.parents)
```

Cross them several times to make offsprings:
```{r crosse_to_make_offsprings}
nb.offs <- G - 2
names.offs <- paste0("off-",
                     sprintf(fmt=paste0("%0", floor(log10(nb.offs))+1, "i"),
                             1:nb.offs))
names.genos <- c(names.parents, names.offs)
head(crosses.off <- data.frame(parent1=rep(names.parents[1], nb.offs),
                               parent2=rep(names.parents[2], nb.offs),
                               child=names.offs,
                               stringsAsFactors=FALSE))
loc.crovers.off <- drawLocCrossovers(crosses=crosses.off,
                                     nb.snps=sapply(haplos.parents, ncol),
                                     simplistic=FALSE,
                                     verbose=1)
haplos.offs <- makeCrosses(haplos=haplos.parents, crosses=crosses.off,
                           loc.crossovers=loc.crovers.off,
                           howto.start.haplo=0)
genos.offs <- segSites2allDoses(seg.sites=haplos.offs,
                                ind.ids=getIndNamesFromHaplos(haplos.offs),
                                snp.ids=rownames(genomes$snp.coords))
dim(genos.doses <- rbind(genos.parents, genos.offs))
genos.classes <- genoDoses2genoClasses(X=genos.doses,
                                       alleles=genomes$alleles)
genos.jm <- genoClasses2JoinMap(x=genos.classes)
genos.jm[1:3, 1:14]
tests.seg <- filterSegreg(genos.jm[,-c(1:8)], return.counts=TRUE)
```

Plot pedigree:
```{r plot_ped}
ped <- data.frame(ind=c(names.parents,
                        crosses.off$child),
                  mother=c(rep(NA, length(names.parents)),
                           crosses.off$parent1),
                  father=c(rep(NA, length(names.parents)),
                           crosses.off$parent2),
                  gen=c(rep(0, length(names.parents)),
                        rep(1, nrow(crosses.off))),
                  stringsAsFactors=FALSE)
ped.tmp <- rbind(ped[1:5,],
                 c(ind="off-...", ped[5, -1]),
                 c(ind="off-....", ped[5, -1]),
                 ped[nrow(ped),])
plotPedigree(inds=ped.tmp$ind, mothers=ped.tmp$mother, fathers=ped.tmp$father,
             generations=ped.tmp$gen, main="Pedigree of the controlled cross")
```

Check additive genetic relationships:
```{r add_gen_rel}
A.vr.cross <- estimGenRel(X=genos.doses, afs=afs.pop, method="vanraden1")
## imageWithScale(A.vr.cross, main="Additive genetic relationships of crosses")
## imageWithScale(A.vr.cross[1:10, 1:10],
##                main="Additive genetic relationships of crosses (subset)",
##                idx.rownames=1:10, idx.colnames=1:10)
summary(diag(A.vr.cross))
summary(A.vr.cross[upper.tri(A.vr.cross)])
summary(A.vr.cross[names.parents[1], grep("off", colnames(A.vr.cross))])
summary(A.vr.cross[names.parents[2], grep("off", colnames(A.vr.cross))])
```

Under HWE in a single population, the additive genetic relationships between all parent-child pairs should be centered around 0.5, corresponding to a coancestry coefficient of 1/4.


## Phenotypes

```{r set_seed_pheno}
set.seed(1859)
```

### Data structure

```{r}
years <- 2001:(2001 + T - 1)
dat.annual <- data.frame(geno=rep(names.genos, each=R),
                         rep=rep(1:R, G))
dat.annual$geno.rep <- paste0(dat.annual$geno, "_", dat.annual$rep)
dat <- dat.annual
for(t in 2:T)
  dat <- rbind(dat, dat.annual)
dat$year <- rep(years, each=G * R)
dat$t <- dat$year - min(dat$year)
dat$year <- as.factor(dat$year)
dat$rep <- as.factor(dat$rep)
dat$geno.rep <- as.factor(dat$geno.rep)
```

```{r}
str(dat)
dim(dat)
head(dat)
```

### Year effects

```{r}
Z.year <- model.matrix(~ 1 + year, data=dat)
dim(Z.year)
year.effs <- c(mean.pheno,
               rnorm(n=T-1, mean=0, sd=sqrt(var.fix)))
year.effs
summary(year.effs[-1])
var(year.effs[-1])
```

### Genotype effects

```{r}
Z.geno <- model.matrix(~ -1 + geno, data=dat)
dim(Z.geno)
geno.vals <- MASS::mvrnorm(n=1, mu=rep(0, G), Sigma=var.geno * diag(G))
summary(geno.vals)
var(geno.vals)
```

### Errors

```{r}
mat.cor.error <- toeplitz(c(1, rho * rep(c(-1,1), length.out=T-1)))
mat.cor.error[1,]
vcov.error <- cor2cov(x=mat.cor.error, sd=sqrt(var.error))
kappa(vcov.error)
eigen(vcov.error)$values
tmp <- mvrnorm(n=G*R, mu=rep(0, T), Sigma=vcov.error)
dim(tmp)
errors <- c(tmp)
```

### Production

```{r}
y <- Z.year %*% year.effs +
  Z.geno %*% geno.vals +
  errors
stopifnot(all(y >= 0))
```

Option to add missing data:
```{r}
y.NA <- y
prop.NA <- 0 # 0.05
idx <- sample.int(n=length(y), size=floor(prop.NA * length(y)))
if(length(idx) > 0)
  y.NA[idx] <- NA
```

```{r}
dat$prod.noNA <- y
dat$prod <- y.NA
rownames(dat) <- NULL
```


# Data exploration

```{r}
trait <- "prod"
```

## Tables

```{r smry_stats}
tmp <- tapply(dat[[trait]], dat$year, prettyPrintBetterSummary)
cor(dat[[trait]][dat$year == "2001"],
    dat[[trait]][dat$year == "2002"])
cor(dat[[trait]][dat$year == "2002"],
    dat[[trait]][dat$year == "2003"])
cor(dat[[trait]][dat$year == "2001"],
    dat[[trait]][dat$year == "2003"])
```

## Plots

```{r plots_data}
hist(dat[[trait]], breaks="FD", col="grey", border="white", las=1,
     main="Trait")
```

```{r}
beanplot(formula(paste0(trait , " ~ year")),
         data=dat, ylim=NULL,
         log="", ll=0.02, side="no", las=1,
         border=NA, col="black",
         main=trait)
```

```{r}
lower.threshold <- 0
upper.threshold <- +Inf
x <- paste0(trait, ".outlier")
dat[[x]] <- FALSE
is.outlier <- dat[[trait]] < lower.threshold |
  dat[[trait]] > upper.threshold
sum(is.outlier, na.rm=TRUE)
dat[[x]][is.outlier] <- TRUE
```

```{r}
set.seed(1234)
(some.genos <- sample(names.genos, 3))
tmp <- dat[dat$geno %in% some.genos &
          dat$rep == "1",
          c("geno", "year", "prod")]
tmp$year <- as.numeric(levels(tmp$year))[tmp$year]
plot(x=tmp$year[tmp$geno == some.genos[1]],
     y=tmp$prod[tmp$geno == some.genos[1]],
     ylim=range(tmp$prod),
     type="b", pch=1, col=1,
     main="Example of genotypes with alternate production",
     xlab="years", las=1,
     ylab="production")
for(i in 2:length(some.genos))
  points(x=tmp$year[tmp$geno == some.genos[i]],
         y=tmp$prod[tmp$geno == some.genos[i]],
         type="b", pch=1, col=i)
legend("bottomright", legend=some.genos, col=1:length(some.genos),
       lty=1, pch=1, bty="n")
```


# Data preparation

```{r}
transfo <- "id"
stopifnot(transfo %in% c("id", "log", "sqrt"))
```

```{r}
if(any(dat[[paste0(trait, ".outlier")]])){
  dat <- droplevels(dat[! dat[[paste0(trait, ".outlier")]],])
  print(str(dat))
}
```

```{r}
if(transfo == "id"){
  response <- trait
} else if(transfo == "log"){
  response <- paste0("log(", trait, ")")
  dat[[response]] <- log(dat[[trait]])
} else if(transfo == "sqrt"){
  response <- paste0("sqrt(", trait, ")")
  dat[[response]] <- sqrt(dat[[trait]])
}
```

```{r}
dat.noNA <- droplevels(na.exclude(dat))
str(dat.noNA)
```


# Statistical model without inter-annual correlation

## Model fitting, comparison and selection

```{r}
form <- paste0(response, " ~ 1",
               " + year",
               " + (1|geno)",
               " + (1|geno.rep)")
globmod.ml <- lmer(as.formula(form),
                   data=dat.noNA,
                   na.action=na.fail,
                   REML=FALSE)
system.time(
    allmod.sel <- suppressMessages(dredge(globmod.ml)))
bestmod.ml <- get.models(allmod.sel, subset=1)[[1]]
formula(bestmod.ml)
bestmod.reml.lme4 <- lmer(formula=formula(bestmod.ml),
                          data=dat.noNA,
                          na.action=na.fail,
                          REML=TRUE)
bestmod.reml <- bestmod.reml.lme4
bestmod.reml.test <- lmerTest::lmer(formula=formula(bestmod.reml),
                                    data=dat.noNA,
                                    na.action=na.fail, REML=TRUE)
allmod.sel <- lmerTest::step(bestmod.reml.test, reduce.fixed=FALSE)
bestmod.reml.test.final <- lmerTest::get_model(allmod.sel)
formula(bestmod.reml.test.final)
bestmod.reml <- bestmod.reml.test.final
bestmod.ml <- lmerTest::lmer(formula=formula(bestmod.reml),
                             data=dat.noNA,
                             na.action=na.fail, REML=FALSE)
```

Parse the formula of the best model:
```{r}
(best.form <- Reduce(paste, deparse(formula(bestmod.ml))))
(all.preds <- trimws(strsplit(best.form, "\\~|\\+")[[1]])[-1])
(all.preds.fix <- all.preds[grep("\\(|\\)", all.preds, invert=TRUE)])
(all.preds.ran <- all.preds[grep("\\(|\\)", all.preds, invert=FALSE)])
```

## Assumption checking

### Residual preparation

```{r prep_res}
fit.all <- cbind(dat,
                 response=dat[[response]],
                 cond.res=NA,
                 scl.cond.res=NA,
                 fitted=NA)
idx.NA <- attr(dat.noNA, "na.action")
if(length(idx.NA) > 0){
  fit.all$cond.res[- idx.NA] <- residuals(bestmod.reml)
  fit.all$scl.cond.res[- idx.NA] <- residuals(bestmod.reml) /
    sigma(bestmod.reml)
  fit.all$fitted[- idx.NA] <- fitted(bestmod.reml)
} else{
  fit.all$cond.res <- residuals(bestmod.reml)
  fit.all$scl.cond.res <- residuals(bestmod.reml) /
    sigma(bestmod.reml)
  fit.all$fitted <- fitted(bestmod.reml)
}
geno.blups <- ranef(bestmod.reml, condVar=TRUE, drop=TRUE)$geno
geno.var.blups <- setNames(attr(geno.blups, "postVar"),
                           names(geno.blups))
```

### Error homoscedasticity

```{r check_error_homoscedasticity}
x.lim <- max(abs(fit.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.all$scl.cond.res, y=fit.all$fitted, las=1,
     xlim=c(-x.lim, x.lim),
     xlab="scaled conditional residuals",
     ylab="fitted responses",
     main=response)
abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
```

```{r check_error_homoscedasticity_year_block}
lattice::xyplot(fitted ~ scl.cond.res | year, #groups=block,
                data=fit.all,
                xlab="scaled conditional residuals",
                ylab="fitted responses",
                main=response,
                auto.key=list(space="right"),
                panel=function(x,y,...){
                  panel.abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
                  panel.xyplot(x,y,...)
                })
```

### Error normality

```{r check_error_normality}
shapiro.test(fit.all$scl.cond.res)
qqnorm(y=fit.all$scl.cond.res,
       main=paste0(response, ": scaled conditional residuals"))
qqline(y=fit.all$scl.cond.res, col="red")
```

### Error temporal independence

```{r}
tmp <- do.call(cbind, lapply(sort(unique(fit.all$year)), function(year){
  fit.all$scl.cond.res[fit.all$year == year]
}))
colnames(tmp) <- sort(unique(fit.all$year))
rownames(tmp) <- fit.all$geno[fit.all$year == unique(fit.all$year)[1]]
panel.cor <- function(x, y, digits=2){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y, use="complete.obs")
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex=cex.cor * abs(r),
       col=ifelse(r >= 0, "red", "blue"))
}
pairs(tmp, lower.panel=panel.cor, upper.panel=panel.smooth)
```

### Error spatial independence

Skipped.

### Outlying genotypes

```{r check_outlying_genos_plot}
x.lim <- max(abs(geno.blups))
par(mar=c(5,6,4,1))
plot(x=geno.blups, y=1:length(geno.blups),
     xlim=c(-x.lim, x.lim),
     xlab="genotypic BLUPs",
     main=response,
     yaxt="n", ylab="")
axis(side=2, at=1:length(geno.blups), labels=names(geno.blups), las=1)
abline(v=0, lty=2)
idx <- which.max(geno.blups)
text(x=geno.blups[idx], y=idx, labels=names(idx), pos=2)
idx <- which.min(geno.blups)
text(x=geno.blups[idx], y=idx, labels=names(idx), pos=4)
summary(geno.var.blups)
head(sort(geno.var.blups))
tail(sort(geno.var.blups))
```

### Normality of genotypic BLUPs

```{r check_normality_geno_BLUPs_qqplot}
shapiro.test(geno.blups)
qqnorm(y=geno.blups, main=paste0(response, ": genotypic BLUPs"), asp=1)
qqline(y=geno.blups, col="red")
```

### Independence between errors and genotypes

```{r check_indep_error_geno_plot}
x.lim <- max(abs(fit.all$scl.cond.res))
lattice::dotplot(geno ~ scl.cond.res, data=fit.all,
                 xlim=c(-x.lim, x.lim),
                 xlab="scaled conditional residuals",
                 main=response,
                 panel=function(x,y,...){
                   panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                   panel.dotplot(x,y,...)
                 })
```

```{r check_indep_error_geno_plot_year_block}
lattice::dotplot(geno ~ scl.cond.res | year, #groups=block,
                 data=fit.all,
                 auto.key=list(space="right"),
                 xlab="scaled conditional residuals",
                 main=response,
                 panel=function(x,y,...){
                   panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                   panel.dotplot(x,y,...)
                 })
```

## Model outputs

### Summary

```{r bestmod_infer}
summary(bestmod.ml)
(vc.ml <- as.data.frame(VarCorr(bestmod.ml)))
summary(bestmod.reml)
(vc.reml <- as.data.frame(VarCorr(bestmod.reml)))
```

### Broad-sense heritability

```{r H2}
reps.geno.year <- tapply(dat.noNA[[trait]],
                         list(dat.noNA$geno, dat.noNA$year),
                         length)
head(reps.geno.year)
(mean.nb.years <- mean(apply(reps.geno.year, 1, function(x){
  sum(! is.na(x))
})))
(mean.nb.reps.per.year <- mean(apply(reps.geno.year, 2, mean, na.rm=TRUE)))
(H2.means <- vc.reml[vc.reml$grp == "geno", "vcov"] /
   (vc.reml[vc.reml$grp == "geno", "vcov"] +
    (vc.reml[vc.reml$grp == "Residual", "vcov"] /
     (mean.nb.years * mean.nb.reps.per.year))))
```

### Confidence intervals

#### Profiling

```{r bestmod_confint_prof, eval=TRUE}
system.time(
    prof <- profile(bestmod.ml, signames=FALSE,
                    parallel="multicore", ncpus=nb.cores))
(ci <- confint(prof, level=0.95))
xyplot(prof, absVal=TRUE, conf=c(0.8, 0.95), #which=1:3,
       main=response)
densityplot(prof,
            main=response)
splom(prof, conf=c(0.8, 0.95), #which=1:3,
      main=response)
```

#### Bootstrapping

```{r bestmod_confint_boot, eval=TRUE}
mySumm <- function(.){
  tmp <- c(ef=fixef(.),
           sd.err=sigma(.),
           sd=sqrt(unlist(VarCorr(.))))
  tmp <- c(tmp,
           tmp["sd.geno"]^2 /
           (tmp["sd.geno"]^2 +
            (tmp["sd.err"]^2 / (mean.nb.years *
                                mean.nb.reps.per.year))))
  names(tmp)[length(tmp)] <- "H2.means"
  tmp <- c(tmp,
           tmp["sd.geno"] / abs(tmp["ef.(Intercept)"]))
  names(tmp)[length(tmp)] <- "CV.geno"
  return(tmp)
}
mySumm(bestmod.ml)
system.time(
      fit.boot <- bootMer(x=bestmod.ml, FUN=mySumm,
                          nsim=1*10^3, seed=1859,
                          use.u=FALSE, type="parametric",
                          parallel="multicore", ncpus=nb.cores))
fit.boot
for(i in seq_along(fit.boot$t0)){
  message(names(fit.boot$t0)[i])
  plot(fit.boot, index=i,
       main=paste0(response, ": ", names(fit.boot$t0)[i]))
  print(boot.ci(fit.boot, conf=c(0.8, 0.95),
                type=c("norm", "basic", "perc"),
                index=i))
}
```

### Hypothesis testing

### Fixed effects

```{r bestmod_lmerTest}
anova(bestmod.reml.test, ddf="lme4")
anova(bestmod.reml.test, ddf="Satterthwaite")
system.time(
    print(anova(bestmod.reml.test, ddf="Kenward-Roger")))
anova(bestmod.reml.test.final)
```

### Variance components

```{r bestmod_lmerTest_varcomp}
ranova(bestmod.reml)
```

### In-sample prediction

```{r bestmod_in-sample_prediction}
plot(fit.all$fitted, fit.all$response, las=1, asp=1,
     xlab="observed response",
     ylab="fitted responses",
     main=response)
abline(a=0, b=1, lty=2)
abline(v=mean(fit.all$fitted, na.rm=TRUE), lty=2)
abline(h=mean(fit.all$response, na.rm=TRUE), lty=2)
```

```{r check_in-sample_prediction_year_block}
lattice::xyplot(response ~ fitted | year, #groups=block,
                data=fit.all,
                auto.key=list(space="right"),
                xlab="observed response",
                ylab="fitted responses",
                main=response,
                panel=function(x,y,...){
                  panel.abline(a=0, b=1, lty=2)
                  panel.abline(v=mean(x, na.rm=TRUE), lty=2)
                  panel.abline(h=mean(y, na.rm=TRUE), lty=2)
                  panel.xyplot(x,y,...)
                })
```

# Statistical modeling with inter-annual correlation

## Model fitting, comparison and selection

```{r}
form.fix <- paste0(response, " ~ 1")
if(length(all.preds.fix) > 0)
  form.fix <- paste0(form.fix,
                     " + ", paste(all.preds.fix, collapse=" + "))
form.fix
fit <- nlme::lme(fixed=as.formula(form.fix),
                 random=list(geno=~1, rep=~1),
                 correlation=corAR1(),
                 data=dat, method="ML")
```

```{r}
stats::AIC(bestmod.ml)
stats::AIC(fit)
```

According to the AIC, the model with inter-annual correlation is better than the model without.

```{r}
bestmod.ml <- fit
bestmod.reml.nlme <- nlme::lme(fixed=as.formula(form.fix),
                               random=list(geno=~1, rep=~1),
                               correlation=corAR1(),
                               data=dat, method="REML")
bestmod.reml <- bestmod.reml.nlme
```

## Assumption checking

### Residual preparation

```{r}
fit.all <- cbind(dat,
                 response=dat[[response]],
                 cond.res=NA,
                 scl.cond.res=NA,
                 fitted=NA)
fit.all$cond.res <- residuals(bestmod.reml)
fit.all$scl.cond.res <- residuals(bestmod.reml) /
  sigma(bestmod.reml)
fit.all$fitted <- fitted(bestmod.reml)
tmp <- nlme::ranef(bestmod.reml)$geno
geno.blups <- setNames(tmp[,"(Intercept)"], rownames(tmp))
geno.var.blups <- NULL # nlme doesn't produce the variance of the BLUPs
```

### Error homoscedasticity

```{r}
x.lim <- max(abs(fit.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.all$scl.cond.res, y=fit.all$fitted, las=1,
     xlim=c(-x.lim, x.lim),
     xlab="scaled conditional residuals",
     ylab="fitted responses",
     main=response)
abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
```

```{r}
lattice::xyplot(fitted ~ scl.cond.res | year, #groups=block,
                data=fit.all,
                xlab="scaled conditional residuals",
                ylab="fitted responses",
                main=response,
                auto.key=list(space="right"),
                panel=function(x,y,...){
                  panel.abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
                  panel.xyplot(x,y,...)
                })
```

### Error normality

```{r}
shapiro.test(fit.all$scl.cond.res)
qqnorm(y=fit.all$scl.cond.res,
       main=paste0(response, ": scaled conditional residuals"))
qqline(y=fit.all$scl.cond.res, col="red")
```

### Error temporal independence

```{r}
tmp <- do.call(cbind, lapply(sort(unique(fit.all$year)), function(year){
  fit.all$scl.cond.res[fit.all$year == year]
}))
colnames(tmp) <- sort(unique(fit.all$year))
rownames(tmp) <- fit.all$geno[fit.all$year == unique(fit.all$year)[1]]
panel.cor <- function(x, y, digits=2){
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- cor(x, y, use="complete.obs")
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex=cex.cor * abs(r),
       col=ifelse(r >= 0, "red", "blue"))
}
pairs(tmp, lower.panel=panel.cor, upper.panel=panel.smooth)
```

### Error spatial independence

Skipped.

### Outlying genotypes

```{r}
x.lim <- max(abs(geno.blups))
par(mar=c(5,6,4,1))
plot(x=geno.blups, y=1:length(geno.blups),
     xlim=c(-x.lim, x.lim),
     xlab="genotypic BLUPs",
     main=response,
     yaxt="n", ylab="")
axis(side=2, at=1:length(geno.blups), labels=names(geno.blups), las=1)
abline(v=0, lty=2)
idx <- which.max(geno.blups)
text(x=geno.blups[idx], y=idx, labels=names(idx), pos=2)
idx <- which.min(geno.blups)
text(x=geno.blups[idx], y=idx, labels=names(idx), pos=4)
summary(geno.var.blups)
head(sort(geno.var.blups))
tail(sort(geno.var.blups))
```

### Normality of genotypic BLUPs

```{r}
shapiro.test(geno.blups)
qqnorm(y=geno.blups, main=paste0(response, ": genotypic BLUPs"), asp=1)
qqline(y=geno.blups, col="red")
```

### Independence between errors and genotypes

```{r}
x.lim <- max(abs(fit.all$scl.cond.res))
lattice::dotplot(geno ~ scl.cond.res, data=fit.all,
                 xlim=c(-x.lim, x.lim),
                 xlab="scaled conditional residuals",
                 main=response,
                 panel=function(x,y,...){
                   panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                   panel.dotplot(x,y,...)
                 })
```

```{r}
lattice::dotplot(geno ~ scl.cond.res | year, #groups=block,
                 data=fit.all,
                 auto.key=list(space="right"),
                 xlab="scaled conditional residuals",
                 main=response,
                 panel=function(x,y,...){
                   panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                   panel.dotplot(x,y,...)
                 })
```

## Model outputs

Useful function:
```{r}
as.data.frame.VarCorr.lme <- function(x){
  data.frame(grp=gsub(" =", "", grep("Intercept", rownames(x),
                                     value=TRUE, invert=TRUE)),
             vcov=as.numeric(x[-grep(" =", rownames(x)),
                               "Variance"]),
             stringsAsFactors=FALSE)
}
```

### Summary

```{r}
summary(bestmod.ml)
(vc.ml <- as.data.frame(VarCorr(bestmod.ml)))
summary(bestmod.reml)
(vc.reml <- as.data.frame(VarCorr(bestmod.reml)))
(phi <- as.numeric(coef(bestmod.reml$modelStruct$corStruct,
                        unconstrained=FALSE)))
```

### Broad-sense heritability

```{r}
reps.geno.year <- tapply(dat[[trait]],
                         list(dat$geno, dat$year),
                         length)
head(reps.geno.year)
(mean.nb.years <- mean(apply(reps.geno.year, 1, function(x){
  sum(! is.na(x))
})))
(mean.nb.reps.per.year <- mean(apply(reps.geno.year, 2, mean, na.rm=TRUE)))
(H2.means <- vc.reml[vc.reml$grp == "geno", "vcov"] /
   (vc.reml[vc.reml$grp == "geno", "vcov"] +
    (vc.reml[vc.reml$grp == "Residual", "vcov"] /
     (mean.nb.years * mean.nb.reps.per.year))))
```

### Confidence intervals

#### Profiling

Not available for nlme.

#### Bootstrapping

```{r bestmod_confint_boot_nlme}
mySumm <- function(x, form.fix, mean.nb.years, mean.nb.reps.per.year){
  if(is.data.frame(x)){
    x <- nlme::lme(fixed=as.formula(form.fix),
                   random=list(geno=~1, rep=~1),
                   correlation=corAR1(),
                   data=x, method="ML")
  }
  as.data.frame.VarCorr.lme <- function(x){
    data.frame(grp=gsub(" =", "", grep("Intercept", rownames(x),
                                       value=TRUE, invert=TRUE)),
               vcov=as.numeric(x[-grep(" =", rownames(x)),
                                 "Variance"]),
               stringsAsFactors=FALSE)
  }
  vc <- as.data.frame(VarCorr(x))
  sd <- setNames(sqrt(vc[-nrow(vc), "vcov"]),
                 paste0("sd.", vc[-nrow(vc), "grp"]))
  tmp <- c(ef=fixef(x),
           sd.err=sigma(x),
           sd)
  tmp <- c(tmp,
           tmp["sd.geno"]^2 /
           (tmp["sd.geno"]^2 +
            (tmp["sd.err"]^2 / (mean.nb.years *
                                mean.nb.reps.per.year))))
  names(tmp)[length(tmp)] <- "H2.means"
  tmp <- c(tmp,
           tmp["sd.geno"] / abs(tmp["ef.(Intercept)"]))
  names(tmp)[length(tmp)] <- "CV.geno"
  tmp <- c(tmp,
           coef(x$modelStruct$corStruct, unconstrained=FALSE))
  names(tmp)[length(tmp)] <- "phi"
  return(tmp)
}
mySumm(bestmod.ml, form.fix, mean.nb.years, mean.nb.reps.per.year)
## set.seed(1859)
## system.time(
##     fit.boot <- lmeresampler::bootstrap(model=bestmod.ml,
##                                         fn=mySumm,
##                                         type="parametric",
##                                         B=1*10^3))
## Error in parametric_bootstrap.lme(model, fn, B) : 
##   not implemented for multiple levels of nesting
mySimulate <- function(data, par.est){
  G <- nlevels(data$geno)
  R <- nlevels(data$rep)
  T <- nlevels(data$year)

  Z.year <- model.matrix(~ 1 + year, data=data)
  year.effects <- c(par.est$intercept,
                    rnorm(n=T-1, mean=0, sd=sqrt(par.est$var.year)))

  Z.geno <- model.matrix(~ -1 + geno, data=data)
  geno.vals <- rnorm(n=G, mean=0, sd=sqrt(par.est$var.geno))

  mat.cor.error <- corrMatAR1(n=T, rho=par.est$phi)
  vcov.error <- cor2cov(x=mat.cor.error, sd=sqrt(par.est$var.err))
  errors <- c(mvrnorm(n=G*R, mu=rep(0, T), Sigma=vcov.error))

  y <- Z.year %*% year.effs + Z.geno %*% geno.vals + errors

  data$prod <- y
  return(data)
}
myMle <- list(intercept=fixef(bestmod.ml)[["(Intercept)"]],
              var.year=var(fitted(bestmod.ml)),
              var.geno=vc.ml[vc.ml == "geno", "vcov"],
              var.err=vc.ml[vc.ml == "Residual", "vcov"],
              phi=phi)
p2f <- "perennial-fruit-plant-biparental-design_with-corr-boot.RData"
if(! file.exists(p2f)){
  st <- system.time(
      fit.boot <- boot(data=dat,
                       statistic=mySumm,
                       R=1*10^3, sim="parametric",
                       ran.gen=mySimulate, mle=myMle,
                       parallel="multicore", ncpus=nb.cores,
                       form.fix=form.fix,
                       mean.nb.years=mean.nb.years,
                       mean.nb.reps.per.year=mean.nb.reps.per.year))
  print(st)
  save(fit.boot, file=p2f)
  print(tools::md5sum(path.expand(p2f)))
} else{
  print(tools::md5sum(path.expand(p2f)))
  load(p2f)
}
fit.boot
for(i in seq_along(fit.boot$t0)){
  message(names(fit.boot$t0)[i])
  plot(fit.boot, index=i,
       main=paste0(response, ": ", names(fit.boot$t0)[i]))
  print(boot.ci(fit.boot, conf=c(0.8, 0.95),
                type=c("norm", "basic", "perc"),
                index=i))
}
```

### In-sample prediction

```{r bestmod_in-sample_prediction_nlme}
plot(fit.all$fitted, fit.all$response, las=1, asp=1,
     xlab="observed response",
     ylab="fitted responses",
     main=response)
abline(a=0, b=1, lty=2)
abline(v=mean(fit.all$fitted, na.rm=TRUE), lty=2)
abline(h=mean(fit.all$response, na.rm=TRUE), lty=2)
```

```{r check_in-sample_prediction_year_block_nlme}
lattice::xyplot(trait ~ fitted | year, #groups=block,
                data=fit.all,
                auto.key=list(space="right"),
                xlab="observed response",
                ylab="fitted responses",
                main=response,
                panel=function(x,y,...){
                  panel.abline(a=0, b=1, lty=2)
                  panel.abline(v=mean(x, na.rm=TRUE), lty=2)
                  panel.abline(h=mean(y, na.rm=TRUE), lty=2)
                  panel.xyplot(x,y,...)
                })
```


# Comparison

```{r}
geno.blups.lme4 <- ranef(bestmod.reml.lme4, condVar=TRUE, drop=TRUE)$geno
geno.var.blups.lme4 <- setNames(attr(geno.blups.lme4, "postVar"),
                                names(geno.blups.lme4))
tmp <- nlme::ranef(bestmod.reml.nlme)$geno
geno.blups.nlme <- setNames(tmp[,"(Intercept)"], rownames(tmp))
## geno.var.blups.lme4 <- setNames(attr(geno.blups.lme4, "postVar"),
##                                 names(geno.blups.lme4))
```

```{r}
cor(geno.blups.lme4, geno.blups.nlme)
plot(geno.blups.lme4, geno.blups.nlme)
abline(lm(geno.blups.nlme ~ geno.blups.lme4), col="red")
```



# Appendix

```{r info}
t1 <- proc.time()
t1 - t0
print(sessionInfo(), locale=FALSE)
```
