---
title: "Variable selection in quantitative genetics"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
opts_knit$set(progress=TRUE, verbose=TRUE)
```


# Overview

This document requires external packages:
```{r load_pkg}
suppressPackageStartupMessages(library(varbvs))
suppressPackageStartupMessages(library(coda))
suppressPackageStartupMessages(library(rgs3))
suppressPackageStartupMessages(library(rutilstimflutre))
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Simulate some data

## Genotypes

```{r}
set.seed(1859)
nb.genos <- 200
Ne <- 10^4
nb.chroms <- 10
chrom.len <- 10^5
mu <- 10^(-8)
c.rec <- 10^(-7)
genomes <- simulCoalescent(nb.inds=nb.genos,
                           nb.reps=nb.chroms,
                           pop.mut.rate=4 * Ne * mu * chrom.len,
                           pop.recomb.rate=4 * Ne * c.rec * chrom.len,
                           chrom.len=chrom.len,
                           get.alleles=TRUE)
plotHistAllelFreq(X=genomes$genos)
X <- discardSnpsLowMaf(X=genomes$genos, thresh=0.01)
```

## Phenotypes

```{r}
set.seed(1859)
phenos <- simulBvsr(Q=1, X=X, pi=0.01, pve=0.7, sigma.a2=1, min.maf=0.01)
sum(phenos$gamma != 0)
summary(abs(phenos$a[phenos$gamma == 1]))
```


# Explore the data

```{r}
hist(phenos$Y[,1], breaks="FD",
     xlab="phenotypic values", main="Simulated data",
     col="grey", border="white", las=1)
```


# Perform inference SNP-by-SNP with GEMMA

```{r}
fit.gemma <- gemma(model="ulmm", y=phenos$Y[,1], X=X, maf=0.01,
                   snp.coords=genomes$snp.coords, alleles=genomes$alleles,
                   W=phenos$W, clean="all")
cols <- rep("black",ncol(X)); cols[phenos$gamma==1] <- "red"
pvadj <- qqplotPval(fit.gemma$tests$p_wald, col=cols, ctl.fdr.bh=TRUE,
                    plot.signif=TRUE)
t(binaryClassif(known.nulls=phenos$gamma == 0,
                called.nulls=pvadj$pv.bh > 0.05))
cor(phenos$a[phenos$gamma == 1],
    fit.gemma$tests[names(phenos$a[phenos$gamma == 1]), "beta"])
```


# Perform inference all SNPs jointly with varbvs

```{r}
fit.varbvs <- varbvs(X=phenos$X.A, Z=NULL, y=phenos$Y[,1], verbose=FALSE)
print(fit.varbvs.s <- summary(fit.varbvs))
names(sort(phenos$a[phenos$gamma == 1], decreasing=TRUE))
subset.snps <- unique(c(as.character(fit.varbvs.s$top.vars$variable),
                        names(sort(phenos$a[phenos$gamma == 1], decreasing=TRUE))))
subset.coords <- genomes$snp.coords[subset.snps,]
(subset.coords <- subset.coords[order(subset.coords$chr, rownames(subset.coords)),])
ld <- estimLd(X=X[, rownames(subset.coords)], snp.coords=subset.coords)
ld[ld$loc1 == "snp2270",]
```

Inference concerning $\pi$:
```{r}
(pi.hat <- 10^(fit.varbvs.s$logodds$x0) / (1 + 10^(fit.varbvs.s$logodds$x0)))
(pi.hat.low <- 10^(fit.varbvs.s$logodds$a) / (1 + 10^(fit.varbvs.s$logodds$a)))
(pi.hat.high <- 10^(fit.varbvs.s$logodds$b) / (1 + 10^(fit.varbvs.s$logodds$b)))
```

Inference concerning the PIPs:
```{r}
w <- c(normalizelogweights(fit.varbvs$logw))
pips <- c(fit.varbvs$alpha %*% w)
cols <- rep("black", ncol(phenos$X.A))
cols[phenos$gamma != 0] <- "red"
plot(x=1:ncol(phenos$X.A), y=pips, col=cols, las=1, xlab="SNPs", ylab="PIP",
     main="Posterior inclusion probabilities (varbvs)")
```

TODO: Inference concerning the SNP effects:



# Perform inference all SNPs jointly with GS3

Prepare the input files:
```{r}
task.id <- "test"
dat <- data.frame(geno.id=rownames(phenos$Y),
                  overall.mean=1,
                  pheno=phenos$Y[,1],
                  weight=NA)
inds <- setNames(object=1:nlevels(dat$geno.id),
                 nm=levels(dat$geno.id))
data.GS3.file <- paste0(task.id, "_data.tsv")
writeDataForGs3(x=dat, file=data.GS3.file, inds=inds,
                col.id=which(colnames(dat) == "geno.id"),
                col.traits=which(colnames(dat) == "pheno"))
genos.GS3.file <- paste0(task.id, "_genos.tsv")
writeGenosForGs3(x=X, file=genos.GS3.file, inds=inds)
```

Set up the configuration:
```{r}
ptl <- data.frame(position=c(2, ncol(dat) + 1),
                  type=c("cross", "add_SNP"),
                  nlevels=c(1, 0))
config <- getDefaultConfig(
    nb.snps=ncol(X),
    rec.id=which(colnames(dat) == "geno.id"),
    twc=c(which(colnames(dat) == "pheno"), 0),
    method="VCE",
    ptl=ptl,
    use.mix="T")
config$niter <- 10^5
config$burnin <- 5*10^3
config$thin <- 5
config$ap
getMeanVarBetaDist(1, 10)
stopifnot(isValidConfig(config))
config.file <- writeConfigForGs3(config=config,
                                 data.file=data.GS3.file,
                                 genos.file=genos.GS3.file,
                                 task.id=task.id)
```

Run GS3, and assess convergence:
```{r, fig.height=15}
stdouterr.GS3.file <- execGs3(config.file, task.id)
vcs <- vcs2mcmc(config$vcs.file)
vcs <- vcs[, c("vara","vare","pa_1","2varapqpi")]
plot(vcs)
raftery.diag(vcs, q=0.5)
summary(vcs)
```

Look at the proportion of variance explained by breeding values:
```{r}
pve.bv <- vcs[[1]][,"2varapqpi"] / (vcs[[1]][,"2varapqpi"] + vcs[[1]][,"vare"])
summary(pve.bv)
plot(pve.bv)
```

Look at SNP effects and inclusion:
```{r}
sols <- read.table(file=config$sol.file, header=TRUE)
table(sols$effect)
sols$solution[sols$effect == 1]
post.snps <- sols[sols$effect == 2,]
rownames(post.snps) <- colnames(X)
summary(post.snps$solution)
summary(post.snps$p)
true.qtls <- names(phenos$gamma[phenos$gamma != 0])
post.snps[true.qtls,]
cols <- rep("black", ncol(X))
cols[phenos$gamma != 0] <- "red"
plot(x=1:ncol(X), y=post.snps[names(phenos$gamma), "p"],
     col=cols, las=1, xlab="SNPs", ylab="PIP",
     main="Posterior inclusion probabilities (GS3 BayesCPi)")
```

Clean files:
```{r}
cleanGs3(config, config.file, task.id)
file.remove(data.GS3.file)
file.remove(genos.GS3.file)
```


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
