---
title: "Tutorial on variance partitioning"
author: "Timothée Flutre (INRAE)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: TRUE
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: 5
    number_sections: TRUE
urlcolor: blue
---


# Preamble

This document was generated from a file in the [Rmd](https://cran.r-project.org/package=rmarkdown) format.
Both the source and target files are under the [CC BY-SA](https://creativecommons.org/licenses/by-sa/3.0/fr/deed.en) license.
The source file is versioned [online](https://github.com/timflutre/rutilstimflutre/tree/master/misc).


# Overview

This tutorial aims at exploring the partitioning of variance between factors in linear models.
In an simulated example, `yield` as the response variable is regressed on two factors, `fertilization` and `genotype`.

External packages are required:
```{r}
suppressPackageStartupMessages(library(MuMIn))
suppressPackageStartupMessages(library(lme4))
## suppressPackageStartupMessages(library(brms))
suppressPackageStartupMessages(library(rsq))
suppressPackageStartupMessages(library(vegan))
```


# Write the model

Let $y_{ijk}$ be the yield of genotype $i$ with fertilization level $j$ in micro-plot $k$:
\[
y_{ijk} = \mu + a_i + b_j + d_{ij} + \epsilon_{ijk}
\]

In matrix notation, when all effects are modeled as fixed:
\[
\boldsymbol{y} = X \boldsymbol{\beta} + \boldsymbol{\epsilon} \text{ with } \boldsymbol{\epsilon} \sim \mathcal{N}_N(\boldsymbol{0}, R)
\]

In matrix notation, when fertilization effects are modeled as fixed and genotype effects as random:
\[
\boldsymbol{y} = X \boldsymbol{\beta} + Z \boldsymbol{u} + \boldsymbol{\epsilon} \text{ with } \boldsymbol{u} \sim \mathcal{N}_I(\boldsymbol{0}, G) \text{ and } \boldsymbol{\epsilon} \sim \mathcal{N}_N(\boldsymbol{0}, R)
\]

Usually, $G = \sigma_u^2 A$ and $R = \sigma^2 \text{Id}$.


# Simulate some data

## Choose the data dimensions

```{r}
I <- 10  # number of genotypes
J <- 3   # number of fertilization level
K <- 4   # number of replicates per treatment
(N <- I * J * K) # total number of observations
```

## Set up the data structure

```{r}
(levGenos <- sprintf("cv%02i", 1:I))
(levFerti <- c("low", "med", "high"))
(levReps <- LETTERS[1:K])
dat <- data.frame(geno=rep(levGenos, each=J*K),
                  ferti=rep(rep(levFerti, each=K), I),
                  rep=rep(levReps, I*J),
                  yield=NA)
dat$ferti <- factor(dat$ferti, levels=levFerti) # so that levels are ordered
head(dat)
str(dat)
```

## Generate the data

Set the seed:
```{r}
set.seed(12345) # to make the simulation reproducible
```

Make the design matrix:
```{r}
X <- model.matrix(~ 1 + geno + ferti + geno:ferti, data=dat,
                  contrasts.arg=list(geno="contr.sum", ferti="contr.sum"))
dim(X)
colnames(X)
```

Fix the intercept:
```{r}
mu <- 100
```

Draw genotype effects:
```{r}
sd_a <- 10
a <- rnorm(n=I-1,
           mean=0, sd=sd_a)
```

Draw fertilization effects:
```{r}
sd_b <- 10
b <- rnorm(n=J-1, mean=0, sd=sd_b)
```

Draw interaction effects:
```{r}
sd_d <- 2
d <- rnorm(n=(I-1)*(J-1), mean=0, sd=sd_d)
```

Make the effect vector:
```{r}
beta <- matrix(c(mu, a, b, d), nrow=ncol(X), ncol=1)
rownames(beta) <- colnames(X)
```

Draw errors:
```{r}
sd_e <- 1
epsilon <- rnorm(n=N, mean=0, sd=sd_e)
```

Compute the yield:
```{r}
y <- X %*% beta + epsilon
dat$yield <- as.vector(y)
```


# Explore the data

## Data means

```{r}
str(dat)
(grandMean <- mean(dat$yield))
(genoMeans <- tapply(dat$yield, dat$geno, mean))
(fertiMeans <- tapply(dat$yield, dat$ferti, mean))
```

## Plots

```{r}
hist(dat$yield, col="grey", border="white", las=1)
abline(v=grandMean, lty=2)
legend("topright", legend="grand mean", lty=2, bty="n")
```

```{r}
colsFerti <- c("yellow", "orange", "red")
boxplot(yield ~ geno + ferti, data=dat, las=2, varwidth=TRUE,
        col=rep(colsFerti, each=I),
        main="Boxplots of dat$yield")
abline(h=grandMean, lty=2)
legend("bottomright", bty="n",
       legend=c("grand mean", paste0("ferti.: ", levFerti)),
       lty=c(2, rep(NA, J)), col=c("black", colsFerti),
       fill=c(NA, colsFerti), border=c("white", colsFerti),
       x.intersp=c(2, rep(0.5, J)))
```


# Perform the statistical analysis

## Model with fixed-effects only

### Fit the models

```{r}
fitFix1 <- lm(yield ~ 1 + geno, data=dat,
              contrasts=list(geno="contr.sum"))
fitFix2 <- lm(yield ~ 1 + ferti, data=dat,
              contrasts=list(ferti="contr.sum"))
fitFix3a <- lm(yield ~ 1 + geno + ferti, data=dat,
               contrasts=list(geno="contr.sum", ferti="contr.sum"))
fitFix3b <- lm(yield ~ 1 + geno + ferti + geno:ferti, data=dat,
               contrasts=list(geno="contr.sum", ferti="contr.sum"))
```

### Compare the models

```{r}
extractAIC(fitFix1)
extractAIC(fitFix2)
extractAIC(fitFix3a)
extractAIC(fitFix3b)
```

Best model:
```{r}
fitGlobModFix <- lm(yield ~ 1 + geno + ferti + geno:ferti, data=dat,
                    contrasts=list(geno="contr.sum", ferti="contr.sum"),
                    na.action="na.fail")
(allModelsFix <- dredge(global.model=fitGlobModFix, rank="AICc"))
```

### Select the best model

```{r}
fitBestModFix <- get.models(allModelsFix, 1)[[1]]
```

### Check the assumptions

```{r}
par(mfrow=c(1,2))
plot(fitBestModFix, which=c(1,2), ask=FALSE)
```

### Look at the results

#### Overall

```{r}
summary(fitBestModFix)
```

#### Analysis of variance

ANOVA table:
```{r}
(aovFix <- anova(fitBestModFix))
```

```{r}
(tmp <- setNames(aovFix[, "Mean Sq"] / sum(aovFix[, "Mean Sq"]),
                 rownames(aovFix)))
sum(tmp)
sum(tmp[-length(tmp)])
```

#### R2

```{r}
summary(fitBestModFix)$r.squared
summary(fitBestModFix)$adj.r.squared
```

```{r}
rsq.partial(objF=fitBestModFix, adj=FALSE, type="v")
rsq.partial(objF=fitBestModFix, adj=TRUE, type="v")
rsq.partial(objF=fitBestModFix, adj=FALSE, type="sse")
rsq.partial(objF=fitBestModFix, adj=TRUE, type="sse")
```

#### Partition of variance

```{r}
vpG <- varpart(dat$yield,
               ~ geno,
               ~ geno + ferti + geno:ferti, data=dat)
vpG
plot(vpG)
```

```{r}
vpF <- varpart(dat$yield,
               ~ ferti,
               ~ geno + ferti + geno:ferti, data=dat)
vpF
plot(vpF)
```

```{r}
vpGF <- varpart(dat$yield,
                ~ geno + ferti,
               ~ geno + ferti + geno:ferti, data=dat)
vpGF
plot(vpGF)
```

## Model with fixed and random effects

### Fit the models

```{r}
fitMix1 <- lmer(yield ~ 1 + (1|geno), data=dat, REML=FALSE)
fitMix2 <- lm(yield ~ 1 + ferti, data=dat,
              contrasts=list(ferti="contr.sum"))
fitMix3a <- lmer(yield ~ 1 + (1|geno) + ferti, data=dat,
                 contrasts=list(ferti="contr.sum"), REML=FALSE)
fitMix3b <- lmer(yield ~ 1 + ferti + (ferti-1|geno), data=dat,
                 contrasts=list(ferti="contr.sum"), REML=FALSE)
```

### Compare the models

```{r}
extractAIC(fitMix1)
extractAIC(fitMix2)
extractAIC(fitMix3a)
extractAIC(fitMix3b)
```

Best model:
```{r}
fitGlobModMix <- lmer(yield ~ 1 + ferti + (ferti-1|geno), data=dat,
                      contrasts=list(ferti="contr.sum"),
                      na.action="na.fail", REML=FALSE)
(allModelsMix <- dredge(global.model=fitGlobModMix, rank="AICc"))
```

### Select the best model

```{r}
fitBestModMix <- get.models(allModelsMix, 1)[[1]]
fitBestModFix <- update(fitBestModMix, REML=TRUE)
```

### Check the assumptions

```{r}
plot(fitBestModMix)
qqmath(fitBestModMix)
```

### Look at the results

#### Overall

```{r}
summary(fitBestModMix)
```

#### Decomposition of the variance

```{r}
anova(fitBestModMix)
as.data.frame(VarCorr(fitBestModMix))
```


# References

- https://cran.r-project.org/web/packages/rsq/rsq.pdf#page=15

- https://cran.r-project.org/web/packages/vegan/vegan.pdf#page=266

- Legendre et Legendre

- Lebarbier et Robin

- Villemereuil et al


# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
