---
title: "Tutorial on parallel computations in R"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
lang: "en"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
##`
## https://github.com/vspinu/polymode/issues/147#issuecomment-399745611
R.v.maj <- as.numeric(R.version$major)
R.v.min.1 <- as.numeric(strsplit(R.version$minor, "\\.")[[1]][1])
if(R.v.maj < 2 || (R.v.maj == 2 && R.v.min.1 < 15))
  stop("requires R >= 2.15", call.=FALSE)

suppressPackageStartupMessages(library(knitr))
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE, fig.align="center")
opts_knit$set(progress=TRUE, verbose=TRUE)
```


# Overview

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```

Since several releases, R natively comes with parallel support:
```{r}
library(parallel)
```

Some context and information can be found in the vignette of this package:
```{r}
#browseVignettes(package="parallel")
```

This package provides a way to determine how many cores are available:
```{r}
(nb.cores <- detectCores())
```

In practice, it is useful to keep at least one core to keep doing other things on the computer while the computations run in parallel:
```{r}
nb.cores <- nb.cores - 1
```


# Make a function

It is frequent to have a set of instructions that one has to launch several times ($n$):
```{r}
n <- 4 # small so that this tutorial doesn't take too long to run
```

When each launch is independent of each other, it is _embarassingly_ easy to parallelize.
But before anything else, it is advised to make a function with all instructions.
Here is an example for the sake of this tutorial:
```{r}
unitRun <- function(){
  Sys.sleep(3) # so that this example takes enough time to run
  runif(1)     # so that this examples involves random numbers
}
```


# Sequential code

## "for" loop

It is usual to start by writing a "for" loop:
```{r}
RNGkind(kind="default")
set.seed(1234) # to be reproducible
out.for.1 <- rep(NA, n)
for(i in 1:n){
  out.for.1[i] <- unitRun()
}
out.for.1
```

Let's do it again to check the reproducibility:
```{r}
RNGkind(kind="default")
set.seed(1234)
out.for.2 <- rep(NA, n)
for(i in 1:n){
  out.for.2[i] <- unitRun()
}
out.for.2
all.equal(out.for.2, out.for.1)
```

## lapply

After having written a "for" loop, it is advised to re-write the code with a call to "lapply":
```{r}
RNGkind(kind="default")
set.seed(1234)
system.time(
  out.lapply.1 <- lapply(X=1:n, FUN=function(i){
    unitRun()
}))
out.lapply.1 <- do.call(c, out.lapply.1) # convert the output list into a vector
out.lapply.1
all.equal(out.lapply.1, out.for.1)
```

# Parallel code

## mclapply

Once the code uses "lapply", it is trivial to re-write it to use "mclapply", but this function works only on Linux:
```{r}
RNGkind(kind="L'Ecuyer-CMRG") # good-quality RNG with multiple independent streams
set.seed(1234)
system.time(
  out.mclapply.1 <- mclapply(X=1:n, FUN=function(i){
    unitRun()
}, mc.cores=nb.cores))
out.mclapply.1 <- do.call(c, out.mclapply.1)
out.mclapply.1
```

Let's do it again to check the reproducibility (only on clusters with the same number of cores):
```{r}
RNGkind(kind="L'Ecuyer-CMRG")
set.seed(1234)
system.time(
  out.mclapply.2 <- mclapply(X=1:n, FUN=function(i){
    unitRun()
}, mc.cores=nb.cores))
out.mclapply.2 <- do.call(c, out.mclapply.2)
out.mclapply.2
all.equal(out.mclapply.2, out.mclapply.1)
```

## parLapply

On Windows, one first has to create a "cluster" object:
```{r}
cl <- makeCluster(spec=nb.cores, type="PSOCK")
```

Then, one can use "parLapply":
```{r}
RNGkind("L'Ecuyer-CMRG")
clusterSetRNGStream(cl=cl, iseed=1234)
clusterExport(cl=cl, varlist="unitRun")
# if unitRun() requires packages, use clusterEvalQ(cl, library(<pkg>))
system.time(
  out.parLapply.1 <- parLapply(cl=cl, X=1:n, fun=function(i){
    unitRun()
}))
out.parLapply.1 <- do.call(c, out.parLapply.1)
out.parLapply.1
```

Let's do it again to check the reproducibility (only on clusters with the same number of cores):
```{r}
RNGkind(kind="L'Ecuyer-CMRG")
clusterSetRNGStream(cl=cl, iseed=1234)
system.time(
  out.parLapply.2 <- parLapply(cl=cl, X=1:n, fun=function(i){
    unitRun()
}))
out.parLapply.2 <- do.call(c, out.parLapply.2)
out.parLapply.2
all.equal(out.parLapply.2, out.parLapply.1)
```

When it's not needed anymore, don't forget to stop the cluster:
```{r}
stopCluster(cl)
```

## Miscellaneous

See this course: https://www.datacamp.com/courses/parallel-programming-in-r.


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
