---
title: "Two-phase analysis in quantitative genetics: example in grapevine"
author: "Timothée Flutre (INRAE)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
colorlinks: true
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: TRUE
urlcolor: blue
---

<!--
This R chunk is used to set up some options.
-->
```{r setup, include=FALSE}
suppressPackageStartupMessages(library(knitr))
if(! "params" %in% ls())
  params <- list(root.dir=NULL)
opts_chunk$set(echo=TRUE, warning=TRUE, message=TRUE, cache=FALSE,
               fig.align="center", root.dir=params$root.dir)
opts_knit$set(progress=TRUE, verbose=TRUE)
```


# Overview

This document aims at exploring two-phase analysis in quantitative genetics as seen with perennial plant species, e.g., grapevine, depending on the amount of missing data, that is when the data set is more or less unbalanced, and the discrepancy between the generative and statistical models.

This document requires external packages to be available:
```{r load_pkg}
suppressPackageStartupMessages(library(parallel))
nb.cores <- ifelse(detectCores() > 20, 20, detectCores() - 1)
suppressPackageStartupMessages(library(scrm))
suppressPackageStartupMessages(library(sp))
suppressPackageStartupMessages(library(randtoolbox))
## suppressPackageStartupMessages(library(DiceDesign))
suppressPackageStartupMessages(library(gstat))
suppressPackageStartupMessages(library(beanplot))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(MuMIn))
suppressPackageStartupMessages(library(boot))
suppressPackageStartupMessages(library(lattice))
suppressPackageStartupMessages(library(lmerTest))
stopifnot(compareVersion("3.0",
                         as.character(packageVersion("lmerTest")))
          != 1)
suppressPackageStartupMessages(library(emmeans))
suppressPackageStartupMessages(library(MM4LMM))
suppressPackageStartupMessages(library(sommer))
## suppressPackageStartupMessages(library(mlmm.gwas))
## suppressPackageStartupMessages(library(varbvs))
## stopifnot(compareVersion("2.5.7",
##                          as.character(packageVersion("varbvs")))
##           != 1)
suppressPackageStartupMessages(library(rutilstimflutre)) # on GitHub
stopifnot(compareVersion("0.170.1",
                         as.character(packageVersion("rutilstimflutre")))
          != 1)
## stopifnot(file.exists(Sys.which("gemma"))) # https://github.com/genetics-statistics/GEMMA/
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```



# Generative model

## Phenotypic level

Dimensions:

* $I = I_p + I_c$: number of genotypes (from a diversity panel, as well as a control)

* $Q = J + K$: number of known, non-genetic factors and covariates

    * $J$: number of blocks (each block contains one plant of each panel genotype as well as several plants of the control)
    
        * $J_T$: number of controls per block
    
    * $K$: number of years

* $R, C$: number of rows and columns of the field layout

* $N = (I-1) * J * K + J_T * J * K$: number of phenotypes

Variables:

* $X$: $N \times Q$ design matrix for factors and covariates

* $\boldsymbol{\beta}$: $Q$-vector of factor and covariate effects, modeled as "fixed"

    * $\boldsymbol{\beta} \sim \mathcal{N}(\boldsymbol{0}, \sigma_\beta^2 \, \text{Id})$ where $\sigma_\beta^2$ is fixed at a "large" value

* $Z$: $N \times I$ design matrix for genotypic values

* $\boldsymbol{g}$: $I$-vector of genotypic values, modeled as "random"
    
    * $\boldsymbol{g} \sim \mathcal{N}(\boldsymbol{0}, \sigma_g^2 \, \text{Id})$
  
    * $H^2 = \frac{\sigma_g^2}{\sigma_g^2 + \sigma^2 / nb.rep}$: broad-sense heritability for the means per genotype

* TODO: geno-year interactions

* $Z_S$: design matrix for spatial heterogeneity

* $\boldsymbol{\eta}$: vector of spatial heterogeneity, here AR1xAR1

    * $\rho_r$: correlation between rows

    * $\rho_c$: correlation between columns

* $\boldsymbol{\epsilon}$: $N$-vector of errors if no spatial heterogeneity, "nugget effect" otherwise

    * $\boldsymbol{\epsilon} \sim \mathcal{N}(\boldsymbol{0}, \sigma^2 \, \text{Id})$

* $\boldsymbol{y}$: $N$-vector of phenotypic values

    * $\boldsymbol{y} = X \, \boldsymbol{\beta} + Z \, \boldsymbol{g} + Z_S \, \boldsymbol{\eta} + \boldsymbol{\epsilon}$

## Genotypic level

* $P$: number of biallelic SNPs

* $\boldsymbol{f}$: $P$-vector of SNP allele frequencies

* $M_a$: $I \times P$ matrix of SNP genotypes coded additively in $[0,2]$

    * $M_{a,c}$: after centering by the $f_p$'s

* $\boldsymbol{\alpha}$: $P$-vector of additive SNP effects

    * dense architecture: $\forall p \in \{1,\ldots,P\}, \; \alpha_p \sim \mathcal{N}(0, \sigma_\alpha^2)$
  
    * sparse architecture: $\forall p \in \{1,\ldots,P\}, \; \alpha_p \sim \pi_0 \, \delta_0 + \pi_1 \, \mathcal{N}(0, \sigma_\alpha^2)$ with $\gamma_{a,p}$ the indicator variable for $\alpha_p \neq 0$
  
    * TODO: BSLMM

* $\boldsymbol{a} = M_{a,c} \, \boldsymbol{\alpha}$: $I$-vector of additive genotypic values

    * $\boldsymbol{a} \sim \mathcal{N}(\boldsymbol{0}, \sigma_a^2 \, A)$
  
    * $A = \frac{M_{a,c} M_{a,c}^T}{2 \, \sum_{p=1}^P f_p (1 - f_p)}$
    
* $M_d$: $I \times P$ matrix of SNP genotypes coded for dominance in $[0,1]$

    * $M_{d,c}$: after centering by the $f_p$'s

* $\boldsymbol{\delta}$: $P$-vector of dominance SNP effects

    * dense architecture: $\forall p \in \{1,\ldots,P\}, \; \delta_p \sim \mathcal{N}(0, \sigma_\delta^2)$
  
    * sparse architecture: $\forall p \in \{1,\ldots,P\}, \; \delta_p \sim \pi_{0,d} \, \delta_0 + \pi_{1,d} \, \mathcal{N}(0, \sigma_\delta^2)$ with $\gamma_{d,p}$ the indicator variable for $\delta_p \neq 0$

* $\boldsymbol{d} = M_{d,c} \, \boldsymbol{\delta}$: $I$-vector of dominance genotypic values

    * $\boldsymbol{d} \sim \mathcal{N}(\boldsymbol{0}, \sigma_d^2 \, D)$
  
    * TODO: $D$

* TODO: epistatic SNP effects (start with additive x additive)

* TODO: epistatic genotypic values
    
* $\boldsymbol{g}$: $I$-vector of (total) genotypic values

    * $\boldsymbol{g} = \boldsymbol{a} + \boldsymbol{d} = M_{a,c} \, \boldsymbol{\alpha} + M_{d,c} \, \boldsymbol{\delta}$ (assuming no epistasis)
  
    * $h^2 = \frac{\sigma_a^2}{\sigma_g^2 + \sigma^2}$: narrow-sense heritability


# Strategy

No unified statistical model can be currently used to reliably perform inference of SNP effects in the presence of genotype-year interactions and spatial heterogeneity.
As a result, a two-phase strategy should be used.

Phases:

1. estimate/predict $\boldsymbol{g}$ as "average" $\boldsymbol{y}$ per genotype;

2. regress $\hat{\boldsymbol{g}}$ on $M$ to infer $\boldsymbol{\alpha}$ and $\boldsymbol{\gamma_a}$ (and $\boldsymbol{\delta}$ and $\boldsymbol{\gamma_d}$ ).

Questions:

* use BLUEs or BLUPs of $\boldsymbol{g}$ in phase 2? $\rightarrow$ BLUPs if unbalanced data set

* account for spatial heterogeneity by kriging thanks to the controls?

* use weights in phase 2 in the presence of missing data?


# Simulation of genetic data

## Initial population

Set the seed:
```{r set_seed_geno}
set.seed(1859)
```

Simulate haplotypes and genotypes in one population at the Hardy-Weinberg equilibrium:
```{r simul_haplos_genos}
nb.genos <- 1 * 10^3
nb.chrs <- 10
chr.len.phy <- 10^5     # chromosome physical length in base pairs
mu <- 5 * 10^(-8)       # neutral mutation rate in events / base / generation
u <- mu * chr.len.phy   #  in events / chrom / gen
chr.len.gen <- 10^(-1)  # chromosome genetic length in Morgans
c.r <- chr.len.gen / chr.len.phy # recomb rate in events / base / gen
r <- c.r * chr.len.phy  #  in events / chrom / gen
Ne <- 10^4              # effective population size
theta <- 4 * Ne * u     # scaled neutral mutation rate in events / chrom
rho <- 4 * Ne * r       # scaled recomb rate in events / chrom
p2f <- "two-phase_quantgen_genomes.RData"
if(! file.exists(p2f)){
  st <- system.time(
      genomes <- simulCoalescent(nb.inds=nb.genos,
                                 nb.reps=nb.chrs,
                                 pop.mut.rate=theta,
                                 pop.recomb.rate=rho,
                                 chrom.len=chr.len.phy,
                                 get.alleles=TRUE))
  print(st)
  save(genomes, file=p2f)
  print(tools::md5sum(path.expand(p2f)))
} else{
  print(tools::md5sum(path.expand(p2f)))
  load(p2f)
}
afs.pop <- estimSnpAf(X=genomes$genos)
mafs.pop <- estimSnpMaf(afs=afs.pop)
A.vr <- estimGenRel(X=genomes$genos, afs=afs.pop, method="vanraden1")
mrk2chr <- setNames(genomes$snp.coords$chr, rownames(genomes$snp.coords))
```

Look at some visual checks:
```{r check_simul_genos}
plotHistAllelFreq(afs=afs.pop, main="Allele frequencies")
plotHistMinAllelFreq(mafs=mafs.pop, main="Minor allele frequencies")
## imageWithScale(A.vr, main="Additive genetic relationships")
summary(diag(A.vr))
hist(diag(A.vr), col="grey", border="white")
summary(A.vr[upper.tri(A.vr)])
hist(A.vr[upper.tri(A.vr)], col="grey", border="white")
```

## Bi-parental cross

Set the seed:
```{r set_seed_geno_cross}
set.seed(1859)
```

Choose two individuals as parents:
```{r choose_parents}
(idx.parents <- sample(x=1:nb.genos, size=2, replace=FALSE))
A.vr[idx.parents, idx.parents]
genos.parents <- genomes$genos[idx.parents,]
names.parents <- rownames(genos.parents)
haplos.parents <- getHaplosInds(haplos=genomes$haplos,
                                ind.names=names.parents)
```

Cross them several times to make offsprings:
```{r cross_to_make_offsprings}
nb.offs <- 200
names.offs <- paste0("off-",
                     sprintf(fmt=paste0("%0", floor(log10(nb.offs))+1, "i"),
                             1:nb.offs))
head(crosses.off <- data.frame(parent1=rep(names.parents[1], nb.offs),
                               parent2=rep(names.parents[2], nb.offs),
                               child=names.offs,
                               stringsAsFactors=FALSE))
loc.crovers.off <- drawLocCrossovers(crosses=crosses.off,
                                     nb.snps=sapply(haplos.parents, ncol),
                                     simplistic=FALSE,
                                     verbose=1)
haplos.offs <- makeCrosses(haplos=haplos.parents, crosses=crosses.off,
                           loc.crossovers=loc.crovers.off,
                           howto.start.haplo=0)
genos.offs <- segSites2allDoses(seg.sites=haplos.offs,
                                ind.ids=getIndNamesFromHaplos(haplos.offs),
                                snp.ids=rownames(genomes$snp.coords))
dim(genos.doses <- rbind(genos.parents, genos.offs))
genos.classes <- genoDoses2genoClasses(X=genos.doses,
                                       alleles=genomes$alleles)
genos.jm <- genoClasses2JoinMap(x=genos.classes)
genos.jm[1:3, 1:14]
tests.seg <- filterSegreg(genos.jm[,-c(1:8)], return.counts=TRUE)
```

The JoinMap format will be used to encode genotypes from both parents and offsprings as it allows to specify segregation types and phases, and can be natively read by the R `qtl` package.

Plot pedigree:
```{r plot_ped}
ped <- data.frame(ind=c(names.parents,
                        crosses.off$child),
                  mother=c(rep(NA, length(names.parents)),
                           crosses.off$parent1),
                  father=c(rep(NA, length(names.parents)),
                           crosses.off$parent2),
                  gen=c(rep(0, length(names.parents)),
                        rep(1, nrow(crosses.off))),
                  stringsAsFactors=FALSE)
ped.tmp <- rbind(ped[1:5,],
                 c(ind="off-...", ped[5, -1]),
                 c(ind="off-....", ped[5, -1]),
                 ped[nrow(ped),])
plotPedigree(inds=ped.tmp$ind, mothers=ped.tmp$mother, fathers=ped.tmp$father,
             generations=ped.tmp$gen, main="Pedigree of the controlled cross")
```

Check additive genetic relationships:
```{r add_gen_rel}
A.vr.cross <- estimGenRel(X=genos.doses, afs=afs.pop, method="vanraden1")
A.t.cross <- estimGenRel(X=genos.doses, afs=afs.pop, method="toro2011_eq10")
cor(c(A.vr.cross), c(A.t.cross))
## imageWithScale(A.vr.cross, main="Additive genetic relationships of crosses")
## imageWithScale(A.vr.cross[1:10, 1:10],
##                main="Additive genetic relationships of crosses (subset)",
##                idx.rownames=1:10, idx.colnames=1:10)
summary(diag(A.vr.cross))
summary(A.vr.cross[upper.tri(A.vr.cross)])
summary(A.vr.cross[names.parents[1], grep("off", colnames(A.vr.cross))])
summary(A.vr.cross[names.parents[2], grep("off", colnames(A.vr.cross))])
summary(A.t.cross[names.parents[1], grep("off", colnames(A.t.cross))])
summary(A.t.cross[names.parents[2], grep("off", colnames(A.t.cross))])
```

Under HWE in a single population, the additive genetic relationships between all parent-child pairs should be centered around 0.5, corresponding to a coancestry coefficient of 1/4.


# Simulation of phenotypic data

## Experimental design

Set the seed:
```{r set_seed_pheno}
set.seed(1859)
```

### Structure for a given year

```{r set_exp_design_year}
I.p <- 279       # nb of genotypes in the panel
I.c <- 1         # control
I <- I.p + I.c   # total nb of genotypes
stopifnot(I <= nb.genos)
J <- 5           # nb of blocks
nb.controls.per.block <- 30
(I.block <- I.p + I.c * nb.controls.per.block) # total nb of plants in a block
(N <- I.block * J) # nb of phenotypic data per year
uniq.geno.names <- sample(x=rownames(genomes$genos),
                          size=I, replace=FALSE)
uniq.geno.names <- sort(uniq.geno.names)
(control.name <- uniq.geno.names[1])
uniq.geno.p.names <- uniq.geno.names[-1] # panel names, w/o the control
uniq.block.names <- LETTERS[1:J]
block.geno.names <- c(uniq.geno.p.names,
                      rep(control.name, nb.controls.per.block))
stopifnot(length(block.geno.names) == I.block)
dat.year <- data.frame(geno=rep(NA, N),
                       control=FALSE,
                       block=rep(uniq.block.names, each=I.block),
                       rank=NA,
                       location=NA,
                       year=NA,
                       stringsAsFactors=TRUE)
dat.year$geno <- do.call(c, lapply(1:J, function(j){
  sample(x=block.geno.names, size=I.block)
}))
dat.year$geno <- factor(dat.year$geno)
dat.year$control[dat.year$geno == control.name] <- TRUE
str(dat.year)
```

### Field layout

Block layout in the field:
```{r set_block_layout}
## field view:
## tions  C (F) ...
## ca     B  E  ...
## lo     A  D  ...
##     ranks ->
## matrix view: ranks are columns, locations are rows
## A D
## B E
## C NA
nb.block.rows <- 3
(nb.block.cols <- ceiling(J / nb.block.rows))
lay.blocks <- matrix(data=NA,
                     nrow=nb.block.rows,
                     ncol=nb.block.cols)
block.idx <- 0
for(j in 1:nb.block.cols){
  for(i in 1:nb.block.rows){
    block.idx <- block.idx + 1
    if(i * j > J)
      next
    lay.blocks[i, j] <- uniq.block.names[block.idx]
  }
}
lay.blocks
ranks.per.block <- 20
(locs.per.block <- ceiling(I.block / ranks.per.block))
```

Use a space-filling design to locate controls in a block (same design in each block):
```{r}
tmp <- randtoolbox::sobol(n=nb.controls.per.block, dim=2)
## tmp <- DiceDesign::dmaxDesign(n=nb.controls.per.block, dimension=2, range=0.9, niter_max=200)$design
## plot(x=tmp[,1], y=tmp[,2], xlim=c(0,1), ylim=c(0,1))
ranks.ctl <- cut(tmp[,1], breaks=seq(0, 1, length.out=ranks.per.block),
                 labels=FALSE)
locs.ctl <- cut(tmp[,2], breaks=seq(0, 1, length.out=locs.per.block),
                labels=FALSE)
coords.ctl <- data.frame(rank=ranks.ctl,
                         loc=locs.ctl)
coords.ctl <- coords.ctl[order(coords.ctl$rank, coords.ctl$loc),]
rownames(coords.ctl) <- NULL
stopifnot(! anyDuplicated(coords.ctl))
plot(x=ranks.ctl, y=locs.ctl, xlim=c(1,ranks.per.block),
     ylim=c(1,locs.per.block))
```

Add spatial coordinates of the genotypes in each block:
```{r set_spatial_coords}
ctl.space.fill <- TRUE # if FALSE, controls are randomly located
count.blocks <- 0
for(block.j in 1:nb.block.cols){
  for(block.i in 1:nb.block.rows){
    count.blocks <- count.blocks + 1
    if(count.blocks > J)
      break
    block <- lay.blocks[block.i, block.j]
    idx.block <- which(dat.year$block == block)
    rank.first <- 1 + (block.j - 1) * ranks.per.block
    rank.last <- rank.first + ranks.per.block - 1
    loc.first <- 1 + (block.i - 1) * locs.per.block
    loc.last <- loc.first + locs.per.block - 1
    message(paste0("block=", block, " [", block.i, ",", block.j, "]",
                   " ranks=", rank.first, "-", rank.last,
                   " locs=", loc.first, "-", loc.last,
                   " (", idx.block[1], ")"))
    ranks <- rank.first:rank.last
    locs <- loc.first:loc.last
    grid <- expand.grid(list(loc=locs, rank=ranks))
    grid <- grid[1:I.block, c("rank","loc")]
    
    if(ctl.space.fill){
      idx.ctl.block <- which(dat.year$control & dat.year$block == block)
      ranks.ctl.block <- ranks.ctl + (block.j-1) * ranks.per.block
      locs.ctl.block <- locs.ctl + (block.i-1) * locs.per.block
      stopifnot(all(ranks.ctl.block >= rank.first),
                all(ranks.ctl.block <= rank.last),
                all(locs.ctl.block >= loc.first),
                all(locs.ctl.block <= loc.last))
      dat.year$rank[idx.ctl.block] <- ranks.ctl.block
      dat.year$location[idx.ctl.block] <- locs.ctl.block
      stopifnot(sum(is.na(dat.year$rank[idx.block])) == I.p,
                sum(is.na(dat.year$location[idx.block])) == I.p)
      to.rmv <- c()
      for(i in 1:nrow(grid)){
        idx <- which(ranks.ctl.block == grid$rank[i])
        if(length(idx) > 0){
          if(grid$loc[i] %in% locs.ctl.block[idx])
            to.rmv <- c(to.rmv, i)
        }
      }
      grid <- grid[-to.rmv,]
      stopifnot(nrow(grid) == I.p)
    }
    
    idx.grid <- 0
    for(idx.dat in which(! dat.year$control & dat.year$block == block)){
      stopifnot(is.na(dat.year$rank[idx.dat]),
                is.na(dat.year$location[idx.dat]))
      idx.grid <- idx.grid + 1
      dat.year$rank[idx.dat] <- grid$rank[idx.grid]
      dat.year$location[idx.dat] <- grid$loc[idx.grid]
    }
  }
}
range(dat.year$rank, na.rm=TRUE)
(nb.ranks <- length(unique(dat.year$rank[! is.na(dat.year$rank)])))
range(dat.year$location, na.rm=TRUE)
(nb.locs <- length(unique(dat.year$location[! is.na(dat.year$location)])))
```

Plot the field layout:
```{r}
lay <- SpatialPointsDataFrame(coords=dat.year[, c("rank","location")],
                              data=dat.year[, c("geno","control","block")])
summary(lay)
## lay@data$control <- as.factor(lay@data$control)
## lay@data$block <- as.factor(lay@data$block)
spplot(obj=lay, zcol="block", scales=list(draw=TRUE),
       xlab="ranks", ylab="locations",
       main="Field layout",
       key.space="right", aspect="fill")
spplot(obj=lay, zcol="control",
       col.regions=c("green","red"),
       scales=list(draw=TRUE),
       xlab="ranks", ylab="locations",
       main="Controls",
       key.space="right", aspect="fill")
```

### Structure for the whole data set

Duplicate the data structure per year:
```{r}
dat <- dat.year
first.year <- 2011
dat$year <- first.year
K <- 2 # nb of years
if(K > 1){
  uniq.year.names <- as.character(first.year:(first.year + K - 1))
  for(year in uniq.year.names[-1]){
    dat <- rbind(dat, dat.year)
    dat$year[is.na(dat$year)] <- year
  }
}
dat$year <- factor(dat$year)
str(dat)
summary(dat)
```

Add the other regressors:
```{r}
(N <- I.block * J * K) # nb of phenotypic data for all years
stopifnot(N == nrow(dat))
dat$covar1 <- rnorm(n=N, mean=0, sd=1)
dat$covar2 <- rnorm(n=N, mean=0, sd=1)
dat$fact1 <- sample(x=c("pos","neg"), size=N, replace=TRUE,
                    prob=c(0.7, 1-0.7))
dat$fact1 <- as.factor(dat$fact1)
dat$fact2 <- sample(x=c("pos","neg"), size=N, replace=TRUE,
                    prob=c(0.9, 1-0.9))
dat$fact2 <- as.factor(dat$fact2)
str(dat)
```

## Simulation of trait1

### Make design matrices

#### Fixed effects

Assuming covar2 and fact2 have no effect:
```{r simul_phenos_design-mat_eff-fix}
truth.form.fix <- "1 + block + year + covar1 + fact1"
X <- model.matrix(formula(paste0("~ ", truth.form.fix)),
                  data=dat)
dim(X)
Q <- ncol(X)
```

#### Genotypic values

```{r simul_phenos_design-mat_geno-val}
Z <- model.matrix(~ -1 + geno, data=dat)
dim(Z)
```

#### Spatial heterogeneity

```{r simul_phenos_design-mat_spatial-het}
Z.S <- model.matrix(~ -1 + factor(location):factor(rank), data=dat)
dim(Z.S)
colnames(Z.S) <- gsub("factor\\(|\\)", "", colnames(Z.S))
Z.S[1:3, 1:3]
```

### Set input values

To obtain realistic data:
```{r fix_simul_val}
mu <- 100    # intercept
y.min <- 20  # minimal phenotypic value
CV.g <- 0.10 # genotypic coef of variation
h2 <- 0.4    # narrow-sense heritability
```

### Deduce the variances

```{r deduce_var}
(sigma.p2 <- ((mu - y.min) / 4)^2) # phenotypic variance
(sigma.g2 <- (CV.g * mu)^2)        # total genetic variance
stopifnot(sigma.g2 < sigma.p2)
sigma.a2 <- sigma.g2               # additive genetic variance
(sigma2 <- ((1 - h2) / h2) * sigma.a2) # error variance
stopifnot(sigma.g2 + sigma2 < sigma.p2)
(sigma.beta2 <- sigma.p2 - (sigma.g2 + sigma2))
(sigma.alpha2 <- sigma.a2 / (2 * sum(afs.pop * (1 - afs.pop)))) # variance of additive SNP effects
```

### Sample effects

#### Fixed effects

```{r simul_phenos_eff-fix}
beta <- setNames(c(mu, rnorm(n=Q - 1, mean=0, sd=sqrt(sigma.beta2))),
                 colnames(X))
beta
summary((X %*% beta)[,1])
var((X %*% beta)[,1])
```

#### Additive SNP effects

```{r simul_phenos_eff-genos_add}
M <- genomes$genos[uniq.geno.names,]
dim(M)
P <- ncol(M)
afs <- estimSnpAf(M) # afs.pop
cor(afs, afs.pop); # plot(afs, afs.pop[names(afs)])
M.a.c <- M - matrix(rep(1, I)) %*% (2 * afs)
alpha <- setNames(rnorm(n=P, mean=0, sd=sqrt(sigma.alpha2)),
                  colnames(M.a.c))
summary(alpha)
hist(alpha)
```

Update the genetic architecture so that there are few large effects ("major QTLs"):
```{r}
nbMajorQtls <- 3
mafs <- estimSnpMaf(afs=afs)
chrsWithMajorQtls <- sample(unique(genomes$snp.coords$chr), nbMajorQtls)
majorQtls <- c()
for(i in seq_along(chrsWithMajorQtls)){
  mafsChr <- mafs[rownames(genomes$snp.coords[genomes$snp.coords$chr ==
                                              chrsWithMajorQtls[i],])]
  majorQtls <- c(majorQtls,
                 sample(names(mafs)[mafs > 0.15], 1))
}
alpha[majorQtls] <- 10 * sign(alpha[majorQtls])
hist(alpha)
```

#### Dominance SNP effects

```{r simul_phenos_eff-genos_dom, eval=TRUE}
boundaries <- seq(from=0, to=2, length.out=4)
is.0 <- (M <= boundaries[2]) # homozygotes for the first allele
is.1 <- (M > boundaries[2] & M <= boundaries[3]) # heterozygotes
is.2 <- (M > boundaries[3]) # homozygotes for the second allele
M.d.c <- matrix(rep(NA, length(M)), nrow=I, ncol=P, dimnames=dimnames(M))
for(i in 1:I){
  M.d.c[i, is.0[i,]] <- - 2 * afs[is.0[i,]]^2
  M.d.c[i, is.1[i,]] <- 2 * afs[is.1[i,]] * (1 - afs[is.1[i,]])
  M.d.c[i, is.2[i,]] <- - 2 * (1 - afs[is.2[i,]])^2
}
delta <- setNames(rep(0, P),
                  colnames(M.d.c))
## delta <- setNames(rnorm(n=P, mean=0, sd=sqrt(sigma.delta2)),
##                   colnames(M.d.c))
summary(delta)
```

#### Total genotypic values

```{r simul_phenos_genovals}
a <- M.a.c %*% alpha
d <- M.d.c %*% delta
g <- as.vector(a) + as.vector(d) # ignoring epistasis
names(g) <- rownames(M)
summary(g)
var(g)
hist(g)
```

```{r}
for(i in seq_along(majorQtls)){
  print(alpha[majorQtls[i]])
  boxplotCandidateQtl(y=M.a.c %*% alpha, X=M, snp=majorQtls[i], main=majorQtls[i],
                      show.points=TRUE)
}
```

### Genotype-year interactions

TODO

### Spatial heterogeneity

#### Draw many samples

Draw several fields with spatial heterogeneity:
```{r simul_phenos_spatial-het_rand-var}
## rho.rank <- rho.loc <- 0
rho.rank <- 0.8
rho.loc <- 0.8
many.etas <- simulAr1Ar1(n=100, R=nb.locs, C=nb.ranks,
                         rho.r=rho.loc, rho.c=rho.rank,
                         sigma.X.2=0,
                         sigma.e.2=sigma2)
dim(many.etas)
stopifnot(dim(many.etas)[1] == nb.locs)  # rows
stopifnot(dim(many.etas)[2] == nb.ranks) # columns
```

Check that the correlations between ranks and between locations correspond to the input values:
```{r simul_phenos_spatial-het_check}
all.cor <- list()
all.cor$loc <- c(apply(many.etas, 3, function(Xi){
  apply(Xi, 2, function(Xi.col){ # per column, hence cor btw rows
    acf(Xi.col, lag.max=1, type="correlation", plot=FALSE)$acf[2]
  })
}))
all.cor$rank <- c(apply(many.etas, 3, function(Xi){
  apply(Xi, 1, function(Xi.row){ # per row, hen cor btw columns
    acf(Xi.row, lag.max=1, type="correlation", plot=FALSE)$acf[2]
  })
}))
sapply(all.cor, length)
sapply(all.cor, summary) # to be compared to rho.loc and rho.rank
```

#### Choose one

Choose one sample with which phenotypes will be simulated:
```{r simul_phenos_spatial-het_choose-sample}
eta.mat <- many.etas[,,1] # extract matrix
dim(eta.mat)
stopifnot(nrow(eta.mat) == nb.locs)
rownames(eta.mat) <- paste0("location", 1:nrow(eta.mat))
stopifnot(ncol(eta.mat) == nb.ranks)
colnames(eta.mat) <- paste0("rank", 1:ncol(eta.mat))
eta.mat[1:3, 1:3]
```

Check the correlations:
```{r simul_phenos_spatial-het_check-sample}
all.cor <- list()
all.cor$loc <- apply(eta.mat, 2, function(eta.col){ # per column, hence cor btw rows
  acf(eta.col, lag.max=1, type="correlation", plot=FALSE)$acf[2]
})
all.cor$rank <- apply(eta.mat, 1, function(eta.row){ # per row, hence cor btw columns
  acf(eta.row, lag.max=1, type="correlation", plot=FALSE)$acf[2]
})
sapply(all.cor, length)
sapply(all.cor, summary) # to be compared to rho.loc and rho.rank
```

Reformat the matrix into a vector:
```{r simul_phenos_spatial-het_vectorize-sample}
eta <- c(eta.mat)         # vectorize it
stopifnot(length(eta) == ncol(Z.S))
summary(eta)
var(Z.S %*% eta)
names(eta) <- paste0(rep(rownames(eta.mat), times=ncol(eta.mat)),
                     ":", rep(colnames(eta.mat), each=nrow(eta.mat)))
stopifnot(all(names(eta) == colnames(Z.S)))
head(eta)
```

#### Plot it

```{r}
coords.sh <- as.matrix(expand.grid(1:nb.locs, 1:nb.ranks))
apply(coords.sh, 2, range)
colnames(coords.sh) <- c("location", "rank")
dim(coords.sh)
head(coords.sh)
eta.df <- data.frame(eta=eta)
head(eta.df)
sh.sp <- SpatialPointsDataFrame(coords=coords.sh, data=eta.df)
spplot(obj=sh.sp, scales=list(draw=TRUE),
       xlab="ranks", ylab="locations",
       main=paste0("Spatial heterogeneity with cor.rank=", rho.rank,
                   ", cor.loc=", rho.loc, " and sigma2=", round(sigma2)),
       key.space="right", aspect="fill")
```

#### Fit a model on it

```{r}
vg.1 <- gstat::variogram(object=eta ~ 1, data=sh.sp)
plot(x=vg.1$dist, y=vg.1$gamma, main="Empirical variogram",
     ylim=c(0, max(vg.1$gamma)))
(vg.1.fit <- suppressWarnings(
     gstat::fit.variogram(object=vg.1,
                          model=vgm(model="Ste", nugget=NA))))
plot(vg.1, vg.1.fit, main="", col="blue",
     key=list(space="top", lines=list(col="blue"),
              text=list("fit of Matérn model")))
```

### Errors

```{r simul_phenos_errors}
if(all(rho.rank == 0, rho.loc == 0)){
  epsilon <- rnorm(n=N, mean=0, sd=sqrt(sigma2))
} else
  epsilon <- rnorm(n=N, mean=0, sd=0)
summary(epsilon)
var(epsilon)
```

### Phenotypic values

```{r}
stopifnot(all(colnames(X) == colnames(beta)))
y <- X %*% beta
stopifnot(all(colnames(Z) == colnames(g)))
y <- y + Z %*% g
if(any(rho.rank != 0, rho.loc != 0)){
  stopifnot(all(colnames(Z.S) == colnames(eta)))
  y <- y + Z.S %*% eta
}
y <- y + epsilon
y <- as.vector(y)
summary(y)
var(y)
dat$trait1.init <- y
```

```{r}
avg_y <- tapply(dat$trait1.init, dat$geno, mean, na.rm=TRUE)
regplot(g, avg_y[names(g)], pred.int=FALSE)
```

## Missing data

Create missing data depending on the year:
```{r miss_dat}
dat$trait1 <- dat$trait1.init
## max.prop.na <- 0
max.prop.na <- 0.1
(prop.na <- setNames(runif(n=K, min=0, max=max.prop.na),
                     uniq.year.names))
for(year in names(prop.na)){
  if(prop.na[year] == 0)
    next
  set.at.na <- sample(x=c(TRUE, FALSE), replace=TRUE,
                      size=sum(dat$year == year),
                      prob=c(prop.na[year], 1 - prop.na[year]))
  if(any(set.at.na)){
    message(paste0(year, ": ", sum(set.at.na), " NAs"))
    dat$trait1[dat$year == year & set.at.na] <- NA
  }
}
```

Set negative values to NA:
```{r}
is.neg <- (dat$trait1 <= 0)
if(any(is.neg)){
  idx.neg <- which(is.neg)
  message(length(idx.neg))
  dat$trait1[idx.neg] <- NA
}
```

## Outliers

Create upper outliers:
```{r outliers}
idx.notNA <- which(! is.na(dat$trait1))
## prop.out <- 0
prop.out <- 0.002
idx.out <- sample(x=idx.notNA, size=floor(prop.out * length(idx.notNA)))
dat$true.outlier <- FALSE
dat$true.outlier[idx.out] <- TRUE
summary(dat$trait1)
(mean.out <- max(dat$trait1, na.rm=TRUE) + median(dat$trait1, na.rm=TRUE))
(sd.out <- 0.5 * sd(dat$trait1, na.rm=TRUE))
dat$trait1[idx.out] <- rnorm(n=length(idx.out),
                             mean=mean.out,
                             sd=sd.out)
```


# Data exploration

## Tables

### Whole panel

```{r explo_tables}
dim(dat)
lapply(setNames(levels(dat$year), levels(dat$year)), function(lev.y){
  do.call(rbind,
          lapply(setNames(levels(dat$block), levels(dat$block)),
                 function(lev.b){
                   betterSummary(subset(dat, year == lev.y & block == lev.b,
                                        select=trait1)$trait1)
                 }))
})
```

### Controls

```{r explo_tables_ctl}
dat.ctl <- droplevels(dat[dat$geno == control.name,])
dim(dat.ctl)
lapply(setNames(levels(dat.ctl$year), levels(dat.ctl$year)), function(lev.y){
  do.call(rbind,
          lapply(setNames(levels(dat.ctl$block), levels(dat.ctl$block)),
                 function(lev.b){
                   betterSummary(subset(dat.ctl, year == lev.y & block == lev.b,
                                        select=trait1)$trait1)
                 }))
})
```

## Plots

### Sources of variation

Outliers:
```{r}
lower.threshold <- -Inf
upper.threshold <- 1800
x <- "trait1.outlier"
dat[[x]] <- FALSE
is.outlier <- dat[["trait1"]] < lower.threshold |
  dat[["trait1"]] > upper.threshold
sum(is.outlier, na.rm=TRUE)
dat[[x]][is.outlier] <- TRUE
```

```{r explo_plot, fig.width=15}
par(mar=c(7, 4, 4, 2) + 0.1)
beanplot(trait1 ~ fact1 + block + year,
         data=dat, log="",
         xlab="", xaxt="n",
         ylab="phenotypes",
         main="Sources of variation",
         side="both", las=1, what=c(1,1,1,0),
         col=list("black","grey"), border=c("black","grey"),
         ylim=c(-5, 1.3 * max(dat$trait1, na.rm=TRUE)))
labels <- paste0(rep(levels(dat$block), nlevels(dat$year)), " in ",
                 rep(levels(dat$year), each=nlevels(dat$block)))
nb.ticks <- nlevels(dat$block) * nlevels(dat$year)
axis(1, at=1:nb.ticks, labels=FALSE)
text(x=1:nb.ticks,
     y=par("usr")[3] - 0.12 * abs(par("usr")[4] - par("usr")[3]),
     srt=35, adj=0.5,
     labels=labels, xpd=TRUE)
mtext(1, text="blocks and years", line=5)
abline(v=seq(from=nlevels(dat$block) + 0.5, to=nb.ticks,
             by=nlevels(dat$block)))
legend("topleft", legend=paste0("fact1: ", levels(dat$fact1)),
       fill=c("black","grey"), border=NA, bty="n")
lgd <- c()
lgd.lty <- c()
lgd.col <- c()
abline(h=c(lower.threshold, upper.threshold), lty=2)
if(any(dat[["trait1.outlier"]])){
  lgd <- c(lgd, "outlier threshold")
  lgd.lty <- c(lgd.lty, 2)
  lgd.col <- c(lgd.col, "black")
}
q1.ctl <- quantile(dat$trait1, probs=0.25, na.rm=TRUE)
q3.ctl <- quantile(dat$trait1, probs=0.75, na.rm=TRUE)
abline(h=c(q1.ctl, q3.ctl), col="red")
lgd <- c(lgd, "controls (Q1, Q3)")
lgd.lty <- c(lgd.lty, 1)
lgd.col <- c(lgd.col, "red")
legend("topright", legend=lgd, lty=lgd.lty, col=lgd.col, bty="n")
```

### Missing data

```{r}
for(trait in c("trait1")){
  for(year in levels(dat$year)){
    lay.tmp <- lay
    lay.tmp@data[["missing"]] <- "false"
    lay.tmp@data[["missing"]][is.na(dat[dat$year == year, trait])] <- "true"
    lay.tmp@data[["missing"]] <- as.factor(lay.tmp@data[["missing"]])
    print(spplot(obj=lay.tmp, zcol="missing", col.regions=c("green","red"),
                 scales=list(draw=TRUE),
                 xlab="ranks", ylab="locations",
                 main=paste0("Missing data for ", trait, " in ", year),
                 key.space="right", aspect="fill"))
  }
}
```


# First phase of the analysis

## Data preparation

```{r}
str(dat)
```

```{r}
trait <- "trait1"
transfo <- "id"
stopifnot(transfo %in% c("id", "log", "sqrt"))
```

```{r}
if(any(dat[[paste0(trait, ".outlier")]])){
  dat <- droplevels(dat[! dat[[paste0(trait, ".outlier")]],])
  print(str(dat))
}
```

```{r}
if(transfo == "id"){
  response <- trait
} else if(transfo == "log"){
  response <- paste0("log(", trait, ")")
  dat[[response]] <- log(dat[[trait]])
} else if(transfo == "sqrt"){
  response <- paste0("sqrt(", trait, ")")
  dat[[response]] <- sqrt(dat[[trait]])
}
```

## Model fitting, comparison and selection

Approach:

1. model fit, comparison and selection on control data,

2. correct spatial heterogeneity via controls,

3. model fit, comparison and selection on panel data.

If spatial heterogeneity is ignored, only step 3 is performed.

```{r model_fit_comp_sel}
glob.form <- paste0(response, " ~ 1",
                    " + block",
                    " + year",
                    " + covar1",
                    " + covar2",
                    ## " + covar1:covar2",
                    " + fact1",
                    " + fact2",
                    " + fact1:fact2")#,
                    ## " + covar1:fact1",
                    ## " + covar1:fact2",
                    ## " + covar2:fact1",
                    ## " + covar2:fact2")
out <- plantTrialLmmFitCompSel(glob.form=glob.form,
                               dat=dat[dat$control,
                                       c(response,"block","year",
                                         "covar1","covar2","fact1","fact2")],
                               part.comp.sel="fixed",
                               saved.file="two-phase_quantgen_ctl_allmod-sel-ML.RData",
                               nb.cores=nb.cores, verbose=1)
```

## Correct spatial heterogeneity

Kriging using controls:
```{r kriging}
fix.preds.corr.spat <- out$all.preds
idx <- grep("\\(*\\|*\\)|year", out$all.preds)
if(length(idx) > 0)
  fix.preds.corr.spat <- out$all.preds[-idx]
dat <- correctSpatialHeterogeneity(dat=dat,
                                   response=response,
                                   fix.eff=fix.preds.corr.spat,
                                   verbose=2)
```

Re-fit the best model using the panel data corrected for spatial heterogeneity:
```{r bestmod_refit_spatial_correct}
corr.spat.het <- TRUE
if(corr.spat.het){
  glob.form <- gsub(response, paste0(response, ".csh"), glob.form)
  response <- paste0(response, ".csh")
}
glob.form <- paste0(glob.form,
                    " + (1|geno)",
                    " + (1|geno:year)")
out <- plantTrialLmmFitCompSel(glob.form=glob.form,
                               dat=dat[! dat$control,
                                       c(response,"block","year",
                                         "covar1","covar2","fact1","fact2",
                                         "geno","rank","location")],
                               part.comp.sel=c("fixed","random"),
                               saved.file="two-phase_quantgen_panel_allmod-sel-ML.RData",
                               nb.cores=nb.cores, verbose=1)
dat.noNA <- out$dat.noNA
bestmod.ml <- out$bestmod.ml
bestmod.reml <- out$bestmod.reml
```

## Best model fit to get genotypic BLUEs and EMMs

Re-fit, modeling genotypic values as fixed effects to get their BLUEs:
```{r bestmod_refit_BLUEs}
best.form.as.fix <- paste0(response, " ~ 1 + ",
                           paste(out$best.preds.fix, collapse=" + "),
                           " + ", paste(out$best.preds.rnd, collapse=" + "))
bestmod.lm <- lm(formula=best.form.as.fix,
                 data=dat.noNA,
                 na.action=na.fail)
```

Re-fit, modeling genotypic values as fixed effects to get their EMMs:
```{r bestmod_refit_EMMs}
bestmod.emm <- emmeans(bestmod.lm, "geno")
```


## Assumption checking

### Residual preparation

```{r prep_res}
fit.all <- cbind(out$dat,
                 response=out$dat[[response]],
                 cond.res=NA,
                 scl.cond.res=NA,
                 fitted=NA)
idx.NA <- attr(dat.noNA, "na.action")
if(length(idx.NA) > 0){
  fit.all$cond.res[- idx.NA] <- residuals(bestmod.reml)
  fit.all$scl.cond.res[- idx.NA] <- residuals(bestmod.reml) /
    sigma(bestmod.reml)
  fit.all$fitted[- idx.NA] <- fitted(bestmod.reml)
} else{
  fit.all$cond.res <- residuals(bestmod.reml)
  fit.all$scl.cond.res <- residuals(bestmod.reml) /
    sigma(bestmod.reml)
  fit.all$fitted <- fitted(bestmod.reml)
}
geno.blups <- ranef(bestmod.reml, condVar=TRUE, drop=TRUE)$geno
geno.var.blups <- setNames(attr(geno.blups, "postVar"),
                           names(geno.blups))
```

TODO:
```{r}
## geno.blues
## geno.emmeans
```

### Error homoscedasticity

```{r check_error_homoscedasticity}
x.lim <- max(abs(fit.all$scl.cond.res), na.rm=TRUE)
plot(x=fit.all$scl.cond.res, y=fit.all$fitted, las=1,
     xlim=c(-x.lim, x.lim),
     xlab="scaled conditional residuals",
     ylab="fitted responses",
     main=response)
abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
```

```{r check_error_homoscedasticity_year_block}
lattice::xyplot(fitted ~ scl.cond.res | year, groups=block,
                data=fit.all,
                xlab="scaled conditional residuals",
                ylab="fitted responses",
                main=response,
                auto.key=list(space="right"),
                panel=function(x,y,...){
                  panel.abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
                  panel.xyplot(x,y,...)
                })
```

```{r}
boxplot(scl.cond.res ~ block + year,
        data=fit.all,
        xlab="scaled conditional residuals",
        ## ylab="blocks and years",
        main=response,
        varwidth=TRUE, notch=TRUE, horizontal=TRUE, las=1)
abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
```

### Error normality

```{r check_error_normality}
shapiro.test(fit.all$scl.cond.res)
qqnorm(y=fit.all$scl.cond.res,
       main=paste0(response, ": scaled conditional residuals"))
qqline(y=fit.all$scl.cond.res, col="red")
```

### Temporal independence of errors

```{r check_error_temp_indep_plot}
for(block in levels(fit.all$block))
  plotResidualsBtwYears(df=fit.all, colname.res="scl.cond.res",
                        years=c("2011","2012"), blocks=rep(block,2),
                        cols.uniq.id=c("geno","block","rank","location"),
                        main=paste0("Block ", block))
plotResidualsBtwYears(df=fit.all, colname.res="scl.cond.res",
                      years=c("2011","2012"), blocks=c("A","B"),
                      cols.uniq.id="geno",
                      main="Blocks A and B")
plotResidualsBtwYears(df=fit.all, colname.res="scl.cond.res",
                      years=c("2011","2012"), blocks=c("A","C"),
                      cols.uniq.id="geno",
                      main="Blocks A and C")
```

### Spatial independence of errors

```{r check_error_spatial_indep_plot, eval=TRUE}
lay.tmp <- lay
for(year in levels(fit.all$year)){
  lay.tmp[[paste0("scl.cond.res.", year)]] <- NA
  for(i in 1:nrow(lay.tmp)){
    j <- which(fit.all$block == lay.tmp$block[i] &
               fit.all$rank == lay.tmp$rank[i] &
               fit.all$location == lay.tmp$location[i] &
               fit.all$year == year)
    if(length(j) == 0){
      next
    } else if(length(j) == 1){
      lay.tmp[[paste0("scl.cond.res.", year)]][i] <- fit.all$scl.cond.res[j]
    } else
      warning(paste0("year=", year, "i=", i))
  }
}
for(year in levels(fit.all$year)){
  out <- spplot(obj=lay.tmp, zcol=paste0("scl.cond.res.", year),
                xlab=colnames(coordinates(lay.tmp))[1],
                ylab=colnames(coordinates(lay.tmp))[2],
                main=paste0(response, ": scaled conditional residuals in ", year),
                key.space="right", scales=list(draw=TRUE), aspect="fill")
  print(out)
}
```

Look at the empirical variogram, and fit a variogram model:
```{r check_error_spatial_indep_variogram, eval=TRUE}
for(year in levels(fit.all$year)){
  tmp <- na.omit(as.data.frame(lay.tmp[,paste0("scl.cond.res.", year)]))
  resid.vg <- variogram(object=formula(paste0("scl.cond.res.",
                                              year, " ~ 1")),
                        locations=~ rank + location, data=tmp)
  ## print(plot(resid.vg,
  ##            main=paste0(response, ": empirical variogram in ", year)))
  resid.vg.fit <- fit.variogram(object=resid.vg,
                                model=vgm(c("Exp", "Sph", "Gau", "Ste")),
                                fit.sills=TRUE,
                                fit.ranges=TRUE,
                                fit.kappa=seq(0.3, 5, 0.05))
  print(resid.vg.fit)
  print(plot(resid.vg, resid.vg.fit, main="", col="blue",
             key=list(space="top", lines=list(col="blue"),
                      text=list(paste0(response, ": fit of variogram model",
                                       " (", resid.vg.fit[2, "model"], ")",
                                       " in ", year)))))
}
```

### Outlying genotypes

```{r check_outlying_genos_plot}
x.lim <- max(abs(geno.blups))
par(mar=c(5,6,4,1))
plot(x=geno.blups, y=1:length(geno.blups),
     xlim=c(-x.lim, x.lim),
     xlab="genotypic BLUPs",
     main=response,
     yaxt="n", ylab="")
axis(side=2, at=1:length(geno.blups), labels=names(geno.blups), las=1)
abline(v=0, lty=2)
idx <- which.max(geno.blups)
text(x=geno.blups[idx], y=idx, labels=names(idx), pos=2)
idx <- which.min(geno.blups)
text(x=geno.blups[idx], y=idx, labels=names(idx), pos=4)
```

```{r check_var_geno_BLUPs}
summary(geno.var.blups)
head(sort(geno.var.blups))
tail(sort(geno.var.blups))
```

### Normality of genotypic BLUPs

```{r check_normality_geno_BLUPs_qqplot}
shapiro.test(geno.blups)
qqnorm(y=geno.blups, main=paste0(response, ": genotypic BLUPs"), asp=1)
qqline(y=geno.blups, col="red")
```

### Independence between errors and genotypes

```{r check_indep_error_geno_plot}
x.lim <- max(abs(fit.all$scl.cond.res))
lattice::dotplot(geno ~ scl.cond.res, data=fit.all,
                 xlim=c(-x.lim, x.lim),
                 xlab="scaled conditional residuals",
                 main=response,
                 panel=function(x,y,...){
                   panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                   panel.dotplot(x,y,...)
                 })
```

```{r check_indep_error_geno_plot_year_block}
lattice::dotplot(geno ~ scl.cond.res | year, groups=block,
                 data=fit.all,
                 auto.key=list(space="right"),
                 xlab="scaled conditional residuals",
                 main=response,
                 panel=function(x,y,...){
                   panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                   panel.dotplot(x,y,...)
                 })
```

## Model outputs

### Summary

```{r bestmod_infer}
(vc.ml <- as.data.frame(VarCorr(bestmod.ml)))
summary(bestmod.reml)
(vc.reml <- as.data.frame(VarCorr(bestmod.reml)))
```

```{r}
if(FALSE){
  summary(bestmod.lm)
  summary(bestmod.emm)
}
```

### Broad-sense heritability

```{r H2}
tmp <- estimH2means(dat=dat.noNA, colname.resp=response,
                    colname.trial="year", vc=vc.reml,
                    geno.var.blups=geno.var.blups)
tmp$mean.nb.trials
tmp$mean.nb.reps.per.trial
tmp$H2.classic
tmp$H2.oakey
mySumm <- tmp$sryStat
```

### Confidence intervals

#### Profiling

```{r bestmod_confint_prof, eval=TRUE}
if(FALSE){
  st <- system.time(
      prof <- profile(bestmod.ml, signames=FALSE,
                      parallel="multicore", ncpus=nb.cores))
  print(st)
  (ci <- confint(prof, level=0.95))
  xyplot(prof, absVal=TRUE, conf=c(0.8, 0.95), which=1:2,
         main=response)
  densityplot(prof,
              main=response)
  splom(prof, conf=c(0.8, 0.95), which=1:2,
        main=response)
}
```

#### Bootstrapping

```{r bestmod_confint_boot, eval=TRUE}
mySumm(bestmod.reml)
mySumm(bestmod.ml)
if(FALSE){
  st <- system.time(
      fit.boot <- bootMer(x=bestmod.ml, FUN=mySumm,
                          nsim=1*10^3, seed=1859,
                          use.u=FALSE, type="parametric",
                          parallel="multicore", ncpus=nb.cores))
  print(st)
  print(fit.boot)
  for(i in seq_along(fit.boot$t0)){
    message(names(fit.boot$t0)[i])
    plot(fit.boot, index=i,
         main=paste0(response, ": ", names(fit.boot$t0)[i]))
    print(boot.ci(fit.boot, conf=c(0.8, 0.95),
                  type=c("norm", "basic", "perc"),
                  index=i))
  }
}
```

### Hypothesis testing

### Fixed effects

```{r bestmod_lmerTest}
anova(bestmod.reml)
## anova(bestmod.reml.lmerTest, ddf="lme4")
## anova(bestmod.reml.lmerTest, ddf="Satterthwaite")
## system.time(
##     print(anova(bestmod.reml.lmerTest, ddf="Kenward-Roger")))
```

### Variance components

```{r bestmod_lmerTest_varcomp}
ranova(bestmod.reml)
```

### In-sample prediction

```{r bestmod_in-sample_prediction}
cor(fit.all$fitted, fit.all$response, use="complete.obs")
plot(fit.all$fitted, fit.all$response, las=1, asp=1,
     xlab="observed response",
     ylab="fitted responses",
     main=response)
abline(a=0, b=1, lty=2)
abline(v=mean(fit.all$fitted, na.rm=TRUE), lty=2)
abline(h=mean(fit.all$response, na.rm=TRUE), lty=2)
```

```{r check_in-sample_prediction_year_block}
tapply(1:nrow(fit.all), fit.all$year, function(idx){
  cor(fit.all$fitted[idx], fit.all$response[idx], use="complete.obs")
})
tapply(1:nrow(fit.all), list(fit.all$year, fit.all$block), function(idx){
  cor(fit.all$fitted[idx], fit.all$response[idx], use="complete.obs")
})
lattice::xyplot(response ~ fitted | year, groups=block,
                data=fit.all,
                auto.key=list(space="right"),
                xlab="observed response",
                ylab="fitted responses",
                main=response,
                panel=function(x,y,...){
                  panel.abline(a=0, b=1, lty=2)
                  panel.abline(v=mean(x, na.rm=TRUE), lty=2)
                  panel.abline(h=mean(y, na.rm=TRUE), lty=2)
                  panel.xyplot(x,y,...)
                })
```

## Comparison with true values

```{r}
regplot(x=geno.blups, y=g[names(geno.blups)],
        xlab="eBLUPs of genotypic values",
        ylab="true genotypic values",
        main="Simulated data")
```



# Second phase of the analysis

## Reformatting

```{r}
matSnps <- M[names(geno.blups),]
afs <- estimSnpAf(matSnps) # afs.pop
cor(afs, afs.pop); # plot(afs, afs.pop[names(afs)])
matAdd <- estimGenRel(X=matSnps, afs=afs, method="vanraden1")
matSnps_ok <- matSnps[, estimSnpMaf(afs=afs) > 0.05]
ncol(matSnps_ok)
snp.coords_ok <- genomes$snp.coords[colnames(matSnps_ok),]
afs.ok <- afs[colnames(matSnps_ok)]
matSnps_ok_ctr <- matSnps_ok - matrix(rep(1,nrow(matSnps_ok))) %*% (2 * afs.ok)
head(matSnps_ok_ctr[,majorQtls])
head(M.a.c[rownames(matSnps_ok_ctr),majorQtls])
```

## univariate, SNP-by-SNP (MM4LMM)

### Fit

```{r}
p2f <- "two-phase_quantgen_MM4LMM.rds"
if(! file.exists(p2f)){
  st <- system.time(
      ## out_mm4lmm_1 <- MMEst(Y=g[names(geno.blups)], X=matSnps_ok_ctr,
      out_mm4lmm_1 <- MMEst(Y=geno.blups, X=matSnps_ok,
                            VarList=list(Additive=matAdd,
                                         Error=diag(length(geno.blups)))))
  print(st)
  saveRDS(out_mm4lmm_1, p2f)
  print(tools::md5sum(path.expand(p2f)))
} else{
  print(tools::md5sum(path.expand(p2f)))
  load(p2f)
}
out_mm4lmm_2 <- AnovaTest(out_mm4lmm_1, Type="TypeI")
```

### Reformat

```{r}
out_mm4lmm_2[[1]]
(snpError <- which(sapply(out_mm4lmm_2, function(x){
  if(! "Xeffect" %in% rownames(x)) TRUE else FALSE
})))
pvals_mm4lmm <- sapply(out_mm4lmm_2, function(x){
  if("Xeffect" %in% rownames(x))  x["Xeffect","pval"] else NA
})
pvals <- cbind(snp.coords_ok[names(pvals_mm4lmm),], pvals_mm4lmm)
colnames(pvals) <- c("Chrom", "Position", "p.val")
pvals$p.val <- -log10(pvals$p.val)
```

### Plot

```{r}
mat <- matrix(c(1,1,2,3), nrow=2, ncol=2, byrow=T)
layout(mat, widths = c(4,4), heights = c(3,3))

sommer::manhattan(pvals, pch=20, cex=1)

plotHistPval(10^-pvals$p.val, main="", ylim=c(0,1.5), legend.x=NULL)

col <- setNames(rep("black", nrow(pvals)), rownames(pvals))
col[majorQtls] <- "red"
pch <- setNames(rep(1, nrow(pvals)), rownames(pvals))
pch[majorQtls] <- 20
tmp <- qqplotPval(10^-pvals$p.val, main="", col=col, pch=pch,
                  ctl.fdr.bh=TRUE, verbose=0)
lim.pv.bh <- attr(tmp, "lim.pv.bh")
segments(x0=par("usr")[1], y0=-log10(lim.pv.bh),
         x1=-log10(lim.pv.bh), y1=-log10(lim.pv.bh),
         lty=3)

10^-pvals[majorQtls,"p.val"]
```

### univariate, SNPs jointly (MLMMGWAS)

TODO

### uniavriate, SNPs jointly (varbvs)

TODO

### multivariate, SNP-by-SNP

TODO

### multivariate, SNPs jointly

TODO


# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```
