---
title: "Introduction to heteroscedasticity in regression"
author: "Timothée Flutre (INRA)"
date: "`r format(Sys.time(), '%d/%m/%Y %H:%M:%S')`"
output:
  rmarkdown::html_vignette:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{Intro heteroscedasticity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
setwd("~/src/rutilstimflutre/vignettes/")

library(rmarkdown)
render("intro-heteroscedasticity.Rmd", "html_document")

tools::buildVignette("intro-heteroscedasticity.Rmd")

library(devtools)
build_vignettes()
-->

```{r, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=6, fig.height=5, fig.align="center",
                      global.par=TRUE)
par(mar=c(5, 4, 2, 0.5))
```


# Preamble

License: [CC BY-SA 4.0](http://creativecommons.org/licenses/by-sa/4.0/)

References:

* [Pinheiro & Bates (2000)](http://dx.doi.org/10.1007/b98882): "Mixed-effects models in S and S-PLUS"

* [Galecki and Burzykowski (2013)](dx.doi.org/10.1007/978-1-4614-3900-4): "Linear mixed-effects models using R: a step-by-step approach"

External packages:

```{r load_pkg}
library(rutilstimflutre)
library(nlme)
library(lme4)
library(lattice)
```

This R chunk is used to assess how much time it takes to execute the R code in this document until the end:
```{r time_0}
t0 <- proc.time()
```


# Generative model

Goal: study the genotype-phenotype map

Notations:

* $g \in \{1,\ldots,G\}$: index of the genotype (assuming a plant species whose genotypes can be vegetatively multipled, hence cloned and planted in multiple blocks)

* $b \in \{1,\ldots,B\}$: index of the block (field plots often are divided into multiple blocks to handle spatial heterogeneity)

* $y_{gb}$: observed value (phenotype) of the $g$-th genotype in the $b$-th block

* $\beta_b$: fixed effect of the $b$-th block

* $\mu$: global intercept

* $u_g$: random effect of the $g$-th genotype

Likelihood:

\[
y_{gb} = \mu + \beta_b + u_g + \epsilon_{gb} \text{ where } \epsilon_{gb} \sim \mathcal{N}(0, \sigma_b^2) \text{ and } u_g \sim \mathcal{N}(0, \sigma_u^2)
\]

This model is heteroscedastic as the errors have different variances depending on the block, e.g. $\sigma_{b=1}$ may be different from $\sigma_{b=2}$.


# Data simulation

```{r}
set.seed(1859)
B <- 5
G <- 10^3
dat <- data.frame(block=rep(LETTERS[1:B], each=G),
                  geno=rep(paste0("g", 1:G), B),
                  y=NA)
X <- model.matrix(~ 1 + block, data=dat)
mu <- 50
beta <- c(mu, rnorm(n=B-1, mean=0, sd=3))
Z <- model.matrix(~ -1 + geno, data=dat)
sigma.u <- 2
u <- rnorm(n=G, mean=0, sd=sigma.u)
(sigma.b <- setNames(sample(1:B), levels(dat$block)))
epsilon <- do.call(c, lapply(sigma.b, function(sd.b){
  rnorm(n=G, mean=0, sd=sd.b)
}))
dat$y <- X %*% beta + Z %*% u + epsilon
```


# Data exploration

Descriptive stats:
```{r}
(tmp <- do.call(rbind, tapply(dat$y, dat$block, function(y.b){
  c(mean=mean(y.b), sd=sd(y.b))
})))
```

Visualization:
```{r}
boxplot(y ~ block, data=dat, las=1, varwidth=TRUE,
        xlab="blocks", ylab="phenotypes", main="Simulated data")
abline(h=mean(dat$y), lty=2)
```


# Model fit and diagnostics

For some fitting functions, before using them, we need to reorder the data according to block variances:
```{r}
(new.order <- rownames(tmp)[order(tmp[,"sd"])])
idx <- do.call(c, lapply(new.order, function(b){
  which(dat$block == b)
}))
dat2 <- dat[idx,]
```

## With `gls` ignoring genotype effects

The `gls` function from the `nlme` package, performing inference via generalized least squares, can't handle random effects.

```{r}
fit.gls <- nlme::gls(model=y ~ 1 + block, data=dat2,
                     weights=varIdent(form=~1|block))
beta.hat <- coef(fit.gls)
cbind(beta, beta.hat)
sigma.b.hat <- c(1, exp(coef(fit.gls$modelStruct$varStruct)))
cbind(sigma.b[new.order], sigma.b.hat)
summary(fit.gls)
scaled.residuals <- residuals(fit.gls) / rep(sigma.b.hat, each=G)
plot(x=scaled.residuals, main="Diagnostics gls",
     y=jitter(fitted(fit.gls)), las=1)
abline(v=c(-2,0,2), lty=2)
tmp <- cbind(dat2, scaled.residuals)
boxplot(scaled.residuals ~ block, data=dat2, las=1, varwidth=TRUE,
        xlab="scaled residuals", ylab="blocks", horizontal=TRUE,
        main="Diagnostics gls")
abline(v=c(-2,0,2), lty=2)
lattice::dotplot(geno ~ scaled.residuals, data=tmp,
                 main="Diagnostics gls",
                 panel=function(x,y,...){
                   panel.dotplot(x,y,...)
                   panel.abline(v=c(-2,0,2), lty=2)
                 })
```

The residuals per block look fine, which is expected as the heteroscedasticity was taken into account.

The residuals per genotype look overdispersed, which is expected as the genotypic variance wasn't taken into account.


## With `lmer` assuming homoscedasticity

The `lmer` function from the `lme4` package, performing inference via ReML, can't handle heteroscedasticity.

```{r}
fit.lmer.hom <- lme4::lmer(formula=y ~ 1 + block + (1|geno), data=dat2)
beta.hat <- fixef(fit.lmer.hom)
cbind(beta, beta.hat)
sigma(fit.lmer.hom)
summary(fit.lmer.hom)
scaled.residuals <- residuals(fit.lmer.hom) / sigma(fit.lmer.hom)
plot(x=scaled.residuals, main="Diagnostics lmer",
     y=fitted(fit.lmer.hom), las=1)
abline(v=c(-2,0,2), lty=2)
tmp <- cbind(dat2, scaled.residuals)
boxplot(scaled.residuals ~ block, data=dat2, las=1, varwidth=TRUE,
        xlab="scaled residuals", ylab="blocks", horizontal=TRUE,
         main="Diagnostics lmer")
abline(v=c(-2,0,2), lty=2)
lattice::dotplot(geno ~ scaled.residuals, data=tmp,
                  main="Diagnostics lmer",
                 panel=function(x,y,...){
                   panel.dotplot(x,y,...)
                   panel.abline(v=c(-2,0,2), lty=2)
                 })
```

The residuals clearly have different variances depending on the blocks, which is expected as heteroscedasticity wasn't taken into account.

The residuals per genotype look fine, as expected thanks to the genotypic variance component.


## With `lme` assuming heteroscedasticity

The `lme` function from the `lme4` package, performing inference via ReML, can handle both random effects and heteroscedasticity.

```{r, fig.width=12}
fit.lme.het <- nlme::lme(fixed=y ~ 1 + block, random=~1|geno,
                         weights=varIdent(form=~1|block),
                         data=dat2)
beta.hat <- fixef(fit.lme.het)
cbind(beta, beta.hat)
sigma.b.hat <- c(1, exp(coef(fit.lme.het$modelStruct$varStruct)))
cbind(sigma.b[new.order], sigma.b.hat)
summary(fit.lme.het)
scaled.residuals <- residuals(fit.lme.het) / rep(sigma.b.hat, each=G)
plot(x=scaled.residuals, main="Diagnostics lme",
     y=fitted(fit.lme.het), las=1)
abline(v=c(-2,0,2), lty=2)
tmp <- cbind(dat2, scaled.residuals)
boxplot(scaled.residuals ~ block, data=dat2, las=1, varwidth=TRUE,
        xlab="scaled residuals", ylab="blocks", horizontal=TRUE,
        main="Diagnostics lme")
abline(v=c(-2,0,2), lty=2)
lattice::dotplot(geno ~ scaled.residuals, data=tmp,
                 main="Diagnostics lme",
                 panel=function(x,y,...){
                   panel.dotplot(x,y,...)
                   panel.abline(v=c(-2,0,2), lty=2)
                 })
```

Residuals per block and per genotype look fine, as expected (except maybe for block "A", as well as a few genotype outliers?).


# More flexible alternative

The `lme4` and `nlme` packages used above are powerful, but not as flexible as other packages when it comes to deal with missing responses or fat-tail error distribution.
This is partly due to the fact that they implement the [ReML algorithm](https://en.wikipedia.org/wiki/Restricted_maximum_likelihood) only for a specific subset of all possible regression models.
An alternative is to choose the Bayesian paradigm, and perform inference via the [Gibbs simpler](https://en.wikipedia.org/wiki/Gibbs_sampling) (with `rjags`) or the [HMC algorithm](https://en.wikipedia.org/wiki/Hybrid_Monte_Carlo) (with `rstan`).

TODO: try `rstan`


# Appendix

```{r info}
t1 <- proc.time()
t1 - t0
print(sessionInfo(), locale=FALSE)
```
