---
title: "Introduction to Bayesian statistics via the univariate Normal model with unknown mean and precision"
author: "Timothée Flutre (INRA)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: TRUE
  pdf_document:
    toc: true
    toc_depth: 4
    number_sections: TRUE
vignette: >
  %\VignetteIndexEntry{Intro Bayes (univ Normal, unknwo mu and tau)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
setwd("~/src/rutilstimflutre/vignettes/")

library(devtools)
build_vignettes()

library(rmarkdown)
render("intro-bayes_univ-normal_unknown-mean-precision.Rmd", "html_document")
-->

# Preamble

License: [CC BY-SA 4.0](http://creativecommons.org/licenses/by-sa/4.0/)

References:

* ["Optimal statistical decisions", DeGroot (1970)](http://dx.doi.org/10.1002/0471729000), theorem 1 in section 9.6 on page 169

* ["A first course in Bayesian statistical methods", Hoff (2009)](http://dx.doi.org/10.1007/978-0-387-92407-6), section 5.3 starting on page 73


# Model

## Parametrization

In this document, the Normal distribution, $\mathcal{N}$, is parametrized in terms of its mean and variance.
Nevertheless, equations are easier to work with in terms of precision, which is the inverse of the variance.

The Gamma distribution, $\mathcal{G}$, is parametrized in terms of shape and rate.


## Likelihood

* data, $\mathcal{D}$:

\[
\{y_1, \ldots, y_N \} = \{ \boldsymbol{y} \}
\]

* parameters, $\Theta$:

\[
\{ \mu, \tau \}
\]

* likelihood, $\mathcal{L}(\Theta) = p(\mathcal{D} | \Theta)$:

\[
\forall n \in \{1,\ldots,N\}, \; y_n | \mu, \tau \; \sim \; \mathcal{N}(\mu, 1/\tau)
\]


## Priors

$p(\Theta) = p(\mu, \tau) = p(\tau) \times p(\mu | \tau)$

* $\tau \sim \mathcal{G}(a_0, b_0)$

* $\mu | \tau \sim \mathcal{N}(m_0, 1 / (\tau t_0))$

Hyper-parameters:

* shape $a_0 > 0$ and rate $b_0 > 0$

* mean $m_0 \in \mathbb{R}$ and precision $t_0 > 0$


## Posteriors

$p(\Theta | \mathcal{D}) \propto p(\Theta) \times p(\mathcal{D} | \Theta)$

* $\tau | \boldsymbol{y} \sim \mathcal{G}(a_N, b_N)$

* $\mu | \boldsymbol{y}, \tau \sim \mathcal{N}(m_N, 1 / t_N)$

with:

$a_N = a_0 + N/2$

$b_N = b_0 + \frac{1}{2} \sum_n (y_n - \bar{y})^2 + \frac{t_0 N (\bar{y} - m_0)^2}{2 (t_0 + N)}$

$m_N = \frac{t_0 m_0 + N \bar{y}}{t_0 + N}$

$t_N = \tau (t_0 + N)$


## Details

TODO

Bayesian inference uses posterior distributions: $p(\Theta | \mathcal{D}) = p(\mu, \tau | \boldsymbol{y}) = p(\tau | \boldsymbol{y}) \, p(\mu | \boldsymbol{y}, \tau)$

Let us start with the conditional posterior of $\mu$:

\begin{align*}
p(\mu | \boldsymbol{y}, \tau) &\propto p(\mu | \tau) \, p(\boldsymbol{y} | \mu, \tau) \\
\end{align*}


# Simulation

```{r setup_simul}
set.seed(7263)
truth <- c()
```

## $\tau$

```{r simul_tau}
a.0 <- 3
truth <- append(x=truth,
                values=setNames(object=a.0, nm="a.0"))
b.0 <- 2
truth <- append(x=truth,
                values=setNames(object=b.0, nm="b.0"))
truth <- append(x=truth,
                values=setNames(object=a.0 / b.0, nm="prior.mean.tau"))
truth <- append(x=truth,
                values=setNames(object=a.0 / b.0^2, nm="prior.var.tau"))
plot(x=seq(0,11,0.1),
     y=dgamma(x=seq(0,11,0.1), shape=a.0, rate=b.0),
     main=paste0("Gamma(shape=", a.0, ", rate=", b.0, ")"),
     xlab="x", ylab="pdf", type="l")
(tau <- rgamma(n=1, shape=a.0, rate=b.0))
truth <- append(x=truth,
                values=setNames(object=tau, nm="tau"))
truth <- append(x=truth,
                values=setNames(object=1 / tau, nm="sigma2"))
truth <- append(x=truth,
                values=setNames(object=sqrt(1 / tau), nm="sigma"))
```

## $\mu$

```{r simul_mu}
m.0 <- 73
truth <- append(x=truth,
                values=setNames(object=m.0, nm="prior.mean.mu"))
t.0 <- 10^(-1)
truth <- append(x=truth,
                values=setNames(object=1 / (tau * t.0), nm="prior.var.mu"))
plot(x=seq(60,86,0.1),
     y=dnorm(x=seq(60,86,0.1), mean=m.0, sd=sqrt(1/(tau*t.0))),
     main=paste0("Normal(mean=", m.0, ", sd=",
                 format(sqrt(1/(tau*t.0)), digits=3), ")"),
     xlab="x", ylab="pdf", type="l")
(mu <- rnorm(n=1, mean=m.0, sd=sqrt(1/(tau * t.0))))
truth <- append(x=truth,
                values=setNames(object=mu, nm="mu"))
```

## Data

```{r simul_data}
N <- 5*10^1
y <- rnorm(n=N, mean=mu, sd=sqrt(1/tau))
summary(y)
hist(y, breaks="FD", freq=FALSE, col="grey", border="white",
     main="Simulated data",
     ylim=c(0, dnorm(x=mu, mean=mu, sd=sqrt(1/tau))))
abline(v=mu, col="red")
curve(dnorm(x=x, mean=mu, sd=sqrt(1/tau)), add=TRUE, col="red")
```


# Inference

## Maximum likelihood estimates

```{r infer_mle}
truth <- append(x=truth,
                values=setNames(object=mean(y), nm="mle.mu"))
truth <- append(x=truth,
                values=setNames(object=1 / var(y), nm="mle.tau"))
truth <- append(x=truth,
                values=setNames(object=var(y), nm="mle.sigma2"))
truth <- append(x=truth,
                values=setNames(object=sd(y), nm="mle.sigma"))
truth[grep("mle", names(truth))]
```


## Posteriors

```{r posteriors}
a.N <- a.0 + N / 2
truth <- append(x=truth,
                values=setNames(object=a.N, nm="a.N"))
b.N <- b.0 + (1/2) * sum((y - mean(y))^2) +
  (t.0 * N * (mean(y) - m.0)^2) / (2 * (t.0 + N))
truth <- append(x=truth,
                values=setNames(object=b.N, nm="b.N"))
truth <- append(x=truth,
                values=setNames(object=a.N / b.N, nm="post.mean.tau"))
truth <- append(x=truth,
                values=setNames(object=a.N / b.N^2, nm="post.var.tau"))
truth <- append(x=truth,
                values=setNames(object=b.N / (a.N - 1), nm="post.mean.sigma2"))
truth <- append(x=truth,
                values=setNames(object=b.N^2 / ((a.N - 1)^2 * (a.N - 2)),
                                nm="post.var.sigma2"))
m.N <- (t.0 * m.0 + N * mean(y)) / (t.0 + N)
truth <- append(x=truth,
                values=setNames(object=m.N, nm="post.mean.mu.cond"))
t.N <- (a.N / b.N) * (t.0 + N)
truth <- append(x=truth,
                values=setNames(object=1 / t.N, nm="post.var.mu.cond"))
```


# Evaluation

```{r truth}
truth
```

```{r infer_graphics, fig.width=7, fig.height=5}
plot(x=0, y=0, type="n",
     xlim=c(truth["prior.mean.mu"] - 2 * sqrt(truth["prior.var.mu"]),
            truth["prior.mean.mu"] + 2 * sqrt(truth["prior.var.mu"])),
     ylim=c(0,2),
     xlab="x", ylab="density",
     main="Bayesian inference of a Normal with unknown mean and variance")
abline(v=truth["mu"], lwd=2, lty=2)
curve(dnorm(x=x, mean=truth["prior.mean.mu"], sd=sqrt(truth["prior.var.mu"])),
      add=TRUE, col="green", lwd=2)
curve(dnorm(x=x, mean=mu, sd=sqrt(1/tau)), add=TRUE, col="black", lwd=2)
curve(dnorm(x=x, mean=truth["post.mean.mu.cond"],
            sd=sqrt(truth["post.var.mu.cond"])),
      add=TRUE, col="red", lwd=2, n=10^3+1)
legend(x="topright", legend=c("true mu|tau", "prior(mu|tau)",
                              "likelihood(y;mu,tau)",
                              "posterior(mu|y,tau)"),
       col=c("black", "green","black","red"), lwd=2, lty=c(2,1,1,1), bty="n")
```


# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
