---
title: "Priors for the regression"
author: "Timothée Flutre (INRA)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Priors for regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
setwd("~/src/rutilstimflutre/vignettes/")

library(devtools)
build_vignettes()

library(rmarkdown)
render("priors-regression.Rmd", "html_document")
-->

# Preamble

License: [CC BY-SA 4.0](http://creativecommons.org/licenses/by-sa/4.0/)

References:

* [Gelman et al (2008)](http://dx.doi.org/10.1214/08-aoas191)

* [Gelman (2006)](http://dx.doi.org/10.1214/06-ba117a)


# Coefficients

Compare the Cauchy prior with the Normal prior for regression coefficients, for instance:

* $c \sim \mathcal{N}(\text{exp}=0, \; \text{stdev}=5)$

* $c \sim \mathcal{C}(\text{loc}=0, \; \text{scale}=5)$

```{r cauchy_normal, fig.width=8, fig.height=6}
plot(x=0, y=0, type="n", xlim=c(-20,20), ylim=c(0,0.08),
     xlab=expression(c), ylab="Density", las=1,
     main="Priors on regression coefficients")
curve(expr=dnorm(x, mean=0, sd=5), from=-25, to=25, n=10^3, add=TRUE,
      col="black")
curve(expr=dcauchy(x, location=0, scale=5), from=-25, to=25, add=TRUE,
      col="red")
curve(expr=dcauchy(x, location=0, scale=25), from=-25, to=25, add=TRUE,
      col="orange")
abline(v=0, lty=2); abline(h=0, lty=2)
legend("topright", legend=c("c ~ Normal(0, 5)","c ~ Cauchy(0, 5)","c ~ Cauchy(0, 25)"),
       col=c("black","red","orange"), lty=1, bty="n")
```


# Variance components

Compare the half-Cauchy prior with the Inverse-Gamma prior for variance components, for instance:

* $\text{variance} \sim \mathcal{IG}(\text{shape}=0.001, \; \text{scale}=1/0.001) \; \; \leftarrow$ classically used by default (e.g. in [OpenBUGS](https://en.wikipedia.org/wiki/OpenBUGS)), but too informative

* $\text{stdev} \sim \mathcal{hC}(\text{loc}=0, \; \text{scale}=5)$

Note that the $x$-axis of the following plots are in units of standard deviation ($\sigma$), not variance ($\sigma^2$):

```{r halfcauchy_invgamma, fig.width=8, fig.height=6}
dinvgamma <- function(x, shape, scale){
  return((scale^shape / gamma(shape)) * (x^(- shape - 1)) * exp(- scale / x))
}
pdfSigma <- function(x, shape, scale){ # see legend of fig 1 of Gelman (2006)
  dgamma(x^(-2), shape=shape, scale=scale) * abs(- 2 * x^(-3))
}
plot(x=0, y=0, type="n", xlim=c(0,10), ylim=c(0,0.27),
     xlab=expression(sigma), ylab="Density", las=1,
     main="Priors on variance components")
curve(pdfSigma(x, shape=0.001, scale=1/0.001), from=0, to=30, n=10^4,
      add=TRUE, col="black")
curve(pdfSigma(x, shape=0.01, scale=1/0.01), from=0, to=30, n=10^4,
      add=TRUE, col="black", lty=2)
curve(pdfSigma(x, shape=0.1, scale=1/0.1), from=0, to=30, n=10^3,
      add=TRUE, col="black", lty=3)
curve(expr=2*dcauchy(x, location=0, scale=5), from=0, to=15, n=10^3,
      add=TRUE, col="red")
curve(expr=2*dcauchy(x, location=0, scale=15), from=0, to=15, n=10^3,
      add=TRUE, col="green")
curve(expr=2*dcauchy(x, location=0, scale=25), from=0, to=15, n=10^3,
      add=TRUE, col="orange")
legend("topright", legend=c("variance ~ inv-Gamma(0.001, 1/0.001)",
                            "variance ~ inv-Gamma(0.01, 1/0.01)",
                            "variance ~ inv-Gamma(0.1, 1/0.1)",
                            "stdev ~ half-Cauchy(0, 5)",
                            "stdev ~ half-Cauchy(0, 15)",
                            "stdev ~ half-Cauchy(0, 25)"),
       col=c(rep("black",3),"red","green","orange"), lty=c(1,2,3,1,1,1), bty="n")
```

And here is a focus on the half-Cauchy (note the different ranges of $x$ and $y$ axes compare to the previous plot):

```{r focus_halfcauchy, fig.width=8, fig.height=6}
plot(x=0, y=0, type="n", xlim=c(0,200), ylim=c(0,0.05),
     xlab=expression(sigma), ylab="Density", las=1,
     main="Focus on the half-Cauchy")
abline(h=0, lty=2); abline(v=0, lty=2)
curve(expr=2*dcauchy(x, location=0, scale=5), from=0, to=200, n=10^3,
      add=TRUE, col="red")
curve(expr=2*dcauchy(x, location=0, scale=15), from=0, to=200, n=10^3,
      add=TRUE, col="green")
curve(expr=2*dcauchy(x, location=0, scale=25), from=0, to=200, n=10^3,
      add=TRUE, col="orange")
legend("topright", legend=c("stdev ~ half-Cauchy(0, 5)",
                            "stdev ~ half-Cauchy(0, 15)",
                            "stdev ~ half-Cauchy(0, 25)"),
       col=c("red","green","orange"), lty=1, bty="n")
```


# Appendix

```{r info}
print(sessionInfo(), locale=FALSE)
```
